### 2023-08-24 doc/15yrsChange.Rmd
initial commit

---
title: ''
output: 
    #pdf_document: 
    bookdown::word_document2: 
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: references.bib
csl: new-phytologist.csl

---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\raggedright

**Title:** Nuisance species compromise the carbon sequestration potential in an Eastern US temperate deciduous forest

**Authors (tentative / incomplete list, not in order):**

Krystal Bagnaschi

Rachel Hoffman

Iris Kennedy

Erin MacMonigle

Caroline Troy

Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)

David Mitre

Jess Shue

Luca Morreale

William McShea

Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)

**Author Affiliations:**

1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, Virginia 22630, USA

2. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama

*corresponding author: teixeirak@si.edu; +1 540 635 6546

**Running head:** <!-- ≤45 characters-->

<!--JOURNAL

Target journal: 
Instructions for authors: 
article type:  -->

```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
# DISPLAY ITEMS
## I do not see any guidance as to the number of display items allowed in MEE.

#Tables
## example = 1

#Figures
## example = 1

```

\newpage

# Abstract  {-#Abstract}

- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm. 
- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
- we show increased biomass mortality, reduced ∆AGB (outside deer exclosure), reduced AGB (outside deer exclosure)?
- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
- thus, C sequestration potential of temperate deciduous forests is overestimated

**Keywords**:

\newpage

# Introduction

The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019]. 
Although currently a C sink, the future of the biome remains uncertain.
The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023; @refs], but with declining CO~2~ sequestration capacity. <!--I think this is true, but check/ find more refs.-->
However, ... [@wu_uncertainty_2023]
<!--Global C models generally agree that the biome's CO~2~ sequestration capacity is likely to decline [@ahlstrom_robustness_2012].-->
Yet, global C models predict an wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014], including the effects of nuisance species.

Nuisance species...

- often non-endemic
- white-tailed deer is important endemic nuisance species, over-abundant because of human influence

Tree mortality ...

Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].

The capacity to regenerate following disturbances, including ongoing gap formation through mortality of canopy trees, critically influences long-term forest dynamics.
Regeneration depends first upon seed production and then upon seedling recruitment, survival, and growth into trees.
When any one of these steps fails, the stage is set for disturbance to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to it's pre-disturbance state in the foreseeable future [@refs].
Global change pressures can set the stage for such critical transitions by gradually shifting baseline conditions, making post-disturbance recovery unlikely despite the persistence of mature trees [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @refs]. 
In the mid-Atlantic region of eastern North America, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of current canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019;@miller_overabundant_2023]. 
Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023].
The juveniles that are present tend to represent a different, more mesophytic set of species (*Acer spp.*, *Fagus grandifolia*) than currently dominate much of the region (*Quercus spp.*, *Carya spp.*, @miller_compounding_2019; @nowacki_climate_2015]--a dynamic driven by fire suppression and mesophication [@refs].

Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
(1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (or at least ∆AGB)
(3) C sequestration (∆AGB) has declined, and will decline further if current trends continue

# Materials and Methods

# Results

# Discussion

# Conclusions (optional)

# Acknowledgements {-#Acknowledgements}

- Funding: ForestGEO

# Conflict of Interest statement {-#Conflict}
The authors declare no conflict of interest.

# Authors' contributions {-#Authors}
*[Name of author 1] and [Name of author 2] conceived the ideas and designed methodology; [Name of author 1] and [Name of author 3] collected the data; [Name of author 2] and [Name of author 4] analysed the data; [Name of author 1] and [Name of author 4] led the writing of the manuscript. All authors contributed critically to the drafts and gave final approval for publication.*

# Data availability {-#Data}

\newpage
# References {-#References}

### 2023-09-05 doc/15yrsChange.Rmd
misc
Update 15yrsChange.Rmd
various updates

@@
 
+Cameron Dow
+
+Jen Jordan
+
 Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
@@
 **Running head:** <!-- ≤45 characters-->
+
+for submission to Anthromes joint special collection in New Phytologist/ Plants, People, Planet: https://www.newphytologist.org/news/view/332
 
@@
 - Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
-- we show increased biomass mortality, reduced ∆AGB (outside deer exclosure), reduced AGB (outside deer exclosure)?
+- we show increased biomass mortality, reduced ∆AGB (only outside deer exclosure?), reduced AGB (only outside deer exclosure?)
 - if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
@@
 
+# Papers to read
+- earth system models (pay special attention to predictions and uncertainty for eastern US)
+  - @wu_uncertainty_2023. - do they say anything about nuisance species?
+  - @ahlstrom_robustness_2012
+- @miller_overabundant_2023
+  - what were there metrics? would it be worth applying their metrics to our forest in order to classify by their system?
+
 # Introduction
@@
 Although currently a C sink, the future of the biome remains uncertain.
-The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023; @refs], but with declining CO~2~ sequestration capacity.
-<!--I think this is true, but check/ find more refs.-->
-However, ... [@wu_uncertainty_2023]
-<!--Global C models generally agree that the biome's CO~2~ sequestration capacity is likely to decline [@ahlstrom_robustness_2012].-->
-Yet, global C models predict an wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014], including the effects of nuisance species.
+The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012].
+Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
+One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
+Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
+This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world [@refs].
 
-Nuisance species...
-
-- often non-endemic
-- white-tailed deer is important endemic nuisance species, over-abundant because of human influence
-
-Tree mortality ...
+## (paragraph on tree mortality from non-endemic pests and pathogens)
 
 Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
+
+## (paragraph on recruitment failure because of deer and non-endemic plants)
+
+- white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence
+
+@mcgarvey_effects_2013
 
@@
 (1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
-(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (or at least ∆AGB)
-(3) C sequestration (∆AGB) has declined, and will decline further if current trends continue
+(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (-∆AGB)
+(3) C stocks have declined (-∆AGB), and will decline further if current trends continue
 
 # Materials and Methods
+
+## Study site
+
+## Data collection
+
+## Analyses
 

### 2023-09-06 doc/15yrsChange.Rmd
incorporate references

@@
 
-# Papers to read
-- earth system models (pay special attention to predictions and uncertainty for eastern US)
-  - @wu_uncertainty_2023. - do they say anything about nuisance species?
-  - @ahlstrom_robustness_2012
-- @miller_overabundant_2023
-  - what were there metrics? would it be worth applying their metrics to our forest in order to classify by their system?
-
 # Introduction
@@
 The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012].
-Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
+Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
 One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
@@
 
+@fei_biomass_2019
+
 ## (paragraph on recruitment failure because of deer and non-endemic plants)
@@
 @mcgarvey_effects_2013
+@holm_interactive_2013
 

### 2023-09-21 doc/15yrsChange.Rmd
outline methods, add photos figure

@@
 #Figures
-## example = 1
-
+photos = 1
+photo_represenative_upland = "1a"
+photo_deer_exclosure = "1b"
+photo_graveyard = "1c"
 ```
@@
 The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012].
-Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014].
+Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
 One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
-This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world [@refs].
+This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
 
@@
 Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
+These can reduce C and need to be considered in future climate change projections.
 
@@
 
+*(describe location, forest type, etc)*
+
+In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot:
+(1) representative upland forest (Fig.
+`r photo_represenative_upland`) - # ha of the plot, excluding areas around streams (*delineated using topographic wetness index, described in @mcgregor_tree_2021*) and the two special focal areas
+(2) deer exclosure (Fig. `r photo_deer_exclosure`)- 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained mainly deer-free for the past ## years;
+(3) "tree graveyard" (Fig.
+`r photo_graveyard`) - a # ha section of upland forest in the NE corner of the plot where extensive mortality of *Fraxinus* and other canopy trees within the past decade has resulted in an open canopy and heavy ground cover of invasive plants.
+The latter was delineated based on *(define & describe criteria)*
+
+![**Photos within the SCBI ForestGEO plot: (a) represenative upland forest, (b) deer exclosure, (c) "tree graveyard" in NE corner ** *(Perhaps combine these with a map of the plot?)*](display/photos.png)
+
 ## Data collection
+
+*(describe main census & years occurred)*
+
+*(describe mortality census)*
+
+*(describe invasive plants survey)*
+
+*(describe seedlings survey, if used)*
 

### 2023-09-21 doc/15yrsChange.Rmd
@llmorreale will be lead author
update list of figures & hypotheses based on group discussion

@@
 
+Luca Morreale
+
 Krystal Bagnaschi
@@
 Jess Shue
-
-Luca Morreale
 
@@
 #Figures
-photos = 1
+map_and_photos = 1
 photo_represenative_upland = "1a"
@@
 photo_graveyard = "1c"
+
+biomass_trends = 2 #line plot showing biomass through time for entire plot and each of the 3 focal sections
+
+loss_gain_net = 3 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) upland, (c) deer exclosure, (d) graveyard
+
+mortality_by_genera = 4 # 4-panel figure showing biomass mortality by genera for for (a) whole plot, (b) upland, (c) deer exclosure, (d) graveyard
+
+regen_failure = 5 #some figure illustrating absence of small canopy trees/ shifting  size distributions
+
 ```
@@
 Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
-(1) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
-(2) growth and recruitment (outside deer exclosure) have not kept pace with tree mortality, resulting in declining AGB (-∆AGB)
-(3) C stocks have declined (-∆AGB), and will decline further if current trends continue
+(1) C stocks have declined (-∆AGB)
+(2) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
+(3) recruitment (outside deer exclosure) and growth have not kept pace with tree mortality, implying that future mortality will result in substantial net biomass loss (regeneration debt)
 
@@
 
-![**Photos within the SCBI ForestGEO plot: (a) represenative upland forest, (b) deer exclosure, (c) "tree graveyard" in NE corner ** *(Perhaps combine these with a map of the plot?)*](display/photos.png)
+![**Figure `r map_and_photos`.
+Photos within the SCBI ForestGEO plot: (a) represenative upland forest, (b) deer exclosure, (c) "tree graveyard" in NE corner ** *(Perhaps combine these with a map of the plot?)*](display/photos.png)
 

### 2023-09-22 doc/15yrsChange.Rmd
change how we describe focal areas
@llmorreale , I think it makes sense to present the areas as (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens (in that order). This stems from your comment yesterday, and also thinking we should present this as progressive degradation.

@@
 map_and_photos = 1
-photo_represenative_upland = "1a"
-photo_deer_exclosure = "1b"
+photo_deer_exclosure = "1a"
+photo_represenative_upland = "1b"
 photo_graveyard = "1c"
@@
 
-loss_gain_net = 3 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) upland, (c) deer exclosure, (d) graveyard
+loss_gain_net = 3 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard
 
-mortality_by_genera = 4 # 4-panel figure showing biomass mortality by genera for for (a) whole plot, (b) upland, (c) deer exclosure, (d) graveyard
+mortality_by_genera = 4 # 4-panel figure showing biomass mortality by genera for for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard
 
@@
 
-In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot:
-(1) representative upland forest (Fig.
-`r photo_represenative_upland`) - # ha of the plot, excluding areas around streams (*delineated using topographic wetness index, described in @mcgregor_tree_2021*) and the two special focal areas
-(2) deer exclosure (Fig. `r photo_deer_exclosure`)- 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained mainly deer-free for the past ## years;
-(3) "tree graveyard" (Fig.
-`r photo_graveyard`) - a # ha section of upland forest in the NE corner of the plot where extensive mortality of *Fraxinus* and other canopy trees within the past decade has resulted in an open canopy and heavy ground cover of invasive plants.
-The latter was delineated based on *(define & describe criteria)*
+In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
+(1) low deer, low non-endemic insects & pathogens (Fig.
+`r photo_deer_exclosure`)- 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
+(2) high deer, low non-endemic insects & pathogens (Fig.
+`r photo_represenative_upland`) - # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
+(3) high deer, high non-endemic insects & pathogens (Fig.
+`r photo_graveyard`) - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens.
+These areas were delineated as follows.
+Upland forest was defined according to topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
+Vulnerability to non-endemic insects & pathogens was defined based on the abundance of canopy species affected by non-endemic pests and pathogens at the time of plot establishment (2008).
 
 ![**Figure `r map_and_photos`.
-Photos within the SCBI ForestGEO plot: (a) represenative upland forest, (b) deer exclosure, (c) "tree graveyard" in NE corner ** *(Perhaps combine these with a map of the plot?)*](display/photos.png)
+Photos within the SCBI ForestGEO plot: (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens ** *(add a map of the plot)* All photos taken September 2023 by K.
+Anderson-Teixeira.](display/photos.png)
 

### 2023-09-25 doc/15yrsChange.Rmd
reference updates from lab meeting

@@
 
+Non-endemic pests and pathogens facilitate invasin of non-endemic plants [@guo_impacts_2023 and refs therein].
+
 ## (paragraph on recruitment failure because of deer and non-endemic plants)
@@
 @holm_interactive_2013
+@knauer_understory_2023
 
@@
 In the mid-Atlantic region of eastern North America, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of current canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019;@miller_overabundant_2023].
-Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023].
+Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023;@gorchov_differential_2021].
+Deer are the biggest problem [@gorchov_differential_2021].
 The juveniles that are present tend to represent a different, more mesophytic set of species (*Acer spp.*, *Fagus grandifolia*) than currently dominate much of the region (*Quercus spp.*, *Carya spp.*, @miller_compounding_2019; @nowacki_climate_2015]--a dynamic driven by fire suppression and mesophication [@refs].
@@
 
+apply the metric of @miller_overabundant_2023 to determine imminent failure/ probable failure/ insecure/ secure?
+
 # Results
@@
 # Discussion
+
+@holtmann_carbon_2021
 

### 2023-10-24 doc/15yrsChange.Rmd
added code to allow HTML tags when knitting to word doc

@@
     bookdown::word_document2: 
+      md_extensions: +raw_html-markdown_in_html_blocks
+      pandoc_args: ['--lua-filter', 'read_html.lua']
       reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
@@
 
+\newpage
+
+# References {-}
+
+<div id="refs"></div>
+
 # Data availability {-#Data}
-
-\newpage
-# References {-#References}

### 2023-10-26 code/mortalityFluxes.r
Code for mortality fluxes
R code for mortality fluxes for each census

@@ -1,0 +1,102 @@
+##mortality figure (figure 3)
+##biomass fluxes per year
+## september 28
+library(tidyverse)
+library(ggplot2)
+library(allodb)
+
+#"C:\Users\irisa\Documents\Smithsonian\15yearsChange\creatingFigure3.r"
+
+#top of script load in current census and subsample which quadrats have been done
+
+
+#download save and then load it in with the path, 1 is oldest, 2 is 2013, 3 is 2018
+#change blob to raw, alternatively you can download and save the raw data and then load in the filepath
+
+census<- read.csv("C:/Users/irisa/Documents/GitHub/2023census/processed_data/scbi.stem4.csv")#loading in current census data
+
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
+Census2013 <- load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
+
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+
+#renaming data
+Census2008 <- scbi.stem1
+Census2013 <- scbi.stem2
+Census2018 <- scbi.stem3
+
+# subsetting data to the quadrats that are completed in the 2023 census
+completeQuadrats <- unique(census$quadrat)##stores complete quadrats as a vector, now we can use subset
+hectaresMeasured <- length(completeQuadrats)*20*20/10000 #fixed number size of our plot, 25.6, dividing 20/20/1000 puts m^2 in hectares
+
+Census2008 <- subset(Census2008, Census2008$quadrat%in%completeQuadrats)
+Census2013 <- subset(Census2013, Census2013$quadrat%in%completeQuadrats)
+Census2018 <- subset(Census2018, Census2018$quadrat%in%completeQuadrats)
+
+
+all(Census2008$tag==Census2013$tag) ## function all checks if everything is true in this statement
+all(Census2008$StemTag==Census2013$StemTag) #if true data sets are ordered the same way
+
+#subsetting what was Alive 'A'in 2008 and Dead 'D' in 2013, %in%
+
+idx <- Census2008$status%in%"A" & Census2013$status%in%"D"
+
+view(Census2013[idx, ])
+Census2008[idx, ]
[…]

### 2023-10-26 code/figure 5 regeneration debt.R
in progress code of regeneration debt map

@@ -1,0 +1,90 @@
+##map plots of scbi forestGEo site
+
+#clear environment
+rm(list = ls())
+
+#load libraries
+library(here)
+library(data.table)
+library(dplyr)
+library(ggplot2)
+library(sf)
+
+# quadrat map
+quadrats <- read_sf("doc/maps/20m_grid/20m_grid.shp")
+
+ggplot() + geom_sf(data = quadrats)
+
+##load cencus data
+cencus2013 <- read.csv("C:/Users/elmgi/Downloads/scbi.stem2.csv")
+cencus2023 <- read.csv("C:/Users/elmgi/Downloads/scbi.stem4.csv")
+
+#read in species table
+spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
+
+#merge cencus data with sp table LOSING ROWS OF UNKNOWN SP, FIX
+cencus2023 <- merge(cencus2023, spTable, by.y = "spcode", by.x = "species", all.x = T)
+cencus2013 <- merge(cencus2013, spTable,  by.y = "spcode", by.x = "sp", all.x = T)
+
+####view sp not included in species table
+###setdiff(cencus2013$sp, spTable$spcode)
+###
+####create missing sp table and input canopy information
+###missing_sp <- data.frame(spcode = setdiff(cencus2013$sp, spTable$spcode), canopy_position = c("1","1","0","1", "1","0","1","1", "0"))
+###cencus2023[is.na(cencus2023$canopy_position), ]
+
+#create canopy trees column
+cencus2023$is_canopy <- ifelse(cencus2023$canopy_position == 'canopy', 1, #this makes canopy trees = 1 and non-canopy trees = 0
+                                     ifelse(cencus2023$canopy_position == 'canopy, emergent', 1, 0))
+
+cencus2013$is_canopy <- ifelse(cencus2013$canopy_position == 'canopy', 1, 
+                                     ifelse(cencus2013$canopy_position == 'canopy, emergent', 1, 0))
+
+#convert new column to numeric for map
+cencus2023$is_canopy <- as.numeric(cencus2023$is_canopy)
+cencus2013$is_canopy <- as.numeric(cencus2013$is_canopy)
+
+####check new attribute
+###check <- select(cencus2023, species, canopy_position, is_canopy, quadrat)
[…]

### 2023-10-26 code/Figure 3 (Woody Growth) Draft 3.R
Draft of code - Woody productivity

@@ -1,0 +1,93 @@
+#Figure 3 (Woody Growth)
+
+#Look at the "Mortality, Growth and Recruitment - Conceptual Differences" word document if you're confused
+
+#Read in data from past censuses and the species table
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+
+#Rename the data
+Census_2008 <- scbi.stem1
+Census_2013 <- scbi.stem2
+Census_2018 <- scbi.stem3
+Species_table <- scbi.spptable
+
+#Combine the census data and species list
+Census_2008_Complete <- merge(Census_2008, Species_table, by ="sp")
+Census_2013_Complete <- merge(Census_2013, Species_table, by ="sp")
+Census_2018_Complete <- merge(Census_2018, Species_table, by ="sp")
+
+#Read in data from 2023
+Census_2023 <- read.csv("/Users/rachelhoffman/Documents/GitHub/2023census/processed_data/scbi.stem4.csv")
+#Combine the 2023 census data and species list
+Census_2023_Complete <- merge(Census_2023, Species_table, by ="sp")
+#Define finished quadrats
+Finished_quadrats <- unique(Census_2023$quadrat)
+
+#Subset the 2008 quadrats by ones that are finished in 2023
+subset(Census_2008_Complete, Census_2008_Complete$quadrat %in% Finished_quadrats)
+Census_2008_Complete_Finishedquadrats <- subset(Census_2008_Complete, Census_2008_Complete$quadrat %in% Finished_quadrats)
+
+#Subset the 2013 quadrats by ones that are finished in 2023
+subset(Census_2013_Complete, Census_2013_Complete$quadrat %in% Finished_quadrats)
+Census_2013_Complete_Finishedquadrats <- subset(Census_2013_Complete, Census_2013_Complete$quadrat %in% Finished_quadrats)
+
[…]

### 2023-10-31 doc/15yrsChange.Rmd
update for special issue abstract submission

@@
 
-Luca Morreale
+Luca L. Morreale^1,2^*
 
@@
 
-Kristina J. Anderson-Teixeira^1,2^* ( Orcid ID : 0000-0001-8461-9713)
+Kristina J. Anderson-Teixeira^1,2^ ( Orcid ID : 0000-0001-8461-9713)
 
@@
 
-*corresponding author: teixeirak@si.edu; +1 540 635 6546
+*corresponding author: MorrealeLL@si.edu; +1 607 227 1350
 
@@
 
-- temperate deciduous forests play a critical role in the global carbon cycle, accounting for a substantial portion of the global forest C sink
-- The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century
-- However, these models do not incorporate nuisance species -- i.e., species whose local abundance has been dramatically increased as a result of human activities and that are causing ecological harm.
-- Nuisance species are known to increase tree mortality (non-endemic pests and pathogens) and decrease recruitment (deer, non-endemic plants) throughout the (Eastern US) temperate deciduous biome, yet we don't know net effect on current and future C cycling
-- Here, we use 15 years of detailed census data from a large forest dynamics plot with a deer exclosure to understand how nuisance species are affecting C cycling
-- we show increased biomass mortality, reduced ∆AGB (only outside deer exclosure?), reduced AGB (only outside deer exclosure?)
-- if current trends continue, forest will continue to lose C sequestration capacity/ likely to be C source in future
-- thus, C sequestration potential of temperate deciduous forests is overestimated
+Temperate deciduous forests play a critical role in the global carbon (C) cycle, accounting for a substantial portion of the global forest C sink.
+The dominant view imbued in Earth System Models is that, at least within the eastern United States, the biome is likely to remain a C sink throughout the 21st century.
+However, these models do not incorporate nuisance species -- *i.e.*, species whose local abundance has increased as a result of human activities and that are causing ecological harm.
+Nuisance species are known to increase tree mortality (*e.g.*, non-endemic pests and pathogens) and decrease recruitment (*e.g.*, deer and non-endemic plants) throughout the temperate deciduous biome, yet the net effect on current and future C cycling is unknown.
+Here, we use 15 years of detailed census data from a large (25.6 ha) forest dynamics plot in a Mid-Atlantic temperate forest, including a 4-ha deer exclosure, to understand how nuisance species are affecting C cycling.
+We show increased biomass mortality, a net reduction in aboveground C storage, and decreased abundance of canopy tree species in the understory.
+Under current trends this forest will continue to lose C sequestration capacity, indicating that the C sequestration of temperate deciduous forests is likely overestimated.
 

### 2024-01-22 code/exploratory/data_exploration_v1.Rmd
region identification, figure 2 adjustments

---
title: "SCBI 2023 Census Exploration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = "hide",message = FALSE, warning = FALSE)
```

```{r init}
library(tidyverse)
library(data.table)
library(sf)
library(RColorBrewer)

curStem <- fread("C:/Work/Smithsonian/Repos/2023census/processed_data/scbi.stem4.csv")

prevStem <- fread(paste0("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/tree_main_census/data/census-csv-files/scbi.stem3.csv"))

load("C:/Work/Smithsonian/Repos/Morreale_BiomassChange/data/census/scbi.stem2.rdata")

load("C:/Work/Smithsonian/Repos/Morreale_BiomassChange/data/census/scbi.stem1.rdata")

quadrats <- st_read( "C:/Work/Smithsonian/Repos/2023census/doc/maps/20m_grid/20m_grid.shp")

```

### Recruitment 2018-2023 ###

#### Trees ####

```{r recruits I}

recruits <- curStem %>% 
  filter(curStem$table == "recruits") %>% 
  select(tag,StemTag,quadrat,sp)

r_stems <- recruits %>% 
  group_by(sp) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))
r_trees <- recruits %>% 
  group_by(tag) %>% 
  slice(1) %>% 
  ungroup() %>% 
  group_by(sp) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))

ggplot(r_trees,aes(x = reorder(sp,-n),y = n, fill = -n)) +
  geom_bar(stat =  "identity",col = "black") +
  scale_fill_gradient(low = "grey20",high = "grey80") +
  theme_classic() +
  ylab("# of Recruits (Trees)") +
  xlab(" ") +
  theme(legend.position = "none")
```

#### Stems ####

```{r recruits II}
ggplot(r_stems,aes(x = reorder(sp,-n),y = n, fill = -n)) +
  geom_bar(stat =  "identity",col = "black") +
  scale_fill_gradient(low = "grey20",high = "grey80") +
  theme_classic() +
  ylab("# of Recruits (Stems)") +
  xlab(" ") +
  theme(legend.position = "none")
```

### Mortality 2018-2023 ###

#### Stems ####

```{r mortality}
ids_prev_living <- prevStem %>% 
  filter(status == "A") %>% 
  mutate(taglink = paste0(tag,StemTag))
m_stems <- curStem %>% 
  mutate(taglink = paste0(tag,StemTag)) %>% 
  filter(taglink %in% ids_prev_living$taglink & status_current != "LI") %>% 
  group_by(sp) %>% 
  summarize(n = n())

ggplot(m_stems,aes(x = reorder(sp,-n),y = n, fill = -n)) +
  geom_bar(stat =  "identity",col = "black") +
  scale_fill_gradient2(low = "grey40",high = "grey80") +
  theme_classic() +
  ylab("Mortality (# of Stems)") +
  xlab(" ") +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45))

```

### Compositional Shifts ###

#### Relative Abundance (by stem count) of top 15 species in 2018 census ####
```{r }
complete_quads <- unique(curStem$quadrat) %>% na.omit()

c2018_ra <- prevStem %>% #calculate relative abundance for 2018 census in quadrats that have been completed so far in 2023
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  group_by(sp) %>% 
  summarize(StemCount = n()) %>% 
  mutate(RelAbund = StemCount/sum(StemCount), Census = "2018") %>% 
  arrange(desc(RelAbund))

c2013_ra <- scbi.stem2 %>%  
  mutate(quadrat = as.numeric(quadrat)) %>% 
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  group_by(sp) %>% 
  summarize(StemCount = n()) %>% 
  mutate(RelAbund = StemCount/sum(StemCount), Census = "2013") %>% 
  arrange(desc(RelAbund))

c2008_ra <- scbi.stem1 %>%  
  mutate(quadrat = as.numeric(quadrat)) %>% 
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  group_by(sp) %>% 
  summarize(StemCount = n()) %>% 
  mutate(RelAbund = StemCount/sum(StemCount), Census = "2008") %>% 
  arrange(desc(RelAbund))

allCens_ra <- curStem %>% 
  filter(status_current == "LI") %>% 
  group_by(sp) %>% 
  summarize(StemCount = n()) %>% 
  mutate(RelAbund = StemCount/sum(StemCount), Census = "2023") %>% 
  bind_rows(c2018_ra, c2013_ra,c2008_ra) %>% 
  #bind_rows(c2018_ra) %>% 
  mutate(Census = as.factor(Census))

topRanked_sp <- c2018_ra %>% 
  slice_head(n = 15)

plotdf <- allCens_ra %>% 
  filter(sp %in% topRanked_sp$sp)

ggplot(plotdf,aes(x = reorder(sp,-RelAbund),y = RelAbund, group = Census, fill = Census)) + 
  geom_bar(stat = "identity",position = "dodge", col = "black") +
  scale_fill_grey() +
  xlab("") +
  ylab("Relative Abundance") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45))
```

### Species Size Class Distributions of top 15 species in 2018 census ###

```{r fig.height= 30, fig.width = 6}
brks <- c(seq(0,120, by = 20),170)

c2018_sp_size_ab <- prevStem %>%  
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  mutate(dbh_classes = cut(as.numeric(dbh)/10,breaks = brks,right = F)) %>% 
  group_by(sp,dbh_classes) %>% 
  summarize(StemCount = n()) %>% 
  ungroup() %>% 
  complete(sp,nesting(dbh_classes),fill = list(StemCount = 0)) %>% 
  mutate(Census = "2018") %>% 
  arrange(desc(StemCount))

c2013_sp_size_ab <- scbi.stem2 %>%  
  mutate(quadrat = as.numeric(quadrat)) %>% 
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  mutate(dbh_classes = cut(as.numeric(dbh)/10,breaks = brks,right = F)) %>% 
  group_by(sp,dbh_classes) %>% 
  summarize(StemCount = n()) %>% 
  ungroup() %>% 
  complete(sp,nesting(dbh_classes),fill = list(StemCount = 0)) %>% 
  mutate(Census = "2013") %>% 
  arrange(desc(StemCount))

c2008_sp_size_ab <- scbi.stem1 %>%  
  mutate(quadrat = as.numeric(quadrat)) %>% 
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  mutate(dbh_classes = cut(as.numeric(dbh)/10,breaks = brks,right = F)) %>% 
  group_by(sp,dbh_classes) %>% 
  summarize(StemCount = n()) %>% 
  ungroup() %>% 
  complete(sp,nesting(dbh_classes),fill = list(StemCount = 0)) %>% 
  mutate(Census = "2008") %>% 
  arrange(desc(StemCount))

allCens_sp_size_ab <- curStem %>% 
  filter(status_current == "LI") %>% 
  mutate(dbh_classes = cut(as.numeric(dbh_current)/10,breaks = brks,right = F)) %>% 
  group_by(sp,dbh_classes) %>% 
  summarize(StemCount = n()) %>% 
  ungroup() %>% 
  complete(sp,nesting(dbh_classes),fill = list(StemCount = 0)) %>% 
  mutate(Census = "2023") %>% 
  bind_rows(c2018_sp_size_ab, c2013_sp_size_ab,c2008_sp_size_ab) %>% 
  #bind_rows(c2018_sp_size_ab) %>% 
  mutate(Census = as.factor(Census))

plotdf_size <- allCens_sp_size_ab %>% 
  filter(sp %in% topRanked_sp$sp)

ggplot(plotdf_size,aes(x = dbh_classes,y = StemCount, group = Census, fill = Census)) + 
  geom_bar(stat = "identity",position = "dodge", col = "black") +
 # geom_line(aes(col = Census)) +
  facet_wrap(~sp,scales = "free",ncol = 1) +
  scale_fill_grey() +
  xlab("") +
  ylab("Number of stems") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45,vjust = .8))

```

### Change in Aboveground Biomass ###

```{r}
library(allodb)

sp.table <- fread(paste0("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")) %>% 
  select(spcode,family,genus,species)
latlong <- c(-78.1454, 38.8935)
area_ha_measured <- (length(unique(curStem$quadrat)) * 400) / 10000

curLive <- curStem %>% 
  filter(status_current == "LI") %>% 
  select(quadrat,dbh = dbh_current,spcode = species) %>% 
  mutate(Census = "2023")
AllLive <- scbi.stem1 %>%  
  mutate(Census = "2008") %>% 
  bind_rows(scbi.stem2 %>% mutate(Census = "2013")) %>% 
  mutate(quadrat = as.numeric(quadrat)) %>% 
  filter(status == "A" & quadrat %in% complete_quads) %>% 
  select(quadrat,dbh, spcode = sp,Census) %>% 
  bind_rows(prevStem  %>% 
              filter(status == "A" & quadrat %in% complete_quads) %>% select(quadrat,dbh,spcode = sp) %>% mutate(Census = "2018")) %>% 
  mutate(dbh = as.numeric(dbh)) %>% 
  bind_rows(curLive)

AGB_stems <- AllLive %>% 
  left_join(sp.table) %>% 
  mutate(AGB = allodb::get_biomass(dbh = dbh/10, genus = genus, species = genus, coords = latlong))

census_AGB <- AGB_stems %>% 
  mutate(AGB_Mg = AGB / 1000) %>% 
  group_by(Census) %>% 
  summarize(AGB_ha = sum(AGB_Mg)/area_ha_measured)

species_AGB <- AGB_stems %>% 
#  filter(spcode %in% topRanked_sp$sp) %>% 
  mutate(plt_sp = if_else(spcode %in% c("litu","fram","qual","quru","quve"),
                          spcode,"Other")) %>% 
  mutate(AGB_Mg = AGB / 1000) %>% 
  group_by(Census,plt_sp) %>% 
  summarize(AGB_ha = sum(AGB_Mg)/area_ha_measured)

ggplot(census_AGB ,aes(x = as.numeric(Census), y = AGB_ha)) +
  geom_point(cex = 2, col = "blue") +
  geom_line(alpha = .2, lwd = 1.2, col = "blue") +
  xlab("Census") +
  ggtitle("SCBI Aboveground Biomass") +
  ylab(expression(AGB~(Mg~ha^-1))) +
  theme_classic() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 20),
        title = element_text(size = 22))

ggplot(species_AGB ,aes(x = as.numeric(Census), y = AGB_ha,group = plt_sp, col = plt_sp)) +
  geom_point(cex = 2) +
  geom_line(alpha = .2, lwd = 1.2) +
  xlab("Census") +
  ggtitle("Selected Species AGB") +
  ylab(expression(AGB~(Mg~ha^-1))) +
  scale_y_continuous(breaks = seq(0,140,by = 20))  +
  scale_color_brewer(palette = "Dark2", name = "Species") +
  theme_bw() +
  theme(axis.text = element_text(size = 14),
        axis.title = element_text(size = 20),
        title = element_text(size = 22))

```

### 2024-01-22 code/figure2_agb.R

@@ -1,6 +1,7 @@
 #figure 2 ABG (Above Ground Biomass)
 #output Line graph of aboveground biomass stocks
 #december 13th, 2023
+#Updated by LLM on 1/22/2024
 
 #clear environment
 rm(list = ls())
@@ -10,6 +11,8 @@ library(allodb)
 library(ggplot2)
 
 #Read in data from past censuses and the species table
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
@@ -35,16 +38,15 @@ spCensus2023$quadrat <- sprintf("%04d", spCensus2023$quadrat)
 latlong <- c(-78.1454, 38.8935)
 
 #######################################################################################################################################
-                                                #for unfinished census data
-
+                                          
 #Define finished quadrats
-Finished_quadrats <- unique(spCensus2023$quadrat)
+#Finished_quadrats <- unique(spCensus2023$quadrat)   #for unfinished census data
+Finished_quadrats <- unique(spCensus2018$quadrat)  #Use this line if you're interested in the biomass for the whole plot
 
 #Finding the physical amount of space we measured
 Hectares_measured <- length(Finished_quadrats)*20*20/10000
 
-###Use this line if you're interested in the biomass for the whole plot = Finished_quadrats <- unique(Census_2018$quadrat)
-##Census_2023_Finished <- subset(spCensus2023, spCensus2023$quadrat %in% Finished_quadrats)
+#Census_2023_Finished <- subset(spCensus2023, spCensus2023$quadrat %in% Finished_quadrats)
 
[…]

### 2024-01-22 code/identify-plot-regions.r

@@ -2,7 +2,7 @@ library(sf)
 library(tidyverse)
 library(terra) 
 library(tidyterra)
-#library("dendextend")
+
 
 setwd("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/")
 load("tree_main_census/data/scbi.stem1.rdata")
@@ -12,7 +12,8 @@ quadrats <- st_read("spatial_data/shapefiles/20m_grid.shp")
 dem <- rast("spatial_data/elevation/rasters/plot_elevation.tif")
 deer_exclosure <- st_read("spatial_data/shapefiles/deer_exclosure_2011.shp")
 twi <- rast("spatial_data/elevation/rasters/plot_TWI.tif")
-#twi_quad <- rast("spatial_data/elevation/rasters/plot_TWI_quadrat.tif")
+
+##### Remove lowland quadrats #####
 twi_nostreams <- twi
 twi_nostreams[twi_nostreams > 10] <- NA
 #plot(twi_nostreams)
@@ -20,20 +21,21 @@ twi_nostreams[twi_nostreams > 10] <- NA
 quadrats <- quadrats  %>% 
     mutate(TWI = unlist(zonal(twi_nostreams,vect(quadrats), fun = mean, na.rm = F)))
 
-###plot TWI with cut-off of 10###
-twi_fig <- ggplot() +
+##plot TWI with cut-off of 10##
+ggplot() +
     geom_spatraster(data  = twi_nostreams) +
+   # geom_sf(data = quadrats,aes( fill = TWI)) + 
 #    geom_spatraster(data  = twi_quad) +
     scale_fill_whitebox_c(palette = "muted") +
     geom_sf(data = deer_exclosure, colour = "black", fill = NA,lwd = 1.5) +
    # geom_sf(data = quadrats, colour = "black", fill = NA,lwd = 1.5) +
     theme_classic()
 
-ggsave(twi_fig,filename = "C:/Work/Smithsonian/Repos/15yrsChange/doc/nondisplay-figs/TWI_nostreams.jpeg", units = "in",width = 5, height = 6)
 
 
+##### get Basal Area of canopy species #####
 
-### get Basal Area of canopy species ###
+affected_species <- c("fram","frni","frpe","frsp","ulam","ulru","ulsp","juci","cade") # species affected by pests & pathogens
 stem.sp <- scbi.stem1  %>% 
     left_join(scbi.spptable)
 
@@ -42,120 +44,49 @@ canopy_sp <- stem.sp  %>%
[…]

### 2024-01-24 code/0_Update_previous_censuses.r
update previous censuses for 2023 census

@@ -1,0 +1,74 @@
+library(tidyverse)
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+census2023 <- read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))
+
+
+sp_correct <- census2023[grep(pattern = "is not",tolower(census2023$notes_current)),]
+
+
+ptrn <- paste(c("species","Species",paste0("(is ",scbi.spptable$sp,")")), collapse = "|")
+outside_col1_tags <- c(33561,53854, 53855)
+
+census2023_corrections <- census2023  %>% 
+    create_stem_UID()  %>% 
+    group_by(tag)  %>% 
+    arrange(tag,StemTag)  %>% 
+    mutate(quadrat_fix = if_else(tag %in% outside_col1_tags, quadrat[2],NA),
+           quadrat_fix = if_else(n_distinct(quadrat) != 1 & is.na(quadrat_fix), quadrat[1],quadrat_fix),
+           quadrat = if_else(!is.na(quadrat_fix),quadrat_fix,quadrat))  %>%  #fix multistems that have multiple quadrats 
+    ungroup()  %>% 
+    mutate(textflag = if_else(grepl(pattern = ptrn,tolower(notes_current)),T,F),
+            true_sp = if_else(textflag,str_extract(tolower(notes_current),paste0("(is) (",paste(scbi.spptable$sp, collapse = "|"),")"), group = 2),NA),
+            true_sp = tolower(if_else(!is.na(true_sp),true_sp,sp)),
+            Status = if_else(status_current %in% c("LI"),"alive","dead"),
+            True_quadrat = sprintf("%04d", quadrat)) ###### add taper correction if needed
+
+lookup_2023 <- census2023_corrections  %>% 
[…]

### 2024-01-24 code/figure2_agb.R

@@ -77,7 +77,8 @@ useCensus2023$Calculated_ABG <- get_biomass(dbh = as.numeric(useCensus2023$dbh_c
 calc_quadrat_abg <- function(abg_df, year){
   outdf <- abg_df  %>% 
     group_by(quadrat)  %>% 
-    summarize(Abg_C_Mg = sum(Calculated_ABG,na.rm = T)/1000 * .47)  %>% 
+    summarize(Abg_C_Mg = sum(Calculated_ABG,na.rm = T)/1000 * .47,
+              Abg_Mg = sum(Calculated_ABG,na.rm = T)/1000)  %>% 
     left_join(grouped_quadrats  %>% select(quadrat,Group))  %>% 
     mutate(Year = year)
   return(outdf)
@@ -92,25 +93,38 @@ quadrat_abg_2023 <- calc_quadrat_abg(useCensus2023,2023)  %>%
 group_trends <- quadrat_abg_2008  %>% 
   bind_rows(quadrat_abg_2013,quadrat_abg_2018,quadrat_abg_2023)  %>% 
   group_by(Year,Group)  %>% 
-  summarize(n_quads = n(),Abg_C_Mg_Ha = sum(Abg_C_Mg) / ((400 * n_quads)/10000))  %>% 
+  summarize(n_quads = n(),Abg_C_Mg_Ha = sum(Abg_C_Mg) / ((400 * n_quads)/10000),
+            Abg_Mg_Ha = sum(Abg_Mg) / ((400 * n_quads)/10000))  %>% 
   drop_na()  %>% 
   mutate(Group = as.character(Group))
 tot_trends <- quadrat_abg_2008  %>% 
   bind_rows(quadrat_abg_2013,quadrat_abg_2018,quadrat_abg_2023)  %>% 
   group_by(Year)  %>% 
-  summarize(n_quads = n(),Abg_C_Mg_Ha = sum(Abg_C_Mg) / ((400 * n_quads)/10000))  %>% 
+  summarize(n_quads = n(),Abg_C_Mg_Ha = sum(Abg_C_Mg) / ((400 * n_quads)/10000),
+            Abg_Mg_Ha = sum(Abg_Mg) / ((400 * n_quads)/10000))  %>% 
   mutate(Group = "Whole Plot")
 
 figure2_agb <- group_trends  %>% 
   bind_rows(tot_trends)
- 
[…]

### 2024-01-24 code/figure3_woodyGrowthDataPrep- temporary.R

@@ -1,0 +1,129 @@
+#Figure 3 (Woody Growth)
+
+###Draft 11
+
+#Look at the "Mortality, Growth and Recruitment - Conceptual Differences" word document if you're confused
+
+#Load libraries 
+library(allodb)
+library(dplyr)
+library(tidyverse)
+
+#Read in data from past censuses and the species table
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+#Read in data from 2023
+Census_2023 <- read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))
+#Rename the data
+Census_2008 <- scbi.stem1
+Census_2013 <- scbi.stem2
+Census_2018 <- scbi.stem3
+Species_table <- scbi.spptable
+
+#Combine the census data and species list
+Census_2008_Species <- merge(Census_2008, Species_table, by ="sp")
+Census_2013_Species <- merge(Census_2013, Species_table, by ="sp")
+Census_2018_Species <- merge(Census_2018, Species_table, by ="sp")
+Census_2023_Species <- merge(Census_2023, Species_table, by ="sp")
+
+#Define finished quadrats
+Finished_quadrats <- unique(Census_2023_Species$quadrat)
+
+##Use this line if you're interested in the biomass for the whole plot = Finished_quadrats <- unique(Census_2018$quadrat)
+#subset(Census_2023_Species, Census_2023_Species$quadrat %in% Finished_quadrats)
[…]

### 2024-01-24 code/figure2_agb.R
update Figure 2 with corrected censuses

@@ -13,15 +13,15 @@ library(ggplot2)
 #Read in data from past censuses and the species table
 grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
-load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem1.rdata"))
-load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem2.rdata"))
-load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
 
 #Rename the data
-census2008 <- scbi.stem1
-census2013 <- scbi.stem2
-census2018 <- scbi.stem3
+census2008 <- scbi.stem1.corrected
+census2013 <- scbi.stem2.corrected
+census2018 <- scbi.stem3.corrected
 speciesTable <- scbi.spptable
 census2023 <- read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))
 

### 2024-01-25 code/difference_measured_vs_tapered.R
Create difference_measured_vs_tapered.R
Preliminary plot comparing differences in diameter for 3 different sets of DBHs

@@ -1,0 +1,33 @@
+library(ggplot2)
+library(allodb)
+library(dplyr)
+library(tidyverse)
+
+# Sample data (replace with your actual data)
+set1 <- data.frame(
+  TreeID = 1:5,
+  Genus = c("Acer", "Acer", "Quercus", "Quercus", "Platinus"),
+  Species = c("rubrum", "rubrum", "alba", "rubra", "occidentalis"),
+  Measured_DBH = c(10, 15, 20, 25, 30),
+  IKTaper_Corrected_DBH = c(11, 16, 21, 26, 31),
+  BCITaper_Corrected_DBH = c(9, 14, 19, 24, 29)
+)
+
+head(set1)
+
+long_data <- set1 %>%
+  pivot_longer(cols = c(IKTaper_Corrected_DBH, BCITaper_Corrected_DBH))
+
+
+
+ggplot(long_data, aes(x = Measured_DBH, y = value, group = name)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  geom_abline(intercept = 0, slope = 1)
+
+
+
+ggplot(long_data, aes(x = Measured_DBH, y = value, group = name)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  geom_abline(intercept = 0, slope = 1)

### 2024-02-05 code/0_Clean_2023_Census.r
more data cleaning

@@ -3,56 +3,74 @@ library(tidyverse)
 ##### load in SCBI 2023 census & SCBI species table from Github #####
 census2023 <- read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+load("./data/taper_coefficients.rdata")
 
 ##### Correction functions #####
-create_stem_UID <- function(census_df,TID_col = "tag",SID_col = "StemTag"){
-    census_df[["UID"]] <- paste(census_df[[TID_col]],census_df[[SID_col]],sep = "_")
-    return(census_df)
+create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
+  census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
+  return(census_df)
 }
 
-mode <- function(codes){
+usfs_taper <- function(dh, h, stump_taper, stem_taper) { #
+  dbh <- dplyr::if_else(h >= 1.3, dh - stem_taper * (h - 1.3),
+                        dplyr::if_else(h < 1.3, dh - stump_taper * (h - 1.3), NA))
+  dbhcm <- dbh * 100
+  return(dbhcm)
+}
+
+mode <- function(codes) {
   names(which.max(table(codes)))
 }
 
 ##### Clean up 2023 census #####
-sp_correct <- census2023[grep(pattern = "is not",tolower(census2023$notes_current)),]
-ptrn <- paste(c("species","Species",paste0("(is ",scbi.spptable$sp,")")), collapse = "|")
-wrong_mainstem_quadrat <- c(33561,53854, 53855)
+sp_correct <- census2023[grep(pattern = "is not", tolower(census2023$notes_current)), ]
+ptrn <- paste(c("species", "Species", paste0("(is ", scbi.spptable$sp, ")")), collapse = "|")
+wrong_mainstem_quadrat <- c(33561, 53854, 53855)
+
+### correct missing hom, flag for species corrections in notes, correct species,  & apply taper corrections to those with HoM != 1.3m ###
+census2023_hom_corrections <- census2023  %>% 
+  mutate(textflag = if_else(grepl(pattern = ptrn, tolower(notes_current)), TRUE, FALSE),
[…]

### 2024-02-05 code/0_Tapering_testing.r

@@ -17,9 +17,12 @@ avgStumpTaper <- weighted.mean(x = stemTaper$Stump.taper.rate, w = stemTaper$n)
 fullspcd <- stemTaper  %>% 
   mutate(spcd = str_split(str_remove(Species.Code, pattern = " "), pattern = ","))  %>%
   unnest(spcd) %>%
+  mutate(spcd = tolower(str_remove(spcd,pattern = " ")))  %>% 
   select(spcd, Stem.taper.rate, Stump.taper.rate)  %>%
   add_row(spcd  = "other", Stem.taper.rate = avgStemTaper, Stump.taper.rate = avgStumpTaper)
 
+# scbi_taper_coefficients <- fullspcd
+# save(scbi_taper_coefficients,file =  "data/taper_coefficients.rdata")
 usfs_taper <- function(dh, h, stump_taper, stem_taper) {
   dbh <- if_else(h > 1.3, dh - stem_taper * (h - 1.3),
                  if_else(h < 1.3, dh - stump_taper * (h - 1.3), NA))
@@ -34,13 +37,6 @@ bci_taper <- function(dbh, hom) { #taper from Cushman 2014
 }
 
 latlong <- c(-78.1454, 38.8935)
-Census2023  %>% 
-# filter(!(hom %in% c(1.3)) & !is.na(hom))  %>% 
- mutate(hom = if_else(hom %in% c(0) & status_current %in% c("LI"), 1.3, hom))  %>% 
- filter(hom %in% c(0))  %>% 
- pull(status_current)  %>% 
- table()
- View()
 
 multi_dbhs <- Census2023  %>% 
   left_join(scbi.spptable)  %>% 
@@ -56,7 +52,7 @@ multi_dbhs <- Census2023  %>%
     USFS_TC_AGB = allodb::get_biomass(dbh = USFS_Taper_Corrected_DBH,genus = Genus, species = Species, coords = latlong),
     BCI_TC_AGB = allodb::get_biomass(dbh = BCI_Taper_Corrected_DBH,genus = Genus, species = Species, coords = latlong))  %>%
   #select(dbh_current,Measured_DBH, USFS_Taper_Corrected_DBH,BCI_Taper_Corrected_DBH,hom)       
-  select(uncorrected_AGB, USFS_TC_AGB, BCI_TC_AGB, hom)  
[…]

### 2024-02-05 code/0_Update_previous_censuses.r

@@ -7,40 +7,40 @@ load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/census2023.lookup.t
 
 
 ##### Correction functions #####
-create_stem_UID <- function(census_df,TID_col = "tag",SID_col = "StemTag"){
-    census_df[["UID"]] <- paste(census_df[[TID_col]],census_df[[SID_col]],sep = "_")
-    return(census_df)
+create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
+  census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
+  return(census_df)
 }
 
-all_previous_census_checks <- function(census_df){
-    census_df  %>% 
-        mutate(live_fix = if_else(Status == "alive" & DFstatus != "alive" & DFstatus != "prior",T,F),
-               sp_fix = if_else(sp != true_sp, T,F),
-                quad_fix = if_else(true_quadrat != quadrat,T,F))   %>% 
-        mutate(DFstatus = if_else(live_fix, Status,DFstatus),
-               sp = if_else(sp_fix,true_sp,sp),
-               quadrat = if_else(quad_fix,true_quadrat,quadrat))  %>% 
-        select(UID,sp,quadrat,dbh,hom,ExactDate,DFstatus)
+all_previous_census_checks <- function(census_df) {
+  census_df  %>% 
+    mutate(live_fix = if_else(Status %in% "alive" & !(DFstatus %in% "alive") & !(DFstatus %in% "prior"), TRUE, FALSE),
+           sp_fix = if_else(sp != true_sp, TRUE, FALSE),
+           quad_fix = if_else(true_quadrat != quadrat, TRUE, FALSE))   %>% 
+    mutate(DFstatus = if_else(live_fix, Status, DFstatus),
+           sp = if_else(sp_fix, true_sp, sp),
+           quadrat = if_else(quad_fix, true_quadrat, quadrat))  %>% 
+    select(UID, sp, quadrat, dbh, hom, ExactDate, DFstatus)
 }
 
 
 ##### correct & save previous census data #####
 
 scbi.stem1.corrected <- scbi.stem1 %>% 
-    create_stem_UID()  %>% 
-    left_join(lookup_2023, by = c("UID"))   %>% 
-    all_checks()
+  create_stem_UID()  %>% 
[…]

### 2024-02-06 code/minor_data_fix_testing.r
hom fixes, multistem tag corrections

@@ -1,0 +1,77 @@
+census2023  %>% 
+#    filter(is.na(hom) & status_current %in% c("LI")) %>% 
+    filter(is.na(hom) & status_current %in% c("LI") & status_previous %in% c("G")) %>% 
+   #filter(!(is.na(hom_alert)| hom_alert %in% "" )& status_current %in% c("LI")) %>% 
+    #filter(hom %in% c(0)  & status_current %in% c("LI"))  %>% 
+#    filter(table %in% c("old_trees") & status_current %in% c("LI"))  %>% 
+    select(tag, table, quadrat, StemTag,dbh_current , status_current, status_previous,hom,hom_alert,notes_current)  %>% 
+    #nrow()
+#    pull(status_previous)  %>% 
+ #   table()
+    #write.csv(file = "C:/Work/Smithsonian/Repos/15yrsChange/data/HoM_fixes_prevstatusA.csv")
+    View()
+
+##### check for DBH < 10 (now flagged in github QAQC) #####
+census2023  %>% 
+    filter(status_current %in% c("LI") & dbh_current < 10)  %>% 
+    select(table,tag, StemTag, quadrat, dbh_current, dbh_previous,status_previous, DBH.check,personnel,codes_current)  %>% 
+    View()
+
+##### check for issues where stam tags are non sequential because secondary stem was added with wrong stem ID (OR issue is from previous census ) #####
+
+st_diff <- census2023  %>% 
+    group_by(tag)  %>% 
+    arrange(StemTag)  %>% 
+    mutate(stemtag_diff = StemTag - lag(StemTag))  %>% 
+  #  select(tag,StemTag,status_current,stemtag_diff)  %>% 
+    
+    filter(stemtag_diff > 1)   
+View(st_diff)
+#write.csv(st_diff,file = "C:/Work/Smithsonian/Repos/15yrsChange/data/nonsequential_stemtags.csv")
+census2023  %>% 
+    filter(tag %in% st_diff$tag)  %>% 
+    arrange(tag,StemTag)  %>% 
+    View()
+
+### check which are issues from previous census ###
+ load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
+
+stem3.diff <- scbi.stem3  %>%  
+   group_by(tag)  %>% 
+    arrange(StemTag)  %>% 
[…]

### 2024-02-06 doc/15yrsChange.Rmd
New Section Headings and Initial Drafts of Paragraphs

@@
 
-# Abstract  {-#Abstract}
+*(Word limit: 6,500 words)*
+
+# Summary  {-#Abstract}
+
+*Summary formats tend to be in a few long bullet points
 
@@
 
-*(describe location, forest type, etc)*
+The study was conducted at The Smithsonian Conservation Biology Institute’s (SCBI) large forest dynamics plot in Front Royal, Virginia, USA (38.8935°, -78.1454°; Bourg et al.
+2013, Anderson-Teixeira et al.
+2015).
+*A sentence about the region and that it is adjacent to the northern entrance of Shenandoah National Park.* The plot is 25.6 ha (640 x 400 m) and includes a 4-ha, 2.4 m high deer exclusion fence, which has decreased the number/decreased the presence of white-tailed deer since 1990 (McShea 2000).
+It consists of a mature secondary eastern deciduous forest that developed following logging and agricultural use in the mid-19th century.
+The species composition reflects what is typical of this forest type, with a canopy that is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*), and the understory that is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*)
+
+*Which definition of forest type is more accurate - mature, mixed deciduous forest or mature secondary eastern deciduous forest?*
+
+*Need to add the Anderson-Teixeira et al. 2015 reference to references page*
 
@@
 
+Data was collected according to the standard ForestGEO protocol, outlined in (Condit 1998).
+A census crew measured the diameter at breast height (DBH), mapped, and tagged every stem on the plot with a DBH greater than or equal to 1 cm.
+DBH was recorded only for living trees.
+If there were multiple stems on a tree, each stem was measured and given a stem number and tag.
+
 *(describe mortality census)*
+
+In addition to the ForestGEO census, an annual [official name of mortality survey] is conducted on the large forest dynamics plot.
+The survey assesses the health of individual trees and identifies factors that contribute to mortality.
+The survey was initiated in 2008, and has been conducted every year since 2013.
+(source).
+The field crew measured the percentage of the crown that was alive and intact, and noted any factors associated with death (i.e.
+wounds) present on the tree.
+Numerous nuisance species and pathogens, including the emerald ash borer, white-tailed deer, and dutch elm disease have been recorded in the region prior to the start of the mortality survey (Herms & McCullough, 2014, McShea 2000,NEED ELM SOURCE).
 
@@
 *(describe seedlings survey, if used)*
+
+*(describe QA/QC and technical workflow)*
+
+Field data was collected on Apple iPads using the ESRI ArcGIS Field Maps application.
+To improve the quality of the data, GitHub Actions continuous integration was implemented for quality control (Kim et al., 2022).
+Data from the previous census was used to populate the Field Maps app, and fields were created to input data.
+There were checks built into the app to help catch errors, for example the DBH field would flag a suspiciously high number and would not allow the form to be submitted if it was blank.
+To use the app in the field the census crew downloaded offline areas of where they were working and synced them to an online map after collecting data.
+Next the data was downloaded from ArcGIS online and pushed to GitHub, where it triggered a GitHub Action running data quality assurance and control (QA/QC) scripts.
+The scripts were written in R and checked the data for errors such as “suspicious positive growth.” After going through the QA/QC process, a csv of errors and warnings with the associated stem were output which could then be corrected in the field.
 
@@
 
-# Conclusions (optional)
+# Conclusions (optional) *(No Conclusions section listed in Submission Guidelines)*
 
@@
 
-# Conflict of Interest statement {-#Conflict}
+# Competing Interests {-#Conflict}
 The authors declare no conflict of interest.
 
-# Authors' contributions {-#Authors}
+# Author Contributions {-#Authors}
 *[Name of author 1] and [Name of author 2] conceived the ideas and designed methodology; [Name of author 1] and [Name of author 3] collected the data; [Name of author 2] and [Name of author 4] analysed the data; [Name of author 1] and [Name of author 4] led the writing of the manuscript.
 All authors contributed critically to the drafts and gave final approval for publication.*
+
+# Data availability {-#Data}
 
@@
 
-# Data availability {-#Data}
+# Supporting Information

### 2024-02-08 doc/15yrsChange.Rmd
Add Outline of Recruitment Failure

@@
 
-*Summary formats tend to be in a few long bullet points
+*Summary format for New Phytologist tends to be in a few long bullet points*
 
@@
 
+Recruitment Failure Paragraph Outline
+
+- Nuisance species have a profound impact on regional ecosystems
+- There is a documented overabundance of white-tailed deer (*Odocoileus virginianus*) (*Publication referenced in McGravey et al.
+2013*) which can be attributed to human influence (*Publication referenced in McGravey et al 2013*)
+- White-tailed deer consume seedlings and saplings, negatively impacting seedling and sapling survival and density (*Publications referenced in McGravey et al. 2013*)
+- Define or contextualize non-endemic plants (Look for a reference)
+  - Working definition: Plants that were not historically present in an area, region or ecosystem or were found at different abundances or densities
+- Non-endemic seedlings and saplings may not be consumed at similar rates as endemic plants (alluded to in *McGravey et al.
+2013*) and may be able to utilize niche space in a forest faster or more efficiently (Reference)
+- Additional information on the effects of non-endemic species
+- As a result, they may be able to out compete endemic species
+- These nuisance species can contribute to recruitment failure
+  - Consider defining and contextualizing recruitment failure a little more here
+  - The effects of recruitment debt on mature tree species composition and forest structure are often not apparent for decades (*McGravey et al.
+2013* - potentially move to regeneration debt paragraph)
+
 - white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence
@@
 
+*What is the best way to describe the forest type in our study site?
+I've seen "mixed mature deciduous forest", "mature secondary eastern deciduous forest", and "mesic temperate deciduous forest".
+I want to make sure I'm being consistent.*
+
 The study was conducted at The Smithsonian Conservation Biology Institute’s (SCBI) large forest dynamics plot in Front Royal, Virginia, USA (38.8935°, -78.1454°; Bourg et al.
@@
 It consists of a mature secondary eastern deciduous forest that developed following logging and agricultural use in the mid-19th century.
-The species composition reflects what is typical of this forest type, with a canopy that is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*), and the understory that is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*)
+*Canopy trees are 84-124 years old (Thompson and McGarvey - unpublished data referenced in McGarvey 2013 - see if you can find the published version of this data)* The species composition reflects what is typical of this forest type, with a canopy that is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*), and the understory that is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*)
 

### 2024-02-09 doc/15yrsChange.Rmd
Update Outline and Add Relevant Literature Notes

--- 2024-02-06 doc/15yrsChange.Rmd
@@
 
-Non-endemic pests and pathogens facilitate invasin of non-endemic plants [@guo_impacts_2023 and refs therein].
+Non-endemic pests and pathogens facilitate invasion of non-endemic plants [@guo_impacts_2023 and refs therein].
 
 ## (paragraph on recruitment failure because of deer and non-endemic plants)
+
+Recruitment Failure Paragraph Outline
+
+- Over-abundant herbivores can have a profound impact on carbon cycling in temperate deciduous forests
+- There is a documented overabundance of white-tailed deer (*Odocoileus virginianus*) (*Publication referenced in McGravey et al.
+2013*) which can be attributed to human influence (*Publication referenced in McGravey et al 2013*)
+- White-tailed deer consume seedlings and saplings, negatively impacting seedling and sapling survival, density and growth (*Publications referenced in McGravey et al. 2013*, *Holm et al. 2013*)
+- This can lead to decreased understory diversity, species richness and decreased abundance of dominant species in a typical forest (in this case Quercus spp.) (*Holm et al. 2013*)
+- Define or contextualize non-endemic plants (Look for a reference)
+  - Working definition: Plants that were not historically present in an area, region or ecosystem or were found at different abundances or densities
+- Deer selectively browse on palatable species, and some non-endemic species, including pawpaw, are considered non-palatable by deer (*McGravey et al.
+2013*)  This can create dense stands of non-endemic species (*Knauer et al.
+2023*)
+- Non-endemic species may be able to utilize niche space in a forest faster or more efficiently (Reference - considered common knowledge?)
+- As a result, they may be able to out compete endemic species
+- These nuisance species can contribute to recruitment failure
+  - Consider defining and contextualizing recruitment failure a little more here
+  - The effects of these recruitment failure on species composition and forest structure are often not apparent for decades (*McGravey et al. 2013* - potentially move to regeneration debt paragraph).
 

### 2024-02-09 doc/15yrsChange.Rmd
Add Recruitment Failure Paragraph

--- 2024-02-06 doc/15yrsChange.Rmd
@@
 
-Non-endemic pests and pathogens facilitate invasin of non-endemic plants [@guo_impacts_2023 and refs therein].
+Non-endemic pests and pathogens facilitate invasion of non-endemic plants [@guo_impacts_2023 and refs therein].
 
 ## (paragraph on recruitment failure because of deer and non-endemic plants)
+
+Recruitment Failure Paragraph
+
+White-tailed deer (Odocoileus virginianus) represent an endemic nuisance species prevalent in eastern deciduous forests.
+Their populations have substantially increased in the past 50 years (McShea et al.
+1997), due in large part to human influence.
+(Found in McGarvey et al.
+2013 – Brown et al.
+2000, Côté et al.
+2004, Rooney 2001).
+White-tailed deer preferentially browse on woody herbaceous species in their earliest life stages (McGarvey et al.
+2013), with overabundant herbivory negatively impacting seedling and sapling survival, growth and density (Found in McGarvey et al.
+2013 – Dzieciolowski 1980, Gill and Beardall 2001, Healy 1997, Konig 1976, Putman et al.
+1989).
+Models predict that chronic overabundance will reduce understory diversity and decrease the abundance of traditionally dominant species (Holm et al.
+2013).
+Due to climate change and human activity, many non-endemic plant species are being introduced or number in forest ecosystems (Reference).
+Deer find many of these species, including pawpaw (Asimina triloba), to be unpalatable (Found in McGarvey 2013 – Asnani 2006), and do not consume them at the same rates relative to their native counterparts (Reference), enabling them to form dense stands in forest understories (Found in Knauer et al.
+2023 – Horsley and Marquis 1983; Stromayer and Warren 1997; Royo and Carson 2006).
+This lack of browsing pressure, coupled with characteristics such as fast growth rate and greater adaptability to altered conditions precipitated by climate change, allow nuisance plant species to outcompete other species in the understory (Reference).
+The interaction of these nuisance species on the landscape contributes to a recruitment failure in forests, [*Where individuals from certain tree species do not enter the mature population - Do we need to define recruitment failure, or is that considered common knowledge?
+If so, reference*).
+The extent of the consequences derived from recruitment failure on species composition and forest structure is often not apparent for years (McGarvey et al.
+2013)
 

### 2024-02-13 doc/15yrsChange.Rmd
Add Study Site, Census and Mortality Descriptions

@@
 
-## Study site
+## Site Description
 
@@
 
-The study was conducted at The Smithsonian Conservation Biology Institute’s (SCBI) large forest dynamics plot in Front Royal, Virginia, USA (38.8935°, -78.1454°; Bourg et al.
-2013, Anderson-Teixeira et al.
-2015).
-*A sentence about the region and that it is adjacent to the northern entrance of Shenandoah National Park.* The plot is 25.6 ha (640 x 400 m) and includes a 4-ha, 2.4 m high deer exclusion fence, which has decreased the number/decreased the presence of white-tailed deer since 1990 (McShea 2000).
-It consists of a mature secondary eastern deciduous forest that developed following logging and agricultural use in the mid-19th century.
-*Canopy trees are 84-124 years old (Thompson and McGarvey - unpublished data referenced in McGarvey 2013 - see if you can find the published version of this data)* The species composition reflects what is typical of this forest type, with a canopy that is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*), and the understory that is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*)
+*Need to add the Anderson-Teixeira et al. 2015 reference to references page*
 
-*Which definition of forest type is more accurate - mature, mixed deciduous forest or mature secondary eastern deciduous forest?*
-
-*Need to add the Anderson-Teixeira et al. 2015 reference to references page*
+The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).
+The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of secondary eastern mixed deciduous forest.
+Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description).
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al.
+2013).
+Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).
+The plot, which includes a 4-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
+It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the World.
 
@@
 
-*(describe main census & years occurred)*
+*(Describe main census & years occurred)*
 
-Data was collected according to the standard ForestGEO protocol, outlined in (Condit 1998).
-A census crew measured the diameter at breast height (DBH), mapped, and tagged every stem on the plot with a DBH greater than or equal to 1 cm.
-DBH was recorded only for living trees.
-If there were multiple stems on a tree, each stem was measured and given a stem number and tag.
+A number of surveys are regularly conducted at the site, four of which are included in our analysis.
+As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).
+This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter at 1.3 meters in height (referred to as diameter at breast height or dbh).
+The census records information regarding the dbh, species, status and location of each stem included in the survey.
+Each plant is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field.
+For multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.
+The location of each individual within its respective 20 x 20 meter quadrat is recorded on a map of the plot on ArcGIS FieldMaps.
+Established in 2008, there have been 4 censuses at the site, comprising 20 years of detailed forest dynamics data.
 
-*(describe mortality census)*
+*Is dbh considered common knowledge? Does it need to be defined?*
 
-In addition to the ForestGEO census, an annual [official name of mortality survey] is conducted on the large forest dynamics plot.
-The survey assesses the health of individual trees and identifies factors that contribute to mortality.
-The survey was initiated in 2008, and has been conducted every year since 2013.
-(source).
-The field crew measured the percentage of the crown that was alive and intact, and noted any factors associated with death (i.e.
-wounds) present on the tree.
-Numerous nuisance species and pathogens, including the emerald ash borer, white-tailed deer, and dutch elm disease have been recorded in the region prior to the start of the mortality survey (Herms & McCullough, 2014, McShea 2000,NEED ELM SOURCE).
+*Check with Luca to use consistent terminology here and in the analysis portion of the methods for ID tag, stem tag, etc.*
 
-*(describe invasive plants survey)*
+*(Describe mortality census and invasive plant survey)*
+
+In addition to the ForestGEO census, an annual mortality survey is conducted on the site.
+Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.
+Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
+Furthermore, an invasive plant survey is conducted in conjunction with [the mortality survey or the census].
+This study evaluates the establishment of non-endemic plant species and patterns of spatial distribution.
+Quadrats are visually examined for non-endemic plant species, with an estimation of the area covered by each species represented on a 1-5 numeric scale.
+
+*Is the invasive plant survey conducted every year or every five years?*
 
@@
 
-Field data was collected on Apple iPads using the ESRI ArcGIS Field Maps application.
 To improve the quality of the data, GitHub Actions continuous integration was implemented for quality control (Kim et al., 2022).

### 2024-02-13 doc/15yrsChange.Rmd
Remove Invasive Species Survey

@@
 
-*(Describe main census & years occurred)*
+*(Describe main census and mortality survey)*
 
-A number of surveys are regularly conducted at the site, four of which are included in our analysis.
+A number of surveys are regularly conducted at the site, two of which are included in our analysis.
 As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).
@@
 Established in 2008, there have been 4 censuses at the site, comprising 20 years of detailed forest dynamics data.
+In addition to the ForestGEO census, an annual mortality survey is conducted on the site.
+Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.
+Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
 
@@
 *Check with Luca to use consistent terminology here and in the analysis portion of the methods for ID tag, stem tag, etc.*
-
-*(Describe mortality census and invasive plant survey)*
-
-In addition to the ForestGEO census, an annual mortality survey is conducted on the site.
-Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.
-Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
-Furthermore, an invasive plant survey is conducted in conjunction with [the mortality survey or the census].
-This study evaluates the establishment of non-endemic plant species and patterns of spatial distribution.
-Quadrats are visually examined for non-endemic plant species, with an estimation of the area covered by each species represented on a 1-5 numeric scale.
-
-*Is the invasive plant survey conducted every year or every five years?*
 

### 2024-02-16 doc/15yrsChange.Rmd
General Notes and Pests/Pathogens Outline

@@
 
+Two Paragraphs
+
+1. The global implications of pests and pathogens
+2. History of pests and pathogens in eastern deciduous forests
+  1. Chestnut blight
+  2. Dutch elm disease
+  3. Hemlock Wooly Adelgid
+  4. Emerald Ash-borer
+
 Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
@@
 
-White-tailed deer (Odocoileus virginianus) represent an endemic nuisance species prevalent in eastern deciduous forests.
-Their populations have substantially increased in the past 50 years (McShea et al.
-1997), due in large part to human influence.
+White-tailed deer (*Odocoileus virginianus*) are a nuisance species prevalent in eastern deciduous forests.*Start with a stronger sentence about the impact of deer on herbivory - Ex.
+White-tailed deer are the dominant herbivore in eastern deciduous forests (Reference)* Their populations have dramatically increased in the past 50 years (McShea et al.
+1997).
+*Find a reference with a specific number of increase in deer populations* due in large part to global change drivers.
 (Found in McGarvey et al.
@@
 2004, Rooney 2001).
-White-tailed deer preferentially browse on woody herbaceous species in their earliest life stages (McGarvey et al.
-2013), with overabundant herbivory negatively impacting seedling and sapling survival, growth and density (Found in McGarvey et al.
+White-tailed deer preferentially browse on woody plant species in their earliest life stages (McGarvey et al.
+2013), with overabundant herbivory reduce seedling and sapling survival, growth and density (Found in McGarvey et al.
 2013 – Dzieciolowski 1980, Gill and Beardall 2001, Healy 1997, Konig 1976, Putman et al.
 1989).
-Models predict that chronic overabundance will reduce understory diversity and decrease the abundance of traditionally dominant species (Holm et al.
-2013).
+Chronic overabundance has been shown to reduce understory diversity and decrease the abundance of traditionally dominant species (*NEW CITATION*).
 Due to climate change and human activity, many non-endemic plant species are being introduced or number in forest ecosystems (Reference).
@@
 
-*What is the best way to describe the forest type in our study site?
-I've seen "mixed mature deciduous forest", "mature secondary eastern deciduous forest", and "mesic temperate deciduous forest".
-I want to make sure I'm being consistent.*
-
 *Need to add the Anderson-Teixeira et al. 2015 reference to references page*
@@
 The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).
-The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of secondary eastern mixed deciduous forest.
+The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
 Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description).
@@
 It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the World.
+
+*Add information about the history of pests and pathogens at this site and at similar forests*
 
@@
 As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).
-This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter at 1.3 meters in height (referred to as diameter at breast height or dbh).
+This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; hereafter referred to as dbh).
 The census records information regarding the dbh, species, status and location of each stem included in the survey.
@@
 
-*Is dbh considered common knowledge? Does it need to be defined?*
+*Add Condit 1998 in references*
 
-*Check with Luca to use consistent terminology here and in the analysis portion of the methods for ID tag, stem tag, etc.*
+*In response to the presence of Emerald Ashborer, a mortality survey was initiated, then improved in year [X].
 

### 2024-02-20 code/0_Update_previous_censuses.r
flux calculations and figures

@@ -4,6 +4,7 @@ load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.stem3.rdata"))
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
 load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/census2023.lookup.table.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem4.corrected.rdata")
 
 
 ##### Correction functions #####
@@ -12,33 +13,49 @@ create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
   return(census_df)
 }
 
+stem_checker <- function(census_df, stem_UID){
+  census_df  %>%
+    create_stem_UID()  %>% 
+    filter(UID %in% stem_UID)
+}
+
+# stem_checker(scbi.stem1, "50567_1")
+# stem_checker(scbi.stem2, "50567_1")
+# stem_checker(scbi.stem3, "50567_1")
+# stem_checker(scbi.stem4.corrected,"50567_1")
+
 all_previous_census_checks <- function(census_df) {
-  census_df  %>% 
+  check_df <- census_df  %>% 
     mutate(live_fix = if_else(Status %in% "alive" & !(DFstatus %in% "alive") & !(DFstatus %in% "prior"), TRUE, FALSE),
            sp_fix = if_else(sp != true_sp, TRUE, FALSE),
            quad_fix = if_else(true_quadrat != quadrat, TRUE, FALSE))   %>% 
     mutate(DFstatus = if_else(live_fix, Status, DFstatus),
            sp = if_else(sp_fix, true_sp, sp),
-           quadrat = if_else(quad_fix, true_quadrat, quadrat))  %>% 
+           quadrat = if_else(quad_fix, true_quadrat, quadrat))  
+  check_counts <- check_df  %>% 
+    count(sp_fix,quad_fix, live_fix) #dataframe recording number of fixes
+  print(check_counts)
+  out_df <- check_df %>% 
     select(UID, sp, quadrat, dbh, hom, ExactDate, DFstatus)
[…]

### 2024-02-20 code/Woodyflux_Calculations.r

@@ -1,0 +1,93 @@
+library(tidyverse)
+library(allodb)
+
+##### helper functions #####
+create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
+  census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
+  return(census_df)
+}
+
+##### Read in data from ALL censuses and the species table #####
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem4.corrected.rdata")
+
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+
+latlong <- c(-78.1454, 38.8935)
+
+scbi.stem1.corrected <- scbi.stem1.corrected  %>% 
+  mutate(Census = 1)
+scbi.stem2.corrected <- scbi.stem2.corrected  %>% 
+  mutate(Census = 2)
+scbi.stem3.corrected <- scbi.stem3.corrected  %>% 
+  mutate(Census = 3)
+scbi.stem4.corrected <- scbi.stem4.corrected  %>% 
+  mutate(Census = 4, hom = as.character(hom), dbh = as.character(dbh))
+
+##### Combine the census data & calculate woody fluxes (AWP, AWM, AWR) #####
+all_censuses <- scbi.stem1.corrected  %>% 
+  bind_rows(scbi.stem2.corrected, scbi.stem3.corrected, scbi.stem4.corrected)  %>% 
+  left_join(scbi.spptable, by = c("sp" = "sp"))  %>% 
+  mutate(ExactDate = as.Date(ExactDate, format = "%Y-%m-%d"),
[…]

### 2024-02-20 code/exploratory/figure3_woodyGrowthDataPrep- temporary.R

@@ -1,0 +1,129 @@
+#Figure 3 (Woody Growth)
+
+library(allodb)
+library(dplyr)
+library(tidyverse)
+
+##### helper functions #####
+create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
+  census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
+  return(census_df)
+}
+
+#Read in data from ALL censuses and the species table
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem4.corrected.rdata")
+
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+
+#Rename the data
+Census_2008 <- scbi.stem1.corrected
+Census_2013 <- scbi.stem2.corrected
+Census_2018 <- scbi.stem3.corrected
+Census_2023 <- scbi.stem4.corrected
+Species_table <- scbi.spptable
+
+#Combine the census data and species list
+Census_2008_Species <- merge(Census_2008, Species_table, by ="sp")
+Census_2013_Species <- merge(Census_2013, Species_table, by ="sp")
+Census_2018_Species <- merge(Census_2018, Species_table, by ="sp")
+Census_2023_Species <- merge(Census_2023, Species_table, by ="sp")
+
+##### Calculate biomass for the Censuses #####
+#Plot size & location
+latlong <- c(-78.1454, 38.8935)
+Hectares_measured <- 25.6
+
+#Reclass date column
+Census_2008_Finished <- Census_2008_Species  %>% 
+  mutate(ExactDate = as.Date(ExactDate, format = "%Y-%m-%d"))  
[…]

### 2024-02-20 code/figure3_output.R

@@ -1,21 +1,0 @@
-library(ggplot2)
-
-##set your own wd##
-setwd("/Users/krystalbagnaschi/Documents/15yrsChange/")
-
-awrtable <- read.csv("doc/figure3_biomassRecruited.csv")
-awmtable <- read.csv("doc/figure3_mortalityFlux.csv")
-awptable <- read.csv("doc/figure3_woodGrowthFlux.csv")
-
-
-##colnames(awrtable) <- c(colnames(awrtable)[1:4],"Value")
-##colnames(awmtable) <- c(colnames(awmtable)[1:4],"Value")
-
-#awrtable[3,4] <- "2018to2023"
-
-biomassfluxtable <- rbind(awrtable, awmtable, awptable)
-
-ggplot(biomassfluxtable, aes(x=Interval, y=Value, group=Flux, col=Flux)) +
-  geom_line()
-
-ggsave("figure3.png", width = 7, height = 7, units = "in", dpi = 300, path = "doc")

### 2024-02-20 code/figure3_woodyGrowthDataPrep- temporary.R

@@ -1,130 +1,0 @@
-#Figure 3 (Woody Growth)
-
-###Draft 11
-
-#Look at the "Mortality, Growth and Recruitment - Conceptual Differences" word document if you're confused
-
-#Load libraries 
-library(allodb)
-library(dplyr)
-library(tidyverse)
-
-#Read in data from past censuses and the species table
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
-  mutate(quadrat = sprintf("%04d",quadrat))
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
-load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
-#Read in data from 2023
-Census_2023 <- read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))  %>%
-  filter(!is.na(quadrat))  %>%  # REMOVE after quadrat NAs are fixed
-  mutate(quadrat = sprintf("%04d",as.numeric(quadrat)))
-
-write.csv(quad_nas,"C:/Work/Smithsonian/Research/2023_Census_Analysis/data/Census2023_quadratsNAs.csv")
-#Rename the data
-Census_2008 <- scbi.stem1.corrected
-Census_2013 <- scbi.stem2.corrected
-Census_2018 <- scbi.stem3.corrected
-Species_table <- scbi.spptable
-
-#Combine the census data and species list
-Census_2008_Species <- merge(Census_2008, Species_table, by ="sp")
-Census_2013_Species <- merge(Census_2013, Species_table, by ="sp")
-Census_2018_Species <- merge(Census_2018, Species_table, by ="sp")
-Census_2023_Species <- merge(Census_2023, Species_table, by ="sp")
-
-#Define finished quadrats
-Finished_quadrats <- unique(Census_2023_Species$quadrat)
[…]

### 2024-02-20 code/figure3_woodyfluxes.R

@@ -1,0 +1,69 @@
+library(tidyverse) 
+
+woodyfluxes <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv")  %>% 
+  mutate(quadrat = sprintf("%04d", quadrat))  %>% 
+  select(-X) 
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d", quadrat))  %>% 
+  select(-X)
+
+group_trends <- woodyfluxes  %>% 
+  left_join(grouped_quadrats, by = c("quadrat" = "quadrat"))  %>% 
+  group_by(Group, Census)  %>% 
+  summarize(qc = n(),
+            AWP = sum(AWP) / qc, 
+            AWM = (sum(AWM) / qc ) * -1, 
+            AWR = sum(AWR) / qc, 
+            NetFlux = sum(NetFlux) / qc) %>% 
+  mutate(Group = as.character(Group))
+
+
+tot_trends <- woodyfluxes  %>% 
+  group_by(Census)  %>% 
+  summarize(qc = n(),
+            AWP = sum(AWP) / qc, 
+            AWM = (sum(AWM) / qc ) * -1, 
+            AWR = sum(AWR) / qc, 
+            NetFlux = sum(NetFlux) / qc)  %>% 
+  mutate(Group = "Whole Plot")
+
+fig3_woodyfluxes <- group_trends  %>% 
+  bind_rows(tot_trends)  %>% 
+  drop_na()   %>% 
+  pivot_longer(c(AWP, AWM, AWR, NetFlux), names_to = c("Flux"),values_to = "MgC_Yr_Ha")  %>% 
+  mutate(Flux = factor(Flux, levels = c("NetFlux","AWM","AWP","AWR")),
+         lwd_col = if_else(Group %in% c("Whole Plot"), .5, 1.2),
+         cex_col = if_else(Group %in% c("Whole Plot"), 5, 3),
+         cens_int = case_when(Census == 2 ~ "2008-\n2013",
+                              Census == 3 ~ "2013-\n2018",
+                              Census == 4 ~ "2018-\n2023")
+         )
+
+colz <- c("#017161","#4c90b0","darkblue","#754792")
+flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
+                            "AWP" = "Aboveground Woody Productivity",
[…]

### 2024-02-20 code/resutrrected-tree-testing.r

@@ -1,0 +1,23 @@
+#issues in quadrat 0515 (not exclusively but by far the largest), trees that were marked dead in 2013 are now marked as alive - need to decide how to gap fill backwards AND deal with outliers (copy from Camille?)
+
+scbi.stem4  <-  read.csv(url("https://raw.githubusercontent.com/SCBI-ForestGEO/2023census/main/processed_data/scbi.stem4.csv"))
+scbi.stem4.corrected  %>% 
+    filter(UID %in% UIDs) 
+scbi.stem3.corrected  %>% 
+    filter(UID %in% UIDs) 
+
+scbi.stem1  %>%
+    create_stem_UID()  %>% 
+    filter(UID %in% UIDs)
+scbi.stem2  %>%
+    create_stem_UID()  %>% 
+    filter(UID %in% UIDs)
+scbi.stem3  %>%
+    create_stem_UID()  %>% 
+    filter(UID %in% UIDs)
+scbi.stem4  %>%
+    create_stem_UID()  %>% 
+    filter(UID %in% UIDs)
+
+UIDs <- c("50569_1","50567_1","50568_1")
+

### 2024-02-21 code/Woodyflux_Calculations.r
Figures 3 & 4

@@ -1,13 +1,13 @@
 library(tidyverse)
 library(allodb)
 
-##### helper functions #####
+##### 0 - helper functions #####
 create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
   census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
   return(census_df)
 }
 
-##### Read in data from ALL censuses and the species table #####
+##### I - Read in data from ALL censuses and the species table #####
 grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
 load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
@@ -28,7 +28,7 @@ scbi.stem3.corrected <- scbi.stem3.corrected  %>%
 scbi.stem4.corrected <- scbi.stem4.corrected  %>% 
   mutate(Census = 4, hom = as.character(hom), dbh = as.character(dbh))
 
-##### Combine the census data & calculate woody fluxes (AWP, AWM, AWR) #####
+##### II - Combine the census data - run this before either section III or IV #####
 all_censuses <- scbi.stem1.corrected  %>% 
   bind_rows(scbi.stem2.corrected, scbi.stem3.corrected, scbi.stem4.corrected)  %>% 
   left_join(scbi.spptable, by = c("sp" = "sp"))  %>% 
@@ -39,6 +39,8 @@ all_censuses <- scbi.stem1.corrected  %>%
   arrange(UID,Census) %>% 
   mutate(Meas_Int = difftime(ExactDate,lag(ExactDate), units = "days") / 365.242)  
 
+
+##### III - Calculate woody fluxes by quadrat #####
 AWP <- all_censuses  %>% 
   filter(DFstatus %in% c("alive"))  %>% 
   mutate(WoodyGrowth = (ABG - lag(ABG)) / as.numeric(Meas_Int))  %>% 
@@ -77,10 +79,11 @@ woody_fluxes <- AWP  %>%
 write.csv(woody_fluxes, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv")
 
 
[…]

### 2024-02-21 code/figure4_mort_recr_by_species.r

@@ -1,0 +1,80 @@
+library(tidyverse) 
+library(patchwork)
+
+##### I - Plot mortality by species #####
+mortal <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/MortalityComposition.csv")
+
+
+mortality_by_sp <- mortal  %>% 
+  group_by(sp, Census)  %>% 
+  summarise(#qc = n(),
+            AWM = sum(AWM) / 25.6)
+mort_top_10 <- mortality_by_sp  %>% 
+  group_by(sp)  %>% 
+  summarize(AWM = sum(AWM))  %>% 
+  arrange(desc(AWM))  %>% 
+  slice(1:10)
+
+plt_mort_slice <- mortality_by_sp  %>% 
+  mutate(plot_sp = if_else(sp %in% mort_top_10$sp, sp, "other")) %>%
+  group_by(plot_sp, Census)  %>% 
+  summarize(AWM = sum(AWM))  %>%  
+  mutate(Cenus = factor(Census),
+         plot_sp = factor(plot_sp,levels = c(mort_top_10$sp, "other" )))
+
+    
+mortsp <- ggplot(plt_mort_slice, aes(x = plot_sp, y = AWM,fill = as.ordered(Census), group = Census, alpha = as.ordered(Census))) + 
+  geom_bar(position = "dodge",stat = "identity", col= "grey60") + 
+  theme_bw() + 
+  ylab(expression("Aboveground Woody Mortality"~(Mg~C~Ha^-1~Yr^-1))) +
+  xlab("Species") + 
+  scale_fill_manual(values = c(alpha("#94dce9",.1),alpha("#316bb8",.5),"#000050"), labels = c("2008 - 2013","2013 - 2018","2018 - 2023"), name = "") + 
+  theme(axis.text = element_text(size = 14),
+        axis.title = element_text(size = 16),
+        panel.grid.major.x = element_blank(),
+        legend.position = "none")
+
+##### II - Plot recruitment by species #####
+recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition.csv")
+
+
+recruitment_by_sp <- recruit  %>% 
+  group_by(sp, Census)  %>% 
+  summarise(AWR = sum(AWR) / 25.6)
+
+recr_top_10 <- recruitment_by_sp  %>% 
+  group_by(sp)  %>% 
+  summarize(AWR = sum(AWR))  %>% 
[…]

### 2024-02-23 doc/15yrsChange.Rmd
Updating Author List
Add Draft of Pests and Pathogens Paragraph

@@
 
+Christopher Pate
+
 Cameron Dow
@@
 Jen Jordan
+
+Lukas Magee
 
@@
 
-*Summary format for New Phytologist tends to be in a few long bullet points*
+*Summary format for New Phytologist must be under 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion*
 
@@
 **Keywords**:
+
+*New Phytologist allows 5-8 keywords. Short phrases are acceptable*
 
@@
 This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
+"Tree mortality is one of the largest uncertainities in projections of future forest dynamics" (Bugmann et al. 2019) - (*Add reference*)
 
-## (paragraph on tree mortality from non-endemic pests and pathogens)
+## (Canopy trees and nonindigenous pests and pathogens)
 
-Two Paragraphs
-
-1. The global implications of pests and pathogens
-2. History of pests and pathogens in eastern deciduous forests
-  1. Chestnut blight
-  2. Dutch elm disease
-  3. Hemlock Wooly Adelgid
-  4. Emerald Ash-borer
+Nuisance pests and pathogens have had a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al.
+2016, Fischer et al.
+2013, Herms and McCullough 2014, Fei et al.
+2014).
+The presence of nonindigenous forest insects and diseases, which have been historically introduced to new ecosystems through intercontinental trade, will continue to proliferate with increased globalization and human movement (Fischer et al.
+2013, Aukema et al.
+2010).
+Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy speciation and structure.
+Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al.
+2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts).
+These previously dominant canopy species are now confined to the understory as non-reproductive individuals (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).
+In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with risk for regional extirpation (Ellison et al.
+2018).
+The loss of mature canopy species to introduced nuisance species contributes to global change driven biomass losses, with estimates that nonindigenous pest and pathogens account for an additional 5.53 TgC per year (Fei et al.
+2013).
+A significant percentage of temperate forests remain at risk for future pest and pathogens (Fei et al.
+2013).
 

### 2024-02-26 code/Woodyflux_Calculations.r
annual mortality data, figures 4 & 5

@@ -1,3 +1,5 @@
+
+
 library(tidyverse)
 library(allodb)
 
@@ -6,6 +8,7 @@ create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
   census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
   return(census_df)
 }
+# Can skip to III and read-in all_censuses #
 
 ##### I - Read in data from ALL censuses and the species table #####
 grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
@@ -39,7 +42,9 @@ all_censuses <- scbi.stem1.corrected  %>%
   arrange(UID,Census) %>% 
   mutate(Meas_Int = difftime(ExactDate,lag(ExactDate), units = "days") / 365.242)  
 
+#save(all_censuses, file = "C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
 
+load(file = "C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
 ##### III - Calculate woody fluxes by quadrat #####
 AWP <- all_censuses  %>% 
   filter(DFstatus %in% c("alive"))  %>% 
@@ -79,7 +84,7 @@ woody_fluxes <- AWP  %>%
 write.csv(woody_fluxes, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv")
 
 
-##### IV - Decomposing woody mortality & recruitment by species #####
+##### IV - Decomposing census mortality & recruitment by species #####
 
 ### Mortality by species ###
 AWM_sp <- all_censuses  %>% 
@@ -92,8 +97,6 @@ AWM_sp <- all_censuses  %>%
   group_by(Census,sp)  %>%
   summarize(AWM = sum(WoodyMort,na.rm = T) / 1000 /.47)
 
-
-write.csv(AWM_sp, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/MortalityComposition.csv")
 
 ### Recruitment by species ###
 AWR_sp <- all_censuses  %>% 
@@ -108,3 +111,52 @@ AWR_sp <- all_censuses  %>%
 
 
[…]

### 2024-02-26 code/figure4_annual_mort_by_species.r

@@ -1,0 +1,80 @@
+library(tidyverse)
+library(patchwork)
+allmort_awm <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/MortalityComposition.csv")
+
+##### I - Identify top species (by mortality flux) & reformat data for plotting #####
+top_sp <- allmort_awm  %>% 
+  mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
+  ungroup()  %>% 
+  group_by(sp)  %>% 
+  mutate(abgmort = if_else(is.na(abgmort),0, abgmort))  %>% 
+  summarize(abgmort = sum(abgmort))  %>% 
+  arrange(desc(abgmort))  %>% 
+  slice_head(n = 7)
+  
+
+allspyr <- allmort_awm  %>% 
+  expand(sp,survey_year)
+mort_plot_df <- allmort_awm  %>% 
+  right_join(allspyr)  %>% 
+  mutate(abgmort = if_else(is.na(abgmort),0, abgmort))  %>% 
+  mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
+  mutate(plotspecies = if_else(sp %in% top_sp$sp,sp, "Other"))  %>% 
+  group_by(survey_year, plotspecies)  %>% 
+  summarize(mort_woody = sum(abgmort))  %>% 
+  mutate(plotspecies = factor(plotspecies, levels = c(top_sp$sp, "Other")))
+
+##### II - Plot mortality by species #####
+
+barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
+  facet_grid(~plotspecies) +
+  geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
+  theme_bw() +
+#  scale_fill_viridis_d()  +
+  scale_fill_grey() +
+  ylab(expression("Aboveground Woody Mortality"~(Mg~C~Ha^-1~Yr^-1))) +
+  theme(axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
+        axis.text.y = element_text(size = 14),
+        axis.title = element_text(size = 16),
+        axis.title.x = element_blank(),
+        panel.spacing.x = unit(0, "lines"),
+        #strip.background = element_blank(),
+        #strip.text = element_blank(),
+        panel.grid.major.x = element_blank(),
[…]

### 2024-02-26 code/figure4_mort_recr_by_species.r

@@ -1,80 +1,0 @@
-library(tidyverse) 
-library(patchwork)
-
-##### I - Plot mortality by species #####
-mortal <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/MortalityComposition.csv")
-
-
-mortality_by_sp <- mortal  %>% 
-  group_by(sp, Census)  %>% 
-  summarise(#qc = n(),
-            AWM = sum(AWM) / 25.6)
-mort_top_10 <- mortality_by_sp  %>% 
-  group_by(sp)  %>% 
-  summarize(AWM = sum(AWM))  %>% 
-  arrange(desc(AWM))  %>% 
-  slice(1:10)
-
-plt_mort_slice <- mortality_by_sp  %>% 
-  mutate(plot_sp = if_else(sp %in% mort_top_10$sp, sp, "other")) %>%
-  group_by(plot_sp, Census)  %>% 
-  summarize(AWM = sum(AWM))  %>%  
-  mutate(Cenus = factor(Census),
-         plot_sp = factor(plot_sp,levels = c(mort_top_10$sp, "other" )))
-
-    
-mortsp <- ggplot(plt_mort_slice, aes(x = plot_sp, y = AWM,fill = as.ordered(Census), group = Census, alpha = as.ordered(Census))) + 
-  geom_bar(position = "dodge",stat = "identity", col= "grey60") + 
-  theme_bw() + 
-  ylab(expression("Aboveground Woody Mortality"~(Mg~C~Ha^-1~Yr^-1))) +
-  xlab("Species") + 
-  scale_fill_manual(values = c(alpha("#94dce9",.1),alpha("#316bb8",.5),"#000050"), labels = c("2008 - 2013","2013 - 2018","2018 - 2023"), name = "") + 
-  theme(axis.text = element_text(size = 14),
-        axis.title = element_text(size = 16),
-        panel.grid.major.x = element_blank(),
-        legend.position = "none")
-
-##### II - Plot recruitment by species #####
-recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition.csv")
-
-
-recruitment_by_sp <- recruit  %>% 
-  group_by(sp, Census)  %>% 
-  summarise(AWR = sum(AWR) / 25.6)
-
-recr_top_10 <- recruitment_by_sp  %>% 
-  group_by(sp)  %>% 
-  summarize(AWR = sum(AWR))  %>% 
[…]

### 2024-02-26 code/figure5_recr_by_species.r

@@ -1,0 +1,61 @@
+library(tidyverse) 
+library(patchwork)
+
+##### I - Plot recruitment by species #####
+recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition.csv")
+spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
+
+missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
+
+canopyposition <- spTable  %>% 
+  select(spcode,canopy_position, life_form)  %>% 
+  mutate(canopy_position_new = if_else(life_form %in% c("tree"),"canopy","understory"))  %>%
+  select(sp = spcode, canopy_position = canopy_position_new)  %>% 
+  bind_rows(missing)
+
+recruitment_by_sp <- recruit  %>% 
+  left_join(canopyposition)  %>% 
+  group_by(sp, Census, canopy_position)  %>% 
+  summarise(AWR = sum(AWR) / 25.6)
+
+recr_top_10 <- recruitment_by_sp  %>% 
+  group_by(sp,canopy_position)  %>% 
+  summarize(AWR = sum(AWR))  %>% 
+  arrange(canopy_position,desc(AWR)) %>% 
+  group_by(canopy_position)  %>% 
+  slice(1:5)  %>% 
+  drop_na()
+
+plt_recr_slice <- recruitment_by_sp  %>% 
+  mutate(plot_sp = case_when(sp %in% recr_top_10$sp ~ sp,
+                             !(sp %in% recr_top_10$sp) & canopy_position == "canopy" ~ "Other Canopy sp.",
+                             !(sp %in% recr_top_10$sp) & canopy_position == "understory" ~ "Other Understory sp.")) %>%
+  group_by(plot_sp, Census, canopy_position)  %>% 
+  summarize(AWR = sum(AWR))  %>%  
+  drop_na()  %>% 
+  mutate(Cenus = factor(Census))#,
+#         plot_sp = factor(plot_sp,levels = c(recr_top_10$sp,  )))
+
+
+    
+#fig5 <-
+ ggplot(plt_recr_slice, aes(x = plot_sp, y = AWR,fill = as.ordered(Census), group = Census)) + 
+  facet_grid(~canopy_position, scales = "free_x") + 
[…]

### 2024-02-26 code/figure_6_temp.r

@@ -1,0 +1,18 @@
+library(tidyverse)
+library(allodb)
+
+##### I - Read in data from ALL censuses and the species table #####
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+
+all_censuses  %>% 
+    head()
+
+# sort species by canopy vs non-canopy status 
+
+# sort stems into canopy vs non-canopy (#if species is canopy, choose size cut-off?)
+
+# sum biomass by quadrat & census for canopy & non-canopy species
+
+# fit linear models for each quadrat on a) canopy biomass b) understory non-canopy biomass

### 2024-02-27 doc/15yrsChange.Rmd
annual mortality data, figures 4 & 5
Minor revisions to introduction

@@
 
+Rachel Hoffman
+
 Krystal Bagnaschi
-
-Rachel Hoffman
 
@@
 2013, Aukema et al.
-2010).
+2010, Brockerhoff et al.
+2006).
 Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy speciation and structure.
@@
 2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts).
-These previously dominant canopy species are now confined to the understory as non-reproductive individuals (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).
-In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with risk for regional extirpation (Ellison et al.
+These previously dominant overstorey species are now confined to the understory as non-reproductive seeds and saplings (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).
+In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with considerable risk of regional extirpation (Ellison et al.
 2018).
-The loss of mature canopy species to introduced nuisance species contributes to global change driven biomass losses, with estimates that nonindigenous pest and pathogens account for an additional 5.53 TgC per year (Fei et al.
-2013).
-A significant percentage of temperate forests remain at risk for future pest and pathogens (Fei et al.
-2013).
+The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens comparable to those attributed to fire and forest removal (Fei et al.
+2013)  Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al.
+2013)
 

### 2024-02-27 code/Woodyflux_Calculations.r
recruitment fixes, figure 5

@@ -68,12 +68,12 @@ AWR <- all_censuses  %>%
   group_by(UID)  %>% 
   arrange(UID, Census)  %>% 
   filter(case_when(first(Census) == 4 ~ Census == 4,
-                   first(Census) != 4 ~ (DFstatus %in% c("alive") & lag(DFstatus) %in% c("prior"))))  %>% 
+                   first(Census) != 4 ~ (DFstatus %in% c("alive") & lag(DFstatus) %in% c("prior"))))  %>%  
+  filter(as.numeric(dbh) <= 60)  %>%
   mutate(WoodyRecr = ABG / 5)  %>% 
   ungroup()  %>% 
   group_by(quadrat,Census)  %>% 
   summarize(AWR = sum(WoodyRecr, na.rm = T) / 1000 / .04 * .47)
-  
 
 woody_fluxes <- AWP  %>% 
   left_join(AWM, by = c("quadrat" = "quadrat", "Census" = "Census"))  %>% 
@@ -104,6 +104,7 @@ AWR_sp <- all_censuses  %>%
   arrange(UID,Census) %>% 
   filter(case_when(first(Census) == 4 ~ Census == 4,
                    first(Census) != 4 ~ (DFstatus %in% c("alive") & lag(DFstatus) %in% c("prior"))))  %>% 
+  filter(as.numeric(dbh) <= 60)  %>%
   mutate(WoodyRecr = ABG / 5)  %>% 
   ungroup()  %>% 
   group_by(Census,sp)  %>%

### 2024-02-27 code/figure5_recr_by_species.r

@@ -15,6 +15,7 @@ canopyposition <- spTable  %>%
 
 recruitment_by_sp <- recruit  %>% 
   left_join(canopyposition)  %>% 
+  mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
   group_by(sp, Census, canopy_position)  %>% 
   summarise(AWR = sum(AWR) / 25.6)
 
@@ -28,34 +29,47 @@ recr_top_10 <- recruitment_by_sp  %>%
 
 plt_recr_slice <- recruitment_by_sp  %>% 
   mutate(plot_sp = case_when(sp %in% recr_top_10$sp ~ sp,
-                             !(sp %in% recr_top_10$sp) & canopy_position == "canopy" ~ "Other Canopy sp.",
-                             !(sp %in% recr_top_10$sp) & canopy_position == "understory" ~ "Other Understory sp.")) %>%
+                             !(sp %in% recr_top_10$sp) & canopy_position == "canopy" ~ "Other canopy sp.",
+                             !(sp %in% recr_top_10$sp) & canopy_position == "understory" ~ "Other understory sp.")) %>%
   group_by(plot_sp, Census, canopy_position)  %>% 
   summarize(AWR = sum(AWR))  %>%  
   drop_na()  %>% 
-  mutate(Cenus = factor(Census))#,
-#         plot_sp = factor(plot_sp,levels = c(recr_top_10$sp,  )))
+  mutate(Census = factor(Census),
+         canopy_position = factor(canopy_position))
 
+f1 <- plt_recr_slice  %>% 
+  group_by(plot_sp,canopy_position)  %>% 
+  summarize(AWR = sum(AWR))  %>% 
+  filter(canopy_position %in% "canopy")  %>% 
+  arrange(desc(AWR))  %>% 
+  filter(plot_sp != "Other canopy sp.")  %>% 
+  pull(plot_sp)  %>% 
+  c(.,"Other canopy sp.")
 
-    
-#fig5 <-
- ggplot(plt_recr_slice, aes(x = plot_sp, y = AWR,fill = as.ordered(Census), group = Census)) + 
-  facet_grid(~canopy_position, scales = "free_x") + 
-  geom_bar(position = "dodge",stat = "identity", col= "black") + 
-  theme_bw() + 
-  ylab(expression("Aboveground Woody Recruitment"~(Mg~C~Ha^-1~Yr^-1))) +
-  xlab("Species") + 
[…]

### 2024-02-27 doc/15yrsChange.Rmd
outline results

@@
 
+## Paragraph discussing broad trend of biomass stock decrease (Figure 2) & net change in biomass & differences among regions in the plot 
+For the first time at the Front Royal ForestGEO site, C stocks have decreased since the last census (Figure 2).
+From 2018 to 2023, whole-plot C stocks have decreased from **"X to X"** Mg C/ha.
+There are different trends for the different regions of the plot.
+Both low nuisance regions show similar decreases in C stocks going from **"X to X"** and **"X to X"** Mg C/ha respectively.
+The magnitude of the decrease in the high der high nuisance area from 2018 to 2023 is greater, going from **"X to X"** Mg C/ha.
+C stock decrease **is related to/ is explained by** aboveground net woody production.
+In the first two census intervals, 2008 to 2013 and 2013 to 2018, there were positive C fluxes **VALUES?** and for the 2018 to 2023 interval there was a negative C flux of **VALUE Mg C/ha/year** (Figure 3a).
+
+## Paragraph discussing constituent fluxes of net biomass change - across the whole plot & within regions
+
+## Species compostion of mortality flux
+
+## Species compositon of recruitment flux
+
+## Regeneration & mortality mismatch
+
 # Discussion

### 2024-02-28 code/figure3_woodyfluxes.R
Update Figure 3

@@ -12,7 +12,7 @@ group_trends <- woodyfluxes  %>%
   group_by(Group, Census)  %>% 
   summarize(qc = n(),
             AWP = sum(AWP) / qc, 
-            AWM = (sum(AWM) / qc ) * -1, 
+            AWM = (sum(AWM) / qc ), 
             AWR = sum(AWR) / qc, 
             NetFlux = sum(NetFlux) / qc) %>% 
   mutate(Group = as.character(Group))
@@ -22,7 +22,7 @@ tot_trends <- woodyfluxes  %>%
   group_by(Census)  %>% 
   summarize(qc = n(),
             AWP = sum(AWP) / qc, 
-            AWM = (sum(AWM) / qc ) * -1, 
+            AWM = (sum(AWM) / qc ),  
             AWR = sum(AWR) / qc, 
             NetFlux = sum(NetFlux) / qc)  %>% 
   mutate(Group = "Whole Plot")
@@ -37,23 +37,31 @@ fig3_woodyfluxes <- group_trends  %>%
          cens_int = case_when(Census == 2 ~ "2008-\n2013",
                               Census == 3 ~ "2013-\n2018",
                               Census == 4 ~ "2018-\n2023")
-         )
+  )
 
-colz <- c("#017161","#4c90b0","darkblue","#754792")
+#colz <- c("#017161","#4c90b0","darkblue","#754792")
+colz <- c("#aa5274","#8578a7","#3A6F9E","#b8d89b")
+#colz <- c("#750000","#C7622B","#E7BC40", "#7e937f")
+#colz <- c("#b30000","#326634","#637eb6", "#000000")
+
+
+###linetype <- c("2", "3", "3", "1")
+
 flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
                             "AWP" = "Aboveground Woody Productivity",
                             "AWR" = "Aboveground Woody Recruitment",
-                            "NetFlux" = "Aboveground Net Woody Production "))
+                            "NetFlux" = "Net Biomass Change"))
 
 fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, group = Group, col = Group)) +
- facet_wrap(~Flux, scales = "free_y", labeller = flux_names) +
- geom_line(lwd = 1)+ 
[…]

### 2024-02-28 code/figure3_woodyfluxes.R
Update Figure 3 (Pt. 2)

@@ -39,13 +39,19 @@ fig3_woodyfluxes <- group_trends  %>%
                               Census == 4 ~ "2018-\n2023")
   )
 
-#colz <- c("#017161","#4c90b0","darkblue","#754792")
-colz <- c("#aa5274","#8578a7","#3A6F9E","#b8d89b")
-#colz <- c("#750000","#C7622B","#E7BC40", "#7e937f")
-#colz <- c("#b30000","#326634","#637eb6", "#000000")
+#install.packages("dichromat")
+#library(dichromat)
+
+#colz <- c("#017161","#4c90b0","darkblue","#754792")     #Colorblind friendly
+colz <- c("#aa5274","#8578a7","#3A6F9E","#b8d89b")       #Not very color blind friendly (though still readable)
+#colz <- c("#750000","#C7622B","#E7BC40", "#7e937f")     #Colorblind friendly 
+#colz <- c("#b30000","#326634","#637eb6", "#000000")     #Not very colorblind friendly (though still readable)
+
+#dichromat(colz, type = c("deutan", "protan", "tritan")) # - checks if a color palette is colorblind friendly
 
 
 ###linetype <- c("2", "3", "3", "1")
+
 
 flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
                             "AWP" = "Aboveground Woody Productivity",

### 2024-02-29 doc/15yrsChange.Rmd
Update 15yrsChange.Rmd
starting the second paragraph in the results (definitely still incomplete)

@@
 From 2018 to 2023, whole-plot C stocks have decreased from **"X to X"** Mg C/ha.
-There are different trends for the different regions of the plot.
 Both low nuisance regions show similar decreases in C stocks going from **"X to X"** and **"X to X"** Mg C/ha respectively.
-The magnitude of the decrease in the high der high nuisance area from 2018 to 2023 is greater, going from **"X to X"** Mg C/ha.
-C stock decrease **is related to/ is explained by** aboveground net woody production.
+The magnitude of the C stock decrease in the high deer high nuisance area from 2018 to 2023 is larger, going from **"X to X"** Mg C/ha.
+C stock decrease **is related to/ is explained by** above ground net woody production.
 In the first two census intervals, 2008 to 2013 and 2013 to 2018, there were positive C fluxes **VALUES?** and for the 2018 to 2023 interval there was a negative C flux of **VALUE Mg C/ha/year** (Figure 3a).
@@
 ## Paragraph discussing constituent fluxes of net biomass change - across the whole plot & within regions
+The trend in net woody production can be explained by its constituent fluxes (Figure 3).
+Trends in woody mortality show that for the first two census intervals, the whole plot has lost **VALUE** and then **VALUE** Mg C/ha/year, increasing to a loss of **VALUE** Mg C/ha/year for the most recent census interval (Figure 3b).
+Woody productivity contributes positively to the flux, showing that the whole plot gained **VALUE** Mg C/ha/year in the first interval, **VALUE** Mg C/ha/year in the second, and **VALUE** Mg C/ha/year in the third (Figure 3c).
+Areas of low deer, low nuisance species were consistently the most productive ending at **VALUE** Mg C/ha/year in the interval 2018 to 2023, and areas of high deer, high nuisance species were the least productive at **VALUE** Mg C/ha/year in the interval 2018 to 2023.
 

### 2024-03-02 code/figure3_woodyfluxes.R
Update Figure 3

@@ -1,9 +1,9 @@
 library(tidyverse) 
 
-woodyfluxes <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv")  %>% 
+woodyfluxes <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv") %>%
   mutate(quadrat = sprintf("%04d", quadrat))  %>% 
   select(-X) 
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv") %>%
   mutate(quadrat = sprintf("%04d", quadrat))  %>% 
   select(-X)
 
@@ -32,42 +32,32 @@ fig3_woodyfluxes <- group_trends  %>%
   drop_na()   %>% 
   pivot_longer(c(AWP, AWM, AWR, NetFlux), names_to = c("Flux"),values_to = "MgC_Yr_Ha")  %>% 
   mutate(Flux = factor(Flux, levels = c("NetFlux","AWM","AWP","AWR")),
-         lwd_col = if_else(Group %in% c("Whole Plot"), .5, 1.2),
+         lwd_col = if_else(Group %in% c("Whole Plot"), 5, 3),  #0.5, 1.2
          cex_col = if_else(Group %in% c("Whole Plot"), 5, 3),
          cens_int = case_when(Census == 2 ~ "2008-\n2013",
                               Census == 3 ~ "2013-\n2018",
                               Census == 4 ~ "2018-\n2023")
   )
 
-#install.packages("dichromat")
-#library(dichromat)
-
-#colz <- c("#017161","#4c90b0","darkblue","#754792")     #Colorblind friendly
-#colz <- c("#aa5274","#8578a7","#3A6F9E","#b8d89b")       #Not very color blind friendly (though still readable)
-colz <- c("#750000","#C7622B","#E7BC40", "#7e937f")     #Colorblind friendly 
-#colz <- c("#b30000","#326634","#637eb6", "#000000")     #Not very colorblind friendly (though still readable)
-
-#dichromat(colz, type = c("deutan", "protan", "tritan")) # - checks if a color palette is colorblind friendly
 
 
-###linetype <- c("2", "3", "3", "1")
-
-
[…]

### 2024-03-04 code/figure3_woodyfluxes.R
fig 3 tweaks

@@ -1,9 +1,9 @@
 library(tidyverse) 
 
-woodyfluxes <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/WoodyFluxes.csv") %>%
+woodyfluxes <- read.csv("data/processed_data/WoodyFluxes.csv") %>%
   mutate(quadrat = sprintf("%04d", quadrat))  %>% 
   select(-X) 
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv") %>%
+grouped_quadrats <- read.csv("data/grouped_quadrats.csv") %>%
   mutate(quadrat = sprintf("%04d", quadrat))  %>% 
   select(-X)
 
@@ -47,7 +47,7 @@ flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
                             "AWR" = "Aboveground Woody Recruitment",
                             "NetFlux" = "Net Biomass Change"))
 
-ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, group = Group, col = Group)) +
+fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, group = Group, col = Group)) +
   facet_wrap(~Flux, scales = "free_y", labeller = flux_names) +
   geom_line(data = fig3_woodyfluxes  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
@@ -68,61 +68,7 @@ ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, group = Gr
         axis.text.y = element_text(size = 18, margin = margin(t = 0,r = .08, b = 0,l = 0, unit = "in")),
         axis.title = element_text(size = 20)) 
 
-ggsave(fig3,filename = "C:/Work/Smithsonian/Repos/15yrsChange/doc/display/Figure3.jpeg", units = "in",
+ggsave(fig3,filename = "doc/display/Figure3.jpeg", units = "in",
        height = 8, width = 10, dpi = 300)
 
-
-
-##Step function (version 1 - with deer intensity)
-
[…]

### 2024-03-04 code/figure3_woodyfluxes.R
Fig 3 Minor Changes

@@ -53,8 +53,8 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
   geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#750000") +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 25, cex = 3, fill = "#C7622B") +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 24, cex = 3, fill = "#E7BC40") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 22, cex = 3, fill = "#C7622B") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#E7BC40") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
   scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), values = colz) +

### 2024-03-04 doc/15yrsChange.Rmd
Update 15yrsChange.Rmd
worked on second paragraph results section

@@
 Areas of low deer, low nuisance species were consistently the most productive ending at **VALUE** Mg C/ha/year in the interval 2018 to 2023, and areas of high deer, high nuisance species were the least productive at **VALUE** Mg C/ha/year in the interval 2018 to 2023.
+Recruitment has consistently added smaller magnitude amounts of carbon with the whole plot gaining **VALUE** Mg C/ha/year (Figure 3d).
+Recruitment was highest at **VALUE** Mg C/ha/year in areas with high deer and high vulnerable species, and lowest at **VALUE** Mg C/ha/year in areas with high deer and low vulnerable species.
 

### 2024-03-05 data/20m_grid_groups.R
Update Figure 1

@@ -1,0 +1,12 @@
+install.packages("sf")
+library(sf)
+library(tidyverse)
+
+grouped_quadrats <- read.csv("/Users/rachelhoffman/Documents/GitHub/15yrsChange/data/grouped_quadrats.csv")
+
+dataframe <- read_sf("/Users/rachelhoffman/Documents/GitHub/15yrsChange/data/20m_grid/20m_grid.shp") %>% 
+  rename(quadrat = PLOT)
+
+merged_shapefile <- merge(x = dataframe2, y = grouped_quadrats, by = "quadrat")
+
+st_write(merged_shapefile, "/Users/rachelhoffman/Documents/GitHub/15yrsChange/data/qgis_data/20m_grid_groups.shp")

### 2024-03-05 doc/15yrsChange.Rmd
Update 15yrsChange.Rmd

@@
 ## Paragraph discussing broad trend of biomass stock decrease (Figure 2) & net change in biomass & differences among regions in the plot 
-For the first time at the Front Royal ForestGEO site, C stocks have decreased since the last census (Figure 2).
+For the first time at the Front Royal ForestGEO site, whole-plot C stocks have decreased since the last census (Figure 2).
 From 2018 to 2023, whole-plot C stocks have decreased from **"X to X"** Mg C/ha.
-Both low nuisance regions show similar decreases in C stocks going from **"X to X"** and **"X to X"** Mg C/ha respectively.
-The magnitude of the C stock decrease in the high deer high nuisance area from 2018 to 2023 is larger, going from **"X to X"** Mg C/ha.
-C stock decrease **is related to/ is explained by** above ground net woody production.
-In the first two census intervals, 2008 to 2013 and 2013 to 2018, there were positive C fluxes **VALUES?** and for the 2018 to 2023 interval there was a negative C flux of **VALUE Mg C/ha/year** (Figure 3a).
+Both low-nuisance regions show similar decreases in C stocks going from **"X to X"** and **"X to X"** Mg C/ha, respectively.
+In the high deer high nuisance regions C stocks have decreased for the past two censuses, with a higher magnitude decrease of **"X to X"** Mg C/ha from 2018 to 2023.
+C stock decrease **is related to/ is explained by** above ground net biomass change.
+In the first two census intervals, 2008 to 2013 and 2013 to 2018, there were positive C fluxes, but for the 2018 to 2023 interval there was a negative C flux of **VALUE** Mg C/ha/year (Figure 3a).
 
@@
 ## Species compostion of mortality flux
+
+Woody mortality was the largest driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality.
+Fraxinus americana had the highest abovegroung woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4).
+Quercus velutina contributed the second-highest levels of mortality, losing **VALUE** Mg C/ha/year in 2022.
+Quercus rubra, Carya species, Quercus prinus, Quercus alba, and Liriodendron tulipifera were all significant contributors as well.
 

### 2024-03-05 code/figure6_canopy_understory_biomass_changes.r
figure 6

@@ -1,0 +1,145 @@
+library(tidyverse)
+library(sf)
+library(allodb)
+
+##### I - Read in data from ALL censuses and the species table #####
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
+quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
+spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
+
+missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
+
+canopyposition <- spTable  %>% 
+  select(spcode,canopy_position, life_form)  %>% 
+  mutate(canopy_position_new = if_else(life_form %in% c("tree"),"canopy","understory"))  %>%
+  select(sp = spcode, canopy_position = canopy_position_new)  %>% 
+  bind_rows(missing)
+
+# sort species by canopy vs non-canopy status 
+all_censuses  %>% 
+  left_join(canopyposition)  %>% 
+  View()
+
+dbh_cutoff <- 50
+
+canopy_abg <- all_censuses  %>% 
+  left_join(canopyposition)  %>% 
+  filter(DFstatus %in% c("alive"))  %>%
+  filter(canopy_position %in% "canopy" & dbh >= dbh_cutoff)  %>%  
+  mutate(quadrat = factor(quadrat), Census = factor(Census))  %>% 
+  group_by(quadrat, Census, .drop = FALSE)  %>% 
+  summarize(WoodyAGB = sum(ABG,na.rm = T) / 1000 / .04 * .47)
+
+understory_abg <- all_censuses  %>% 
+  left_join(canopyposition)  %>% 
+  filter(DFstatus %in% c("alive"))  %>%
+  filter(canopy_position %in% "canopy" & dbh < dbh_cutoff)  %>%  
+  mutate(quadrat = factor(quadrat), Census = factor(Census))  %>%
[…]

### 2024-03-05 doc/15yrsChange.Rmd
Update 15yrsChange.Rmd
worked on results paragraphs for first 5 figures

@@
 Woody mortality was the largest driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality.
-Fraxinus americana had the highest abovegroung woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4).
+Fraxinus americana had the highest aboveground woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4).
 Quercus velutina contributed the second-highest levels of mortality, losing **VALUE** Mg C/ha/year in 2022.
@@
 ## Species compositon of recruitment flux
+Certain species dominated aboveground woody recruitment, with understory species being recruited at higher rates than canopy species (Figure 5).
+Out of the canopy species, Carya species showed the highest levels of recruitment with a rate of **VALUE** Mg C/ha/year recruited from 2018 to 2023 (Figure 5a).
+Acer negundo had the second highest rate of recruitment for canopy species at **VALUE** Mg C/ha/year from 2018 to 2023.
+Fagus grandifolia, Fraxinus americana, and Ulmus rubra are the other species with the highest recruitment rates.
+Most canopy species had lower recruitment rates in the 2018 to 2023 census interval than the 2008 to 2013 interval.
+Lindera benzoin had by far the highest recruitment rate of the understory species at **VALUE** Mg C/ha/year from 2018 to 2023 (Figure 3b).
+Asimina triloba had the second highest rate of recruitment of the understory species at **VALUE** Mg C/ha/year recruited from 2018 to 2023.
+Other understory species with high recruitment were Carpinus caroliniana, Hamamelis virginiana, and Celtis occidentalis.
 

### 2024-03-06 code/figure5_recr_by_species.r
figure updates - effect of deer exclosure

@@ -2,74 +2,102 @@ library(tidyverse)
 library(patchwork)
 
 ##### I - Plot recruitment by species #####
-recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition.csv")
+recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition_annual.csv")
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+
 
 missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
 
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+
 canopyposition <- spTable  %>% 
   select(spcode,canopy_position, life_form)  %>% 
-  mutate(canopy_position_new = if_else(life_form %in% c("tree"),"canopy","understory"))  %>%
+  mutate(canopy_position_new = case_when(spcode %in% "acne" ~ "understory",
+                                         life_form %in% c("tree") ~ canopy,
+                                        .default = "understory"))  %>%  
   select(sp = spcode, canopy_position = canopy_position_new)  %>% 
   bind_rows(missing)
 
 recruitment_by_sp <- recruit  %>% 
   left_join(canopyposition)  %>% 
   mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
-  group_by(sp, Census, canopy_position)  %>% 
-  summarise(AWR = sum(AWR) / 25.6)
+  group_by(quadrat,sp, canopy_position)  %>% 
+  summarise(AWR_yr_ha = sum(AWR_yr) / .04, n_stems_yr_ha = sum(n_stems_yr) / .04)
 
 recr_top_10 <- recruitment_by_sp  %>% 
   group_by(sp,canopy_position)  %>% 
-  summarize(AWR = sum(AWR))  %>% 
+  summarize(AWR = sum(AWR_yr_ha))  %>% 
[…]

### 2024-03-06 code/figure6_canopy_understory_biomass_changes.r

@@ -8,6 +8,9 @@ grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped
 load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
 deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
 quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
+grouped_quadrats <- read.csv("data/grouped_quadrats.csv") %>%
+  mutate(quadrat = sprintf("%04d", quadrat))  %>% 
+  select(-X)
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
 
 missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
@@ -68,8 +71,8 @@ understory_slopes <-  understory_trends %>%
   
 ##### Plotting #####
 fillcols <- c("#b7c3e0", "#f3f2f2", "#ffc080","#85a4eb", "#e2dede", "#fd9e3f","#004494", "#a6a6a6", "#cc4c00")
-bivariate_color_scale <- tibble(canopy = rep(3:1, 3),
-                                understory = rep(c(1,2,3), each = 3),
+bivariate_color_scale <- tibble(understory = rep(3:1, 3),
+                                canopy = rep(1:3, each = 3),
                                 fill = fillcols)
 
 # bivariate_color_scale <- tibble(
@@ -108,20 +111,21 @@ map <- ggplot(data = plotdf) +
 #    geom_sf(data = deer_exclosure, colour = "black", fill = NA,lwd = 1.5) +
     scale_fill_identity() +
     # geom_sf(data = deer_exclosure, colour = "black", fill = NA,lwd = 1.5) +
-    theme_classic()
+    theme_classic() +
+    theme(plot.margin = unit(c(0,0,0,0), "cm"))
 
 
 legend <- ggplot() +
   geom_tile(
     data = bivariate_color_scale,
     mapping = aes(
-      x = factor(canopy),
-      y = factor(understory),
+      y = factor(canopy),
[…]

### 2024-03-06 code/figure2_agb.R
Update Author List and Tweak Figure 4

@@ -3,7 +3,7 @@
 
 #install.packages('allodb')
 library(allodb)
-library(dyplr)
+library(tidyverse)
 
 #Read in data from all censuses and the species table
 grouped_quadrats <- read.csv("data/grouped_quadrats.csv") %>% 
@@ -79,8 +79,13 @@ colz <- c("#750000","#C7622B", "#E7BC40","#7e937f")
 
 fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Group)) +
 #fig2 <- ggplot(figure2_agb, aes(y=Abg_Mg_Ha, x=Year, group = Group, col = Group)) +
-  geom_line(lwd = c(rep(1,12),rep(3,4))) +
-  geom_point(cex = c(rep(3,12),rep(5,4))) +
+  geom_line(data = figure2_agb  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
+  geom_line(data = figure2_agb %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
+  geom_line(data = figure2_agb %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#750000") +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("2")), pch = 22, cex = 3, fill = "#C7622B") +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#E7BC40") +
+  geom_point (data = figure2_agb %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
     minor_breaks = c())+
   scale_color_manual(name = element_blank(),labels = c("Low deer, low nuisance","High deer, low nuisance", "High deer, high nuisance", "Plot"),values = colz) + 

### 2024-03-06 code/figure4_annual_mort_by_species.r

@@ -1,6 +1,7 @@
+#install.packages("patchwork")
 library(tidyverse)
 library(patchwork)
-allmort_awm <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/MortalityComposition.csv")
+allmort_awm <- read.csv("data/processed_data/MortalityComposition.csv")
 
 ##### I - Identify top species (by mortality flux) & reformat data for plotting #####
 top_sp <- allmort_awm  %>% 
@@ -24,13 +25,21 @@ mort_plot_df <- allmort_awm  %>%
   summarize(mort_woody = sum(abgmort))  %>% 
   mutate(plotspecies = factor(plotspecies, levels = c(top_sp$sp, "Other")))
 
+mort_plot_df$plotspecies <- gsub('fram', 'Fraxinus americana', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('Hickory spp.', 'Carya spp.', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('litu', 'Liriodendron tulipifera', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('qual', 'Quercus alba', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('qupr', 'Quercus prinus', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('quru', 'Quercus rubra', mort_plot_df$plotspecies)
+mort_plot_df$plotspecies <- gsub('quve', 'Quercus velutina', mort_plot_df$plotspecies)
+
 ##### II - Plot mortality by species #####
 
 barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
-  facet_grid(~plotspecies) +
+  facet_grid(~plotspecies, labeller = label_parsed()) +
   geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
   theme_bw() +
-#  scale_fill_viridis_d()  +
+  #scale_fill_viridis_d()  +
   scale_fill_grey() +
   ylab(expression("Aboveground Woody Mortality"~(Mg~C~Ha^-1~Yr^-1))) +
   theme(axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
@@ -39,13 +48,17 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
         axis.title.x = element_blank(),
[…]

### 2024-03-08 code/figure2_agb.R
Figure Updates

@@ -75,16 +75,16 @@ tot_trends <- quadrat_abg_2008  %>%
 figure2_agb <- group_trends  %>% 
   bind_rows(tot_trends)
 
-colz <- c("#750000","#C7622B", "#E7BC40","#7e937f")
+colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 
 fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Group)) +
 #fig2 <- ggplot(figure2_agb, aes(y=Abg_Mg_Ha, x=Year, group = Group, col = Group)) +
   geom_line(data = figure2_agb  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
   geom_line(data = figure2_agb %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
   geom_line(data = figure2_agb %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
-  geom_point(data = figure2_agb %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#750000") +
-  geom_point(data = figure2_agb %>% filter(Group %in% c("2")), pch = 22, cex = 3, fill = "#C7622B") +
-  geom_point(data = figure2_agb %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#E7BC40") +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#E7BC40") +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("2")), pch = 25, cex = 3, fill = "#C7622B") +
+  geom_point(data = figure2_agb %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = figure2_agb %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
     minor_breaks = c())+

### 2024-03-08 code/figure3_woodyfluxes.R

@@ -41,7 +41,7 @@ fig3_woodyfluxes <- group_trends  %>%
 
 
 
-colz <- c("#750000","#C7622B","#E7BC40", "#7e937f")
+colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
                             "AWP" = "Aboveground Woody Productivity",
                             "AWR" = "Aboveground Woody Recruitment",
@@ -52,9 +52,9 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_line(data = fig3_woodyfluxes  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#750000") +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 22, cex = 3, fill = "#C7622B") +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#E7BC40") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#E7BC40") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 25, cex = 3, fill = "#C7622B") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
[…]

### 2024-03-08 code/figure4_annual_mort_by_species.r

@@ -36,7 +36,7 @@ mort_plot_df$plotspecies <- gsub('quve', 'Quercus velutina', mort_plot_df$plotsp
 ##### II - Plot mortality by species #####
 
 barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
-  facet_grid(~plotspecies, labeller = label_parsed()) +
+  facet_grid(~plotspecies) +
   geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
   theme_bw() +
   #scale_fill_viridis_d()  +

### 2024-03-08 code/figure5_recr_by_species.r

@@ -2,15 +2,15 @@ library(tidyverse)
 library(patchwork)
 
 ##### I - Plot recruitment by species #####
-recruit <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition_annual.csv")
+recruit <- read.csv("data/processed_data/RecruitmentComposition_annual.csv")
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+grouped_quadrats <- read.csv("data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
 
 
 missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
 
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+load("data/census_data/all_censuses_agb.rdata")
 
 canopyposition <- spTable  %>% 
   select(spcode,canopy_position, life_form)  %>% 
@@ -89,7 +89,7 @@ fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,f
       ylab(expression(atop("# of Recruits", (Ha^-1~Yr^-1)))) +
       xlab("Species") + 
       guides(alpha = "none", fill = guide_legend(override.aes = list(size = 8))) +
-      scale_fill_manual(values = c("#750000","#C7622B","#E7BC40","#7e937f"), labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), name = "") + 
+      scale_fill_manual(values = c("#E7BC40","#C7622B","#750000","#7e937f"), labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), name = "") + 
       theme(axis.text = element_text(size = 14, angle = 45, hjust = 1),
             axis.title = element_text(size = 16),
             panel.grid.major.x = element_blank(),
[…]

### 2024-03-09 code/figure3_woodyfluxes.R
Figure Updates (Pt. 2)

@@ -1,4 +1,5 @@
 library(tidyverse) 
+library(ggplot2)
 
 woodyfluxes <- read.csv("data/processed_data/WoodyFluxes.csv") %>%
   mutate(quadrat = sprintf("%04d", quadrat))  %>% 
@@ -57,7 +58,7 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
-  scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), values = colz) +
+  scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), values = colz, guide = guide_legend(nrow = 2)) +
   labs(y = expression("Carbon Flux"~(Mg~C~Ha^-1~Yr^-1)), x = "Census") +
   theme_bw() +
   theme(legend.position = "top",

### 2024-03-09 code/figure4_annual_mort_by_species.r

@@ -33,10 +33,16 @@ mort_plot_df$plotspecies <- gsub('qupr', 'Quercus prinus', mort_plot_df$plotspec
 mort_plot_df$plotspecies <- gsub('quru', 'Quercus rubra', mort_plot_df$plotspecies)
 mort_plot_df$plotspecies <- gsub('quve', 'Quercus velutina', mort_plot_df$plotspecies)
 
+facet_order <- c("Fraxinus americana", "Quercus velutina", "Quercus rubra", "Carya spp.", 
+                 "Quercus prinus", "Quercus alba", "Liriodendron tulipifera", "Other")
+
+facet_years <- c(2013, 2018, 2023)
+
 ##### II - Plot mortality by species #####
 
 barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
-  facet_grid(~plotspecies) +
+  facet_grid(~ factor(plotspecies, levels = facet_order)) +
+  scale_x_continuous(breaks = facet_years) +
   geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
   theme_bw() +
   #scale_fill_viridis_d()  +
@@ -48,12 +54,12 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
         axis.title.x = element_blank(),
         panel.spacing.x = unit(0, "lines"),
         #strip.background = element_blank(),
-        strip.text = element_text(size = 8, face = "italic"),
-
+        strip.text = element_text(size = 8, 
+                                  face = "italic"), 
         panel.grid.major.x = element_blank(),
         panel.grid.minor.x = element_blank(),
-        legend.position = "none")
-
+        legend.position = "none") 
+  
 
 print(barp)
 

### 2024-03-11 doc/15yrsChange.Rmd
Update Author List and Tweak Figure 4
Updated author list
Updated author list (Pt. 2)
cleaning R markdown for export to google doc & figure 5 tweaks

@@
 
-**Authors (tentative / incomplete list, not in order):**
+**Authors :**
 
@@
 
-Christopher Pate
+Cameron Dow
 
-Cameron Dow
+Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
 
@@
 
-Valentine Herrmann^1^ ( Orcid ID : 0000-0002-4519-481X)
+David Mitre
 
-David Mitre
+Christopher Pate
 
@@
 
-*corresponding author: MorrealeLL@si.edu; +1 607 227 1350
+*corresponding author: MorrealeLL@si.edu;
 
@@
 map_and_photos = 1
-photo_deer_exclosure = "1a"
-photo_represenative_upland = "1b"
-photo_graveyard = "1c"
 
@@
 
-mortality_by_genera = 4 # 4-panel figure showing biomass mortality by genera for for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard
+mortality_by_genera = 4 # 2-panel figure showing annual biomass mortality by genera for (a) canopy species and (b) understory species
 
-regen_failure = 5 #some figure illustrating absence of small canopy trees/ shifting  size distributions
+recruitment_by_genera = 5 # 2-panel figure showing # of recruited stems by genera for (a) canopy species and (b) understory species by plot regions
+
+regen_failure = 6 #some figure illustrating absence of small canopy trees/ shifting  size distributions
 
@@
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
-This is problematic in that nuisance species -- including non-endemic insect pests and pathogens, non-endemic plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
-"Tree mortality is one of the largest uncertainities in projections of future forest dynamics" (Bugmann et al. 2019) - (*Add reference*)
+This is problematic in that nuisance species -- including nonindigenous insect pests and pathogens, nonindigenous plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
+"Tree mortality is one of the largest uncertainities in projections of future forest dynamics" (Bugmann et al. 2019)
 
-## (Canopy trees and nonindigenous pests and pathogens)
-
-Nuisance pests and pathogens have had a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al.
+Nuisance pests and pathogens can have a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al.
 2016, Fischer et al.
@@
 2006).
-Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy speciation and structure.
+Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy composition and structure.
 Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al.
@@
 2018).
-The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens comparable to those attributed to fire and forest removal (Fei et al.
+The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens likely comparable to those attributed to fire and forest removal (Fei et al.
 2013)  Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al.
@@
 Non-endemic pests and pathogens facilitate invasion of non-endemic plants [@guo_impacts_2023 and refs therein].
-
-## (paragraph on recruitment failure because of deer and non-endemic plants)
-
-Recruitment Failure Paragraph
 
@@
 Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
-(1) C stocks have declined (-∆AGB)
-(2) canopy tree mortality and associated biomass loss are increasing, in large part due to non-endemic nuisance species (pests & pathogens)
-(3) recruitment (outside deer exclosure) and growth have not kept pace with tree mortality, implying that future mortality will result in substantial net biomass loss (regeneration debt)
+(1) ) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB)
+(2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to nonindigenous nuisance species
+(3) Recruitment and woody growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
 
@@
 ## Site Description
-
-*Need to add the Anderson-Teixeira et al. 2015 reference to references page*
 
@@
 Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).
-The plot, which includes a 4-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
-It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the World.
+The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
+It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the world.
 
-*Add information about the history of pests and pathogens at this site and at similar forests*
+* Add information about the history of pests and pathogens at this site and at similar forests
 
 In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
-(1) low deer, low non-endemic insects & pathogens (Fig.
-`r photo_deer_exclosure`)- 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
-(2) high deer, low non-endemic insects & pathogens (Fig.
-`r photo_represenative_upland`) - # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
-(3) high deer, high non-endemic insects & pathogens (Fig.
-`r photo_graveyard`) - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens.
+(1) low deer, low non-endemic insects & pathogens - 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
+(2) high deer, low non-endemic insects & pathogens # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens; 
+(3) high deer, high non-endemic insects & pathogens - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens.
 These areas were delineated as follows.
@@
 Photos within the SCBI ForestGEO plot: (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens ** *(add a map of the plot)* All photos taken September 2023 by K.
-Anderson-Teixeira.](display/photos.png)
+Anderson-Teixeira.](display/Figure1.png)
 
@@
 
-*(Describe main census and mortality survey)*
-
-A number of surveys are regularly conducted at the site, two of which are included in our analysis.
+A number of field surveys are regularly conducted at the site, two of which are included in our analysis.
 As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).
-This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; hereafter referred to as dbh).
+This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; dbh).
 The census records information regarding the dbh, species, status and location of each stem included in the survey.
-Each plant is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field.
-For multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.
-The location of each individual within its respective 20 x 20 meter quadrat is recorded on a map of the plot on ArcGIS FieldMaps.
-Established in 2008, there have been 4 censuses at the site, comprising 20 years of detailed forest dynamics data.
-In addition to the ForestGEO census, an annual mortality survey is conducted on the site.
+Each invidual is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field.
+In multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.
+Location is recorded on a digitized map of the plot on the ESRI ArcGIS FieldMaps application (*citation*).
+Established in 2008, there have been four censuses at the site, comprising 15 years of detailed forest dynamics data.
+In addition to the ForestGEO census, since 2014 an annual mortality survey has been conducted on the site.
+Initiated, in part, as a response to the presence of Emerald Ashborer, the mortality survey is conducted on all trees that had a dbh greater than 10 centimeters in the previous full census.
 Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.
 Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
-
-*Add Condit 1998 in references*
-
-*In response to the presence of Emerald Ashborer, a mortality survey was initiated, then improved in year [X].
-
-*(describe seedlings survey, if used)*
-
-*(describe QA/QC and technical workflow)*
 
@@
 
-apply the metric of @miller_overabundant_2023 to determine imminent failure/ probable failure/ insecure/ secure?
+* We  
+*
 
@@
 
-## Paragraph discussing broad trend of biomass stock decrease (Figure 2) & net change in biomass & differences among regions in the plot 
-For the first time at the Front Royal ForestGEO site, whole-plot C stocks have decreased since the last census (Figure 2).
+For the first time since the establishment of  the SCBI ForestGEO site, whole-plot aboveground C stocks decreased (Figure 2).
 From 2018 to 2023, whole-plot C stocks have decreased from **"X to X"** Mg C/ha.
@@
 
-## Paragraph discussing constituent fluxes of net biomass change - across the whole plot & within regions
+![**Figure `r biomass_trends`.
+Aboveground biomass decreased at the site for the first time since establishment.
+The net decrease at the plot-level was driven by large losses in regions with high vulnerability to nuisance species.**](display/Figure2.jpeg)
+
 The trend in net woody production can be explained by its constituent fluxes (Figure 3).
@@
 
-## Species compostion of mortality flux
+![**Figure `r loss_gain_net`.
+Fluxes of aboveground woody biomass.
+Large increases in woody mortality combined with sustained decreases in woody productivity result in decreasing net biomass in the most recent census.**](display/Figure3.jpeg)
 
-Woody mortality was the largest driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality.
+## Species composition of mortality flux
+
+Woody mortality was the primary driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality.
 Fraxinus americana had the highest aboveground woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4).
@@
 
-## Species compositon of recruitment flux
+![**Figure `r mortality_by_genera`. Annual woody mortality flux by species/genera. We report large increases in mortality of white ash, red oak, and black oak.**](display/Figure4.jpeg)
+
+## Species composition of recruitment flux
 Certain species dominated aboveground woody recruitment, with understory species being recruited at higher rates than canopy species (Figure 5).
@@
 
+![**Figure `r recruitment_by_genera`.
+Annual recruitment of canopy & understory species/genera across all censuses.
+Recruitment of canopy species is significantly higher within the deer exclosure.
+Recruitment of understory species is 20x that of canopy species.**](display/Figure5.jpeg)
+
 ## Regeneration & mortality mismatch
+
+![**Figure `r regen_failure`.**](display/Figure6.jpeg)
 
@@
 
-# Conclusions (optional) *(No Conclusions section listed in Submission Guidelines)*
+# Conclusions /Future Directions
 

### 2024-03-11 code/exploratory/plot-biomass-maps-size-classes.r

@@ -1,0 +1,128 @@
+library(tidyverse)
+library(sf)
+library(allodb)
+
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+
+
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+
+deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
+
+quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
+
+deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
+
+
+spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
+
+missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
+
+canopyposition <- spTable  %>% 
+  select(spcode,canopy_position, life_form)  %>% 
+  mutate(canopy_position_new = if_else(life_form %in% c("tree"),"canopy","understory"))  %>%
+  select(sp = spcode, canopy_position = canopy_position_new)  %>% 
+  bind_rows(missing)
+
+newdf <- all_censuses  %>% 
+  left_join(canopyposition)  %>% 
+  filter(DFstatus %in% c("alive"))  %>% 
+  mutate(quadrat = factor(quadrat),
+         canopy_position = factor(canopy_position),
+         dbh_cutoffs = factor(cut(as.numeric(dbh), breaks = c(0,20,100,10000),labels = c("0-2cm","2-10cm","10+ cm")))) 
+
+plotdf_biomass <- function(df,census_num, position_class, size_class){
+    tempdf <- df  %>% 
+    filter(Census %in% census_num & canopy_position %in% position_class & dbh_cutoffs %in% size_class)  %>% 
+    group_by(quadrat, .drop = FALSE)  %>% 
[…]

### 2024-03-11 code/figure5_recr_by_species.r

@@ -15,7 +15,7 @@ load("data/census_data/all_censuses_agb.rdata")
 canopyposition <- spTable  %>% 
   select(spcode,canopy_position, life_form)  %>% 
   mutate(canopy_position_new = case_when(spcode %in% "acne" ~ "understory",
-                                         life_form %in% c("tree") ~ canopy,
+                                         life_form %in% c("tree") ~ "canopy",
                                         .default = "understory"))  %>%  
   select(sp = spcode, canopy_position = canopy_position_new)  %>% 
   bind_rows(missing)
@@ -71,13 +71,13 @@ f2 <-  wp_recr  %>%
   group_by(plot_sp,canopy_position)  %>% 
   summarize(n_stems = sum(n_stems))  %>% 
   filter(canopy_position %in% "understory")  %>% 
-  arrange(desc(AWR))  %>% 
+  arrange(desc(n_stems))  %>% 
   filter(plot_sp != "Other understory sp.")  %>% 
   pull(plot_sp)  %>% 
   c(.,"Other understory sp.")
 
-plotdf <- plt_recr_slice  %>%   
-  bind_rows(wp_recr)
+plotdf <- plt_recr_slice  #%>%   
+#  bind_rows(wp_recr)
 
 
 
@@ -89,7 +89,7 @@ fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,f
       ylab(expression(atop("# of Recruits", (Ha^-1~Yr^-1)))) +
       xlab("Species") + 
       guides(alpha = "none", fill = guide_legend(override.aes = list(size = 8))) +
-      scale_fill_manual(values = c("#E7BC40","#C7622B","#750000","#7e937f"), labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), name = "") + 
+      scale_fill_manual(values = c("#E7BC40","#C7622B","#750000","#7e937f"), labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), name = "") +  
       theme(axis.text = element_text(size = 14, angle = 45, hjust = 1),
             axis.title = element_text(size = 16),
             panel.grid.major.x = element_blank(),

### 2024-03-11 code/figure6_canopy_understory_biomass_changes.r
Figure 6 changes

@@ -8,30 +8,26 @@ grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped
 load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
 deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
 quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
-grouped_quadrats <- read.csv("data/grouped_quadrats.csv") %>%
-  mutate(quadrat = sprintf("%04d", quadrat))  %>% 
-  select(-X)
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
 
 missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
 
 canopyposition <- spTable  %>% 
   select(spcode,canopy_position, life_form)  %>% 
-  mutate(canopy_position_new = if_else(life_form %in% c("tree"),"canopy","understory"))  %>%
+  mutate(canopy_position_new = case_when(spcode %in% "acne" ~ "understory",
+                                         life_form %in% c("tree") ~ "canopy",
+                                        .default = "understory"))  %>%  
   select(sp = spcode, canopy_position = canopy_position_new)  %>% 
   bind_rows(missing)
 
 # sort species by canopy vs non-canopy status 
-all_censuses  %>% 
-  left_join(canopyposition)  %>% 
-  View()
 
-dbh_cutoff <- 50
+dbh_cutoff <- 30
 
 canopy_abg <- all_censuses  %>% 
   left_join(canopyposition)  %>% 
   filter(DFstatus %in% c("alive"))  %>%
-  filter(canopy_position %in% "canopy" & dbh >= dbh_cutoff)  %>%  
+  filter(canopy_position %in% "canopy" & as.numeric(dbh) >= dbh_cutoff)  %>%  
   mutate(quadrat = factor(quadrat), Census = factor(Census))  %>% 
   group_by(quadrat, Census, .drop = FALSE)  %>% 
   summarize(WoodyAGB = sum(ABG,na.rm = T) / 1000 / .04 * .47)
[…]

### 2024-03-11 doc/15yrsChange.Rmd
Update Author/Affiliations and Add Citation Checklist

@@
 
-Rachel Hoffman
+Rachel Hoffman^1^
 
-Krystal Bagnaschi
+Krystal Bagnaschi^1^
 
-Iris Kennedy
+Iris Kennedy^1^
 
-Erin MacMonigle
+Erin MacMonigle^1^
 
-Caroline Troy
+Caroline Troy^1^
 
-Cameron Dow
+Cameron Dow^1,3^
 
@@
 
-Jen Jordan
+Jennifer Jordan^1,4^
 
-Lukas Magee
+Lukas Magee^1,5^
 
-David Mitre
+David Mitre^2^
 
-Christopher Pate
+Christopher Pate^1^
 
-Jess Shue
+Jessica Shue^6^
 
-William McShea
+Norman A. Bourg^1^
+
+William J. McShea^1^
 
@@
 
-1. Conservation Ecology Center; Smithsonian's National Zoo & Conservation Biology Institute; Front Royal, Virginia 22630, USA
+1. Conservation Ecology Center, Smithsonian's National Zoo & Conservation Biology Institute, Front Royal, Virginia 22630, USA
 
 2. Forest Global Earth Observatory, Smithsonian Tropical Research Institute, Apartado Postal 0843-03092, Panama City, Panama
+
+3. Department of Biology, West Virginia University, Morgantown, West Virginia 26506, USA
+
+4. Department of _____, George Mason University, Fairfax, Virginia 22030, USA
+
+5. School of Forest, Fisheries and Geomatics Sciences, University of Florida, Gainesville, Florida 32603, USA
+
+6. Forest Global Earth Observatory, Smithsonian Environmental Research Center, Edgewater, Maryland 21037, USA
 

### 2024-03-13 doc/15yrsChange.Rmd
text results, text changes, fig 6 changes

@@
 
-**Title:** Nuisance species compromise the carbon sequestration potential in an Eastern US temperate deciduous forest
+**Title:** Nuisance species compromise carbon sequestration potential in an Eastern US temperate deciduous forest
 
@@
 
-The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with deciduous forests representing a substantial portion of this, sequestering >300 Tg C yr^-1^ [@pugh_role_2019].
-Although currently a C sink, the future of the biome remains uncertain.
-The dominant view imbued in Earth System Models is that, at least within the eastern US, the biome is likely to remain a C sink for the remainder of the 21st century [@finzi_carbon_2020; @wu_uncertainty_2023], albeit with declining CO~2~ sequestration capacity [@ahlström_robustness_2012].
+The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019].
+Although currently a net C sink, the future of the temperate forest biome remains uncertain.
+The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012].
 Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
-One mechanism that is not represented in global C models is the impact of nuisance species -- i.e., endemic or non-endemic species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
-Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning/niche models that seek to project future forest distribution and carbon stocks [**?** @wu_uncertainty_2023], nor in carbon offset projects [**???** @refs].
-This is problematic in that nuisance species -- including nonindigenous insect pests and pathogens, nonindigenous plants, and over-abundant herbivores -- are dramatically impacting carbon cycling in temperate deciduous forests around the world, as described below.
-"Tree mortality is one of the largest uncertainities in projections of future forest dynamics" (Bugmann et al. 2019)
+One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
+Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
+This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.
 
-Nuisance pests and pathogens can have a protracted and enduring impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories and geomorphic processes (Lovett et al.
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes (Lovett et al.
 2016, Fischer et al.
@@
 2014).
-The presence of nonindigenous forest insects and diseases, which have been historically introduced to new ecosystems through intercontinental trade, will continue to proliferate with increased globalization and human movement (Fischer et al.
+Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement (Fischer et al.
 2013, Aukema et al.
@@
 2006).
-Many of the prominent pests and pathogens in temperate deciduous forests have devastating effects on canopy composition and structure.
-Once abundant in temperate deciduous forests throughout North America, the American Chestnut (*Castanea dentata*) and the Ash genera (*Fraxinus spp.*) have been decimated by pests and pathogens, with stand-clearing disturbances occuring within a few years after initial infestation (Anderson 2020, Klooster et al.
-2013, Need a source for Ash abundance and length of time between infestation and mortality for Chestnuts).
-These previously dominant overstorey species are now confined to the understory as non-reproductive seeds and saplings (Elliot and Swank 2008, Anagnostakis 1987, Need a source for Ash).
-In other species, such as the Eastern Hemlock (*Tsuga canadensis*), the existence of pests and pathogens induces close to total population mortality, with considerable risk of regional extirpation (Ellison et al.
+Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species (Aukema et.
+al 2010,Anderson-Teixeria 2021, Klooster et al.
+2013, Ellison et al.
 2018).
-The loss of mature canopy species to nuisance species contributes to global change driven biomass losses, with biomass losses due to pest and pathogens likely comparable to those attributed to fire and forest removal (Fei et al.
-2013)  Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al.
-2013)
-
-Non-endemic pests and pathogens have been important driver of mortality [@anderson-teixeira_longterm_2021].
-These can reduce C and need to be considered in future climate change projections.
-
-@fei_biomass_2019
-
-Non-endemic pests and pathogens facilitate invasion of non-endemic plants [@guo_impacts_2023 and refs therein].
+Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al.
+2013).
 

### 2024-03-13 code/Woodyflux_Calculations.r

@@ -73,7 +73,7 @@ AWR <- all_censuses  %>%
   mutate(WoodyRecr = ABG / 5)  %>% 
   ungroup()  %>% 
   group_by(quadrat,Census)  %>% 
-  summarize(AWR = sum(WoodyRecr, na.rm = T) / 1000 / .04 * .47)
+  summarize(AWR = sum(WoodyRecr, na.rm = T) / 1000 / .04 * .47) 
 
 woody_fluxes <- AWP  %>% 
   left_join(AWM, by = c("quadrat" = "quadrat", "Census" = "Census"))  %>% 
@@ -107,11 +107,11 @@ AWR_sp <- all_censuses  %>%
   filter(as.numeric(dbh) <= 60)  %>%
   mutate(WoodyRecr = ABG / 5)  %>% 
   ungroup()  %>% 
-  group_by(Census,sp)  %>%
-  summarize(AWR = sum(WoodyRecr,na.rm = T) / 1000 /.47)
+  group_by(sp,quadrat)  %>%
+  summarize(AWR_yr = sum(WoodyRecr,na.rm = T) / 1000 /.47 / 15, n_stems_yr = n() / 15)
 
 
-write.csv(AWR_sp, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition.csv")
+write.csv(AWR_sp, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_data/RecruitmentComposition_annual.csv")
 
 
 

### 2024-03-13 code/figure2_agb.R

@@ -101,3 +101,7 @@ fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Grou
 ggsave(fig2,filename = "doc/display/Figure2.jpeg", units = "in",
         height = 8, width =7, dpi = 300)
 
+
+### Save text results for manuscript ###
+
+save(figure2_agb,file = "doc/results-text/Figure2_textdata.Rdata")

### 2024-03-13 code/figure3_woodyfluxes.R

@@ -74,3 +74,9 @@ ggsave(fig3,filename = "doc/display/Figure3.jpeg", units = "in",
        height = 8, width = 10, dpi = 300)
 
 
+### save results for text ###
+fig3_textresults  <- fig3_woodyfluxes  %>% 
+  select(Group, Census, Flux, MgC_Yr_Ha, cens_int)
+
+save(fig3_textresults, file = "doc/results-text/Figure3_textdata.Rdata")
+

### 2024-03-13 code/figure4_annual_mort_by_species.r

@@ -97,3 +97,7 @@ ggsave(barp,filename = "doc/display/Figure4.jpeg", units = "in",
 #   theme(plot.tag = element_text(size = rel(2), angle = 90),
 #         plot.tag.position = "left")
 
+##### IV - Save text results #####
+figure4_textresults <- mort_plot_df
+
+save(figure4_textresults,file = "doc/results-text/Figure4_textdata.Rdata")

### 2024-03-13 code/figure5_recr_by_species.r

@@ -16,8 +16,8 @@ canopyposition <- spTable  %>%
   select(spcode,canopy_position, life_form)  %>% 
   mutate(canopy_position_new = case_when(spcode %in% "acne" ~ "understory",
                                          life_form %in% c("tree") ~ "canopy",
-                                        .default = "understory"))  %>%  
-  select(sp = spcode, canopy_position = canopy_position_new)  %>% 
+                                         .default = "understory"))  %>%
+  select(sp = spcode, canopy_position = canopy_position_new)  %>%
   bind_rows(missing)
 
 recruitment_by_sp <- recruit  %>% 
@@ -98,7 +98,10 @@ fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,f
             legend.background = element_blank()
             )
 
-print(fig5)
-
+fig5
 ggsave(fig5,filename = "doc/display/Figure5.jpeg", units = "in", height = 8, width = 12, dpi = 300)
 
+##### IV - Save text results #####
+figure5_textresults <- plotdf
+
+save(figure5_textresults,file = "doc/results-text/Figure5_textdata.Rdata")

### 2024-03-13 code/figure6_canopy_understory_biomass_changes.r

@@ -70,9 +70,9 @@ understory_biomass <- understory_abg  %>%
   
 ##### Plotting #####
 #fillcols <- c("#b7c3e0", "#f3f2f2", "#ffc080","#85a4eb", "#e2dede", "#fd9e3f","#004494", "#a6a6a6", "#cc4c00")
-fillcols <- c("#85a4eb","#004494", "#e2dede", "#a6a6a6","#fd9e3f", "#cc4c00")
-bivariate_color_scale <- tibble(understory = rep(2:1, 3),
-                                canopy = rep(1:3, each = 2),
+fillcols <- c( "#85a4eb", "#800303","#004494","#fd5c5c")
+bivariate_color_scale <- tibble(understory = rep(2:1, 2),
+                                canopy = rep(1:2, each = 2),
                                 fill = fillcols)
 
 # bivariate_color_scale <- tibble(
@@ -98,9 +98,8 @@ plotdf <- quadrat_shp  %>%
   left_join(canopy_slopes)  %>% 
   left_join(understory_biomass)  %>% 
   mutate(WoodyAGB = if_else(is.na(WoodyAGB),0,WoodyAGB))  %>% 
-  mutate(canopy = case_when(canopy_trend >= 3 ~ 3,
-                              canopy_trend < 3 & canopy_trend > -3 ~ 2,
-                              canopy_trend <= -3 ~ 1),
+  mutate(canopy = case_when(canopy_trend >= 0 ~ 2,
+                            canopy_trend < 0 ~ 1),
          understory = case_when(WoodyAGB >= .02 ~ 2,
                                 WoodyAGB < .02 ~ 1 ))  %>% 
   left_join(bivariate_color_scale) 
@@ -131,7 +130,7 @@ legend <- ggplot() +
   labs(y = "Canopy biomass ⟶️",
        x = "Understory biomass \n of canopy sp. ⟶️") +
   scale_x_discrete(labels  = c("1" = "Low","2" = "High")) +
-  scale_y_discrete(labels  = c("1" = "Decreasing","2" = "Stable", "3" = "Increasing")) +
+  scale_y_discrete(labels  = c("1" = "Decreasing","2" = "Increasing")) +
   theme_classic() +
   # make font small enough
   theme(
@@ -172,7 +171,7 @@ barp_df2 <- barp_df  %>%
 
 barp <- ggplot(barp_df) +
[…]

### 2024-03-18 code/figure2_agb.R
Update Figure 2 and 5

@@ -1,7 +1,7 @@
 #figure 2 ABG (Aboveground Biomass)
 #output Line graph of aboveground biomass stocks
 
-#install.packages('allodb')
+install.packages('allodb')
 library(allodb)
 library(tidyverse)
 
@@ -89,9 +89,9 @@ fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Grou
   scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
     minor_breaks = c())+
   scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"),values = colz) + 
-  labs(y = "Carbon Stock (Mg C/Ha)", x = "Year") +
+  labs(y = "Aboveground Biomass (Mg C/ha)", x = "Year") +
   theme_bw() +
-  theme(legend.position = c(.3,.2),
+  theme(legend.position = c(.38,.175),
         legend.text = element_text(size = 18),
         legend.background = element_blank(),
         axis.text.x = element_text(size = 18, margin = margin(t = .15,r = 0, b = .05,l = 0,unit = "in")),

### 2024-03-18 code/figure5_recr_by_species.r

@@ -1,3 +1,4 @@
+install.packages("tidyverse")
 library(tidyverse) 
 library(patchwork)
 
@@ -23,6 +24,7 @@ canopyposition <- spTable  %>%
 recruitment_by_sp <- recruit  %>% 
   left_join(canopyposition)  %>% 
   mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
+  #mutate(sp = if_else(sp %in% c("qual", "quco", "qufa", "qumi", "qupr", "quru", "quve"), "Oak spp.", sp)) %>%
   group_by(quadrat,sp, canopy_position)  %>% 
   summarise(AWR_yr_ha = sum(AWR_yr) / .04, n_stems_yr_ha = sum(n_stems_yr) / .04)
 
@@ -76,30 +78,52 @@ f2 <-  wp_recr  %>%
   pull(plot_sp)  %>% 
   c(.,"Other understory sp.")
 
-plotdf <- plt_recr_slice  #%>%   
+plotdf <- plt_recr_slice
+
+##plotdf <- plt_recr_slice  #%>%   
 #  bind_rows(wp_recr)
 
+
+spTable_latinNames <- spTable %>%
+  mutate(scientific_name = paste(genus, species, sep = " ")) 
+
+plotdf$scientific_name <- if_else(plotdf$plot_sp %in% spTable_latinNames$spcode, 
+                                  spTable_latinNames$scientific_name[match(plotdf$plot_sp, spTable_latinNames$spcode)], NA)
+
+plotdf %>% 
+  mutate(scientific_name = case_when(
+        plot_sp == "Hickory spp." ~ "Carya spp.",
+        plot_sp == "Other canopy sp." ~ "Other canopy sp.",
+        plot_sp == "Other understory sp." ~ "Other understory sp.", 
+        TRUE ~ scientific_name
+  ) )
+
+plotdf %>%
+  mutate (plot_sp = scientific_name)
 
 
 
 fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,fill = as.ordered(Group), group = Group)) + 
-      facet_wrap(~canopy_position, scales = "free") + 
+      facet_wrap(~canopy_position, labeller = labeller(canopy_position = c("canopy" = "Canopy", "understory" = "Understory")), scales = "free") + 
[…]

### 2024-03-18 code/figure2_agb.R
update figures (pt. 2)

@@ -75,10 +75,79 @@ tot_trends <- quadrat_abg_2008  %>%
 figure2_agb <- group_trends  %>% 
   bind_rows(tot_trends)
 
+########## Attempt at fixing overlapping legend ################################
+
+figure2_agb_lwd <- figure2_agb %>%
+  mutate(line_width = case_when(
+    Group == "Whole Plot" ~ 1.5,
+    Group == "1" ~ 1.2,
+    Group == "2" ~ 1.2,
+    Group == "3" ~ 1.2))
+
+figure2_agb_lty <- figure2_agb_lwd %>%
+  mutate(line_type = case_when(
+    Group == "Whole Plot" ~ "solid", 
+    Group == "1" ~ "dotted", 
+    Group == "2" ~ "longdash",
+    Group == "3" ~ "longdash"))
+
+figure2_agb_pch <- figure2_agb_lty %>%
+  mutate(point_type = case_when(
+    Group == "Whole Plot" ~ "circle", 
+    Group == "1" ~ "triangle", 
+    Group == "2" ~ "triangle",
+    Group == "3" ~ "square"))
+
+figure2_agb_cex <- figure2_agb_pch %>%
+  mutate(point_size = case_when(
+    Group == "Whole Plot" ~ 3,
+    Group == "1" ~ 3,
+    Group == "2" ~ 3,
+    Group == "3" ~ 3))
+
+figure2_agb_fill <- figure2_agb_cex %>%
+  mutate(point_fill = case_when(
+    Group == "Whole Plot" ~ "#7e937f",
+    Group == "1" ~ "#E7BC40",
+    Group == "2" ~ "#C7622B",
+    Group == "3" ~ "#750000"))
+  
+figure2_agb <- figure2_agb_fill
+
 colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 
+fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, 
+                                group = Group, col = Group,
+                                linewidth = line_width,
+                                linetype = line_type,
+                                shape = point_type, 
+                                fill = point_fill)) +
+  geom_line() +
+  geom_point() +
+  scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
+                     minor_breaks = c())+
+  scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species",
+                                                       "High deer, low vulnerable species", 
[…]

### 2024-03-18 code/figure5_recr_by_species.r

@@ -83,6 +83,7 @@ plotdf <- plt_recr_slice
 ##plotdf <- plt_recr_slice  #%>%   
 #  bind_rows(wp_recr)
 
+###Attempt at changing species code to species name in ggplot###
 
 spTable_latinNames <- spTable %>%
   mutate(scientific_name = paste(genus, species, sep = " ")) 
@@ -101,7 +102,7 @@ plotdf %>%
 plotdf %>%
   mutate (plot_sp = scientific_name)
 
-
+################################################################
 
 fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,fill = as.ordered(Group), group = Group)) + 
       facet_wrap(~canopy_position, labeller = labeller(canopy_position = c("canopy" = "Canopy", "understory" = "Understory")), scales = "free") + 

### 2024-03-19 code/figure2_agb.R
Figure 2 and 5 tweaks

@@ -77,64 +77,22 @@ figure2_agb <- group_trends  %>%
 
 ########## Attempt at fixing overlapping legend ################################
 
-figure2_agb_lwd <- figure2_agb %>%
-  mutate(line_width = case_when(
-    Group == "Whole Plot" ~ 1.5,
-    Group == "1" ~ 1.2,
-    Group == "2" ~ 1.2,
-    Group == "3" ~ 1.2))
-
-figure2_agb_lty <- figure2_agb_lwd %>%
-  mutate(line_type = case_when(
-    Group == "Whole Plot" ~ "solid", 
-    Group == "1" ~ "dotted", 
-    Group == "2" ~ "longdash",
-    Group == "3" ~ "longdash"))
-
-figure2_agb_pch <- figure2_agb_lty %>%
-  mutate(point_type = case_when(
-    Group == "Whole Plot" ~ "circle", 
-    Group == "1" ~ "triangle", 
-    Group == "2" ~ "triangle",
-    Group == "3" ~ "square"))
-
-figure2_agb_cex <- figure2_agb_pch %>%
-  mutate(point_size = case_when(
-    Group == "Whole Plot" ~ 3,
-    Group == "1" ~ 3,
-    Group == "2" ~ 3,
-    Group == "3" ~ 3))
-
-figure2_agb_fill <- figure2_agb_cex %>%
-  mutate(point_fill = case_when(
-    Group == "Whole Plot" ~ "#7e937f",
-    Group == "1" ~ "#E7BC40",
-    Group == "2" ~ "#C7622B",
-    Group == "3" ~ "#750000"))
-  
-figure2_agb <- figure2_agb_fill
-
 colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 
 fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, 
-                                group = Group, col = Group,
-                                linewidth = line_width,
-                                linetype = line_type,
-                                shape = point_type, 
-                                fill = point_fill)) +
-  geom_line() +
-  geom_point() +
+                                group = Group, col = Group)) + 
+  geom_line(aes(linetype = Group, linewidth = Group)) + 
+  geom_point(aes(shape = Group, size = Group)) +
   scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
                      minor_breaks = c())+
   scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species",
[…]

### 2024-03-19 code/figure5_recr_by_species.r

@@ -1,4 +1,4 @@
-install.packages("tidyverse")
+#figure 5 Recruitment by Species
 library(tidyverse) 
 library(patchwork)
 
@@ -24,7 +24,7 @@ canopyposition <- spTable  %>%
 recruitment_by_sp <- recruit  %>% 
   left_join(canopyposition)  %>% 
   mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
-  #mutate(sp = if_else(sp %in% c("qual", "quco", "qufa", "qumi", "qupr", "quru", "quve"), "Oak spp.", sp)) %>%
+  mutate(sp = if_else(sp %in% c("qual", "quco", "qufa", "qumi", "qupr", "quru", "quve"), "Oak spp.", sp)) %>%
   group_by(quadrat,sp, canopy_position)  %>% 
   summarise(AWR_yr_ha = sum(AWR_yr) / .04, n_stems_yr_ha = sum(n_stems_yr) / .04)
 
@@ -91,16 +91,17 @@ spTable_latinNames <- spTable %>%
 plotdf$scientific_name <- if_else(plotdf$plot_sp %in% spTable_latinNames$spcode, 
                                   spTable_latinNames$scientific_name[match(plotdf$plot_sp, spTable_latinNames$spcode)], NA)
 
-plotdf %>% 
-  mutate(scientific_name = case_when(
-        plot_sp == "Hickory spp." ~ "Carya spp.",
-        plot_sp == "Other canopy sp." ~ "Other canopy sp.",
-        plot_sp == "Other understory sp." ~ "Other understory sp.", 
-        TRUE ~ scientific_name
-  ) )
+# plotdf %>%
+  # mutate(scientific_name = case_when(
+  # plot_sp == "Hickory spp." ~ "Carya spp.",
+  # plot_sp == "Oak spp." ~ "Quercus spp."
+  # plot_sp == "Other canopy sp." ~ "Other canopy sp.",
+  # plot_sp == "Other understory sp." ~ "Other understory sp.",
+  # FALSE ~ scientific_name
+ # ) )
 
-plotdf %>%
-  mutate (plot_sp = scientific_name)
+plotdf_final <- plotdf %>% 
+  mutate(scientific_name = if_else(is.na(scientific_name), plot_sp, scientific_name))
 
 ################################################################
 
[…]

### 2024-03-20 code/figure2_agb.R
Figure Updates

@@ -85,15 +85,15 @@ fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year,
   geom_point(aes(shape = Group, size = Group)) +
   scale_x_continuous(breaks = c(2008, 2013,2018, 2023), labels = c("2008","2013","2018","2023"),
                      minor_breaks = c())+
-  scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species",
-                                                       "High deer, low vulnerable species", 
-                                                       "High deer, high vulnerable species", 
-                                                       "Plot"),values = colz) +
+  scale_color_manual(name = element_blank(),labels = c("Low deer, low canopy vulnerability",
+                                                       "High deer, low canopy vulnerability", 
+                                                       "High deer, high canopy vulnerability", 
+                                                       "Whole plot"),values = colz) +
   scale_linetype_manual(values = c("dotted", "longdash", "longdash", "solid"), guide = "none") +
   scale_linewidth_manual(values = c(1.2, 1.2, 1.2, 1.5), guide = "none") + 
   scale_size_manual(values = c(3, 3, 3, 3), guide = "none") +
   scale_shape_manual(values = c("triangle", "triangle", "square", "circle"), guide = "none") +
-  labs(y = "Aboveground Biomass (Mg C/ha)", x = "Year") +
+  labs(y = expression("Aboveground Biomass" ~ (Mg~C~'/'~ha^-1)), x = "Year") +
   theme_bw() +
   theme(legend.position = c(.38,.175),
         legend.text = element_text(size = 18),
@@ -106,27 +106,27 @@ fig2
 
 ################################################################################
 
-fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Group)) +
-  geom_line(data = figure2_agb  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
-  geom_line(data = figure2_agb %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
-  geom_line(data = figure2_agb %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
[…]

### 2024-03-20 code/figure3_woodyfluxes.R

@@ -53,13 +53,13 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_line(data = fig3_woodyfluxes  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
   geom_line(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), pch = 25, cex = 3, fill = "#E7BC40") +
-  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 25, cex = 3, fill = "#C7622B") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("1")), pch = 17, cex = 3, fill = "#E7BC40") +
+  geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("2")), pch = 17, cex = 3, fill = "#C7622B") +
   geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
-  scale_color_manual(name = element_blank(),labels = c("Low deer, low vulnerable species","High deer, low vulnerable species", "High deer, high vulnerable species", "Plot"), values = colz, guide = guide_legend(nrow = 2)) +
-  labs(y = expression("Carbon Flux"~(Mg~C~Ha^-1~Yr^-1)), x = "Census") +
+  scale_color_manual(name = element_blank(),labels = c("Low deer, low canopy vulnerability","High deer, low canopy vulnerability", "High deer, canopy vulnerability", "Whole Plot"), values = colz, guide = guide_legend(nrow = 2)) +
+  labs(y = expression("Carbon Flux"~(Mg~C~ha^-1~yr^-1)), x = "Census") +
   theme_bw() +
   theme(legend.position = "top",
[…]

### 2024-03-20 code/figure5_recr_by_species.r

@@ -78,18 +78,16 @@ f2 <-  wp_recr  %>%
   pull(plot_sp)  %>% 
   c(.,"Other understory sp.")
 
-plotdf <- plt_recr_slice
-
-##plotdf <- plt_recr_slice  #%>%   
-#  bind_rows(wp_recr)
+plotdf <- plt_recr_slice #%>%   
+  #bind_rows(wp_recr)
 
 ###Attempt at changing species code to species name in ggplot###
 
-spTable_latinNames <- spTable %>%
-  mutate(scientific_name = paste(genus, species, sep = " ")) 
+#spTable_latinNames <- spTable %>%
+  #mutate(scientific_name = paste(genus, species, sep = " ")) 
 
-plotdf$scientific_name <- if_else(plotdf$plot_sp %in% spTable_latinNames$spcode, 
-                                  spTable_latinNames$scientific_name[match(plotdf$plot_sp, spTable_latinNames$spcode)], NA)
+#plotdf$scientific_name <- if_else(plotdf$plot_sp %in% spTable_latinNames$spcode, 
+                                  #spTable_latinNames$scientific_name[match(plotdf$plot_sp, spTable_latinNames$spcode)], NA)
 
 # plotdf %>%
   # mutate(scientific_name = case_when(
@@ -100,8 +98,8 @@ plotdf$scientific_name <- if_else(plotdf$plot_sp %in% spTable_latinNames$spcode,
   # FALSE ~ scientific_name
  # ) )
 
-plotdf_final <- plotdf %>% 
-  mutate(scientific_name = if_else(is.na(scientific_name), plot_sp, scientific_name))
+#plotdf_final <- plotdf %>% 
+  #mutate(scientific_name = if_else(is.na(scientific_name), plot_sp, scientific_name))
 
 ################################################################
 
@@ -126,7 +124,7 @@ fig5 <- ggplot(plotdf, aes(x = factor(plot_sp, levels = c(f1,f2)), y = n_stems,f
 
 
 fig5
-ggsave(fig5,filename = "doc/display/Figure5_withOakGenera.jpeg", units = "in", height = 8, width = 12, dpi = 300)
+ggsave(fig5,filename = "doc/display/Figure5.jpeg", units = "in", height = 8, width = 12, dpi = 300)
 
 ##### IV - Save text results #####
 figure5_textresults <- plotdf

### 2024-03-20 code/figure6_canopy_understory_biomass_changes.r

@@ -3,9 +3,9 @@ library(sf)
 library(allodb)
 
 ##### I - Read in data from ALL censuses and the species table #####
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+grouped_quadrats <- read.csv("data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+load("data/census_data/all_censuses_agb.rdata")
 deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
 quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")

### 2024-03-20 code/figure4_annual_mort_by_species.r
Figure tweaks

@@ -47,8 +47,8 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
   theme_bw() +
   #scale_fill_viridis_d()  +
   scale_fill_grey() +
-  ylab(expression("Aboveground Woody Mortality"~(Mg~C~Ha^-1~Yr^-1))) +
-  theme(axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
+  ylab(expression("Aboveground Woody Mortality"~(Mg~C~ha^-1~yr^-1))) +
+  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
         axis.text.y = element_text(size = 14),
         axis.title = element_text(size = 16),
         axis.title.x = element_blank(),
@@ -61,9 +61,7 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
         legend.position = "none") 
   
 
-print(barp)
-
-
+barp
 ggsave(barp,filename = "doc/display/Figure4.jpeg", units = "in",
         height = 6, width = 10, dpi = 300)
 

### 2024-03-20 doc/15yrsChange.Rmd
Update author list
reknitting with figure and text updates

@@
 
-Jennifer Jordan^1,4^
+Jennifer Jordan^1,4^ ( Orcid ID : 0000-0001-5521-360X)
 
@@
 
-4. Department of _____, George Mason University, Fairfax, Virginia 22030, USA
+4. Department of Environmental Science and Public Policy, George Mason University, Fairfax, Virginia 22030, USA
 
@@
 
-regen_failure = 6 #some figure illustrating absence of small canopy trees/ shifting  size distributions
+regen_failure = 6 #2-panel - map & barplot illustrating areas losing canopy species in the understory and biomass in the overstory
 
@@
 
-- white-tailed deer (*Odocoileus virginianus* ) is important endemic nuisance species, over-abundant because of human influence
-
-@mcgarvey_effects_2013
-@holm_interactive_2013
-@knauer_understory_2023
-
 The capacity to regenerate following disturbances, including ongoing gap formation through mortality of canopy trees, critically influences long-term forest dynamics.
@@
 
-Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses:
-(1) ) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB)
-(2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to nonindigenous nuisance species
-(3) Recruitment and woody growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
+Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest understory and canopy over extended time periods.
+Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses: (1) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB) (2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species (3) Recruitment and biomass gains from tree growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
 
@@
 
-The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).
+The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
 The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
-Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description).
+Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
 The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al.
@@
 
-* Add information about the history of pests and pathogens at this site and at similar forests
+In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
+(1) Low deer, low canopy vulnerability: A 4-ha portion of upland forest in the SE quarter of the plot, fenced in 1990 and maintained with only occasional deer presence for the past 24 years that has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm *et al.* 2013);
+(2) high deer, low canopy vulnerability: 16.3 ha of upland forest outside of the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens; 
+(3) high deer, high canopy vulnerability: 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. 
+Upland forest was defined according to the topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
+Vulnerability to non-indigenous insects and pathogens was defined based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment.
+The following canopy species were classified as affected: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Carya dentata*).
 
-In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
-(1) low deer, low non-endemic insects & pathogens - 4 ha portion of upland forest in the SE quarter of the plot, fenced in YEAR and maintained with only occasional deer presence for the past ## years and with relatively low abundance of canopy species affected by non-endemic pests and pathogens;
-(2) high deer, low non-endemic insects & pathogens # ha of upland forest outside the deer exclosure with relatively low abundance of canopy species affected by non-endemic pests and pathogens; 
-(3) high deer, high non-endemic insects & pathogens - a # ha section of upland forest outside the deer exclosure with relatively high abundance of canopy species affected by non-endemic pests and pathogens.
-These areas were delineated as follows.
-Upland forest was defined according to topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
-Vulnerability to non-endemic insects & pathogens was defined based on the abundance of canopy species affected by non-endemic pests and pathogens at the time of plot establishment (2008).
-
-![**Figure `r map_and_photos`.
-Photos within the SCBI ForestGEO plot: (a) low deer, low non-endemic insects & pathogens, (b) high deer, low non-endemic insects & pathogens, (c) high deer, high non-endemic insects & pathogens ** *(add a map of the plot)* All photos taken September 2023 by K.
-Anderson-Teixeira.](display/Figure1.png)
+![**_Figure `r map_and_photos`.
+Map and photos of the Forest Global Earth Observatory (ForestGEO) plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA._** Our analysis focuses on three habitat types: (a) low deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, (b) high deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, and (c) high deer density, high abundance of canopy trees vulnerable to non-indigenous insects & pathogens.
+*All photos taken September 2023 by K.
+Anderson-Teixeira; aerial photography from ESRI World Imagery*.](display/Figure1.png)
 
@@
 
-A number of field surveys are regularly conducted at the site, two of which are included in our analysis.
-As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).
-This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; dbh).
-The census records information regarding the dbh, species, status and location of each stem included in the survey.
-Each invidual is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field.
-In multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.
-Location is recorded on a digitized map of the plot on the ESRI ArcGIS FieldMaps application (*citation*).
-Established in 2008, there have been four censuses at the site, comprising 15 years of detailed forest dynamics data.
-In addition to the ForestGEO census, since 2014 an annual mortality survey has been conducted on the site.
-Initiated, in part, as a response to the presence of Emerald Ashborer, the mortality survey is conducted on all trees that had a dbh greater than 10 centimeters in the previous full census.
-Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.
+As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in Condit (1998).
+This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh).
+The census records information regarding the dbh, species, living status and location of each stem included in the survey.
+Each individual is marked with an identifying number for sequential data collection in subsequent censuses.
+In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag.
+The plot was established in 2008 and recensused three times since, comprising 15 years of forest dynamics data.
+In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated by GitHub actions (Kim et al., 2022).
+Each individual's location was recorded in the Field Maps application using a digitized map of the plot (*citation*).
+Additional checks and fixes were performed at the conclusion of the last census.
+In addition to the ForestGEO census, since 2014, a tree mortality survey has been conducted annually at the site.
+Initiated, in part, as a response to the presence of emerald ash borer (*Agrilus planipennis*), the mortality survey is conducted on all trees that had a dbh $\geq$ 10 centimeters in the previous census.
+Through assessing individual tree health, we document trends in forest mortality and identify factors that are associated with death.
 Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
 
-To improve the quality of the data, GitHub Actions continuous integration was implemented for quality control (Kim et al., 2022).
-Data from the previous census was used to populate the Field Maps app, and fields were created to input data.
-There were checks built into the app to help catch errors, for example the DBH field would flag a suspiciously high number and would not allow the form to be submitted if it was blank.
-To use the app in the field the census crew downloaded offline areas of where they were working and synced them to an online map after collecting data.
-Next the data was downloaded from ArcGIS online and pushed to GitHub, where it triggered a GitHub Action running data quality assurance and control (QA/QC) scripts.
-The scripts were written in R and checked the data for errors such as “suspicious positive growth.” After going through the QA/QC process, a csv of errors and warnings with the associated stem were output which could then be corrected in the field.
+## Calculating biomass dynamics and stem recruitment
 
-## Analyses
+We estimated total aboveground biomass (AGB) from the measured dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024).
+*allodb* creates species-specific allometric equations through weighting of all applicable allometries within its database filtered by tree size, geography, and taxa information.
+Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot regions and the plot as a whole.
+We evaluated trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region.
+We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.
 
-* We  
-*
+In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems.
+We categorized each species as either canopy or understory based on expert knowledge of life history traits.
+*Quercus* spp.
+and *Carya* spp.
+were grouped by genera to better assess differences between the recruitment of canopy and understory species.
+We then calculated the number of new stems for each census interval for each  species or genera.
+Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species.
+For each .4 hectare quadrat, we calculated the net change in AGB from 2008 to 2023 and AGB stock of canopy species in the understory in 2023.
 
@@
 
-![**Figure `r biomass_trends`.
-Aboveground biomass decreased at the site for the first time since establishment.
-The net decrease at the plot-level was driven by large losses in regions with high vulnerability to nuisance species.**](display/Figure2.jpeg)
+![**_Figure `r biomass_trends`.
+Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)
 
@@
 
-![**Figure `r loss_gain_net`.
-Fluxes of aboveground woody biomass.
-Large increases in woody mortality combined with sustained decreases in woody productivity result in decreasing net biomass in the most recent census.**](display/Figure3.jpeg)
+![**_Figure `r loss_gain_net`.
+Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)
 
@@
 
-![**Figure `r mortality_by_genera`. Annual woody mortality flux by species/genera. We report large increases in mortality of white ash, red oak, and black oak.**](display/Figure4.jpeg)
+![**_Figure `r mortality_by_genera`.
+Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the seven canopy taxa with highest average mortality rate in descending order and a category including all others.
+The census included all stems ≥ 10cm in the most recent full census.](display/Figure4.jpeg)
 
@@
 
-![**Figure `r recruitment_by_genera`.
-Annual recruitment of canopy & understory species/genera across all censuses.
-Recruitment of canopy species is significantly higher within the deer exclosure.
-Recruitment of understory species is 20x that of canopy species.**](display/Figure5.jpeg)
+![**_Figure `r recruitment_by_genera`.
+Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment >20x that of canopy species recruitment.](display/Figure5.jpeg)
 
@@
 
-@holtmann_carbon_2021
+We found that the combination of decreasing biomass (driven by increasing woody mortality) and limited recruitment of canopy species in the understory is compromising the potential the carbon sink, now and into the future.
+Differences in AGB were the biggest in the regions of the plot with high vulnerability to nuisance species.
+These regions also had the lowest amount of canopy species in the understory.
+Areas with low deer population had the highest amount of canopy species in the understory.
+Some of the plot is increasing biomass, notably those with low proportions of vulnerable species. On net, these increases are not outpacing mortality. 
+Suggests that the combination of heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to pathogen-derived mortality where understory species that are resistant to deer and fast growers can take advantage, further competing with canopy species.
+While Emerald Ash-borer has had a large impact on biomass loss at our site, mortality among oak species, which is not linked to a non-indigenous pest, is substantial as well.
+Oak species in the region are generally found in older, even aged cohorts, which contributes to accelerated mortality rates (Oak et al., 2016).
+However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
+These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
+Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).
+
+Using current understory as a predictor of the future canopy is imperfect. 
+Some canopy species (examples) recruit in pulse events - if there is a shift in deer population and enough survival that coincides with a mast year than you could get regeneration into the canopy.
+However, current recruitment levels are below a rate at which we would expect in maturing secondary forests and may be insufficient to prevent large periods without a close canopy.
+Our data is from just the single site, which can make generalizability challenging. However, there is a lot of support that the drivers we identify here are representative of the region.
+Mixed hardwood forests are typical of mid-Atlantic forests and deer populations are high throughout the region.
+Vulnerable species (ash) are also common throughout these forests, and emerald ash borer continues to spread. Where it does spread it consistently decimates adult ash.
+Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forests. 
+Nuisance species (driven by human pressures/actions) interact with other anthropogenic changes to modify forest ecosystems. 
+Have the capacity to alter short-term dynamics: large carbon losses that would not be expected in models that don’t exclude it.
+Longer-term have capacity to alter successional dynamics: can delay recovery from disturbances (including other disturbances) or even suppress regeneration into a closed-canopy for sustained periods. 
+Decrease resilience to other pressures as pests and pathogens decrease diversity through shrinking the number of possible canopy species.
+Positive takeaways: this work highlights the potential of active management. Preserving the forests that we have is more valuable than planting new forests. 
+Pests and pathogens are driving mortality - may be opportunities for treatment/prevention of spread. 
+Research into natural immunity (ash species in original range of EAB have immunity).
+Management of deer populations to decrease browsing pressure - deer exclosure shows the importance to maintain an understory.
 
 # Conclusions /Future Directions
+
+Expand analyses across other temperate forests.
+Different forest compositions and herbivory dynamics, but many pests and pathogens.
+Leverage other forest dynamics sites to quantify the impact of nuisance species across large regions.
+Investigate productivity responses of surviving canopy. We would expect survivors adjacent to gaps increase productivity as competition decreases, but may be hampered by climate change.
+Interactions of global change drivers are critical to understand (though difficult to study).
 

### 2024-03-21 code/figure3_woodyfluxes.R
Figure tweaks

@@ -41,10 +41,9 @@ fig3_woodyfluxes <- group_trends  %>%
   )
 
 
-
 colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
-                            "AWP" = "Aboveground Woody Productivity",
+                            "AWP" = "Aboveground Woody Growth",
                             "AWR" = "Aboveground Woody Recruitment",
                             "NetFlux" = "Net Biomass Change"))
 
@@ -58,7 +57,7 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
-  scale_color_manual(name = element_blank(),labels = c("Low deer, low canopy vulnerability","High deer, low canopy vulnerability", "High deer, canopy vulnerability", "Whole Plot"), values = colz, guide = guide_legend(nrow = 2)) +
+  scale_color_manual(name = element_blank(),labels = c("Low deer, low canopy vulnerability","High deer, low canopy vulnerability", "High deer, high canopy vulnerability", "Whole Plot"), values = colz, guide = guide_legend(nrow = 2)) +
   labs(y = expression("Carbon Flux"~(Mg~C~ha^-1~yr^-1)), x = "Census") +
   theme_bw() +
   theme(legend.position = "top",

### 2024-03-21 code/figure4_annual_mort_by_species.r

@@ -5,7 +5,7 @@ allmort_awm <- read.csv("data/processed_data/MortalityComposition.csv")
 
 ##### I - Identify top species (by mortality flux) & reformat data for plotting #####
 top_sp <- allmort_awm  %>% 
-  mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
+  #mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
   ungroup()  %>% 
   group_by(sp)  %>% 
   mutate(abgmort = if_else(is.na(abgmort),0, abgmort))  %>% 
@@ -33,15 +33,20 @@ mort_plot_df$plotspecies <- gsub('qupr', 'Quercus prinus', mort_plot_df$plotspec
 mort_plot_df$plotspecies <- gsub('quru', 'Quercus rubra', mort_plot_df$plotspecies)
 mort_plot_df$plotspecies <- gsub('quve', 'Quercus velutina', mort_plot_df$plotspecies)
 
-facet_order <- c("Fraxinus americana", "Quercus velutina", "Quercus rubra", "Carya spp.", 
+facet_order <- c("Fraxinus americana", "Quercus velutina", "Quercus rubra", 
                  "Quercus prinus", "Quercus alba", "Liriodendron tulipifera", "Other")
 
-facet_years <- c(2013, 2018, 2023)
+facet_years <- c(2013, 2023)
+
+#custom_labels <- function(plotspecies, facet_order) {
+  #if_else(facet_order == "Other", as.character("Other"), paste0("*", facet_order, "*"))
+#}
 
 ##### II - Plot mortality by species #####
 
 barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
   facet_grid(~ factor(plotspecies, levels = facet_order)) +
+             #labeller = as_labeller(custom_labels)
   scale_x_continuous(breaks = facet_years) +
   geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
   theme_bw() +
@@ -54,7 +59,7 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
         axis.title.x = element_blank(),
         panel.spacing.x = unit(0, "lines"),
         #strip.background = element_blank(),
-        strip.text = element_text(size = 8, 
+        strip.text = element_text(size = 9.5, 
[…]

### 2024-03-21 code/figure2_agb.R
figure 2 tweaks

@@ -93,20 +93,19 @@ fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year,
   scale_linewidth_manual(values = c(1.2, 1.2, 1.2, 1.5), guide = "none") + 
   scale_size_manual(values = c(3, 3, 3, 3), guide = "none") +
   scale_shape_manual(values = c("triangle", "triangle", "square", "circle"), guide = "none") +
-  labs(y = expression("Aboveground Biomass" ~ (Mg~C~'/'~ha^-1)), x = "Year") +
+  labs(y = expression("Aboveground Biomass" ~ (Mg~C~ha^-1)), x = "Year") +
   theme_bw() +
   theme(legend.position = c(.38,.175),
-        legend.text = element_text(size = 18),
+        legend.text = element_text(size = 12),
         legend.background = element_blank(),
-        axis.text.x = element_text(size = 18, margin = margin(t = .15,r = 0, b = .05,l = 0,unit = "in")),
-        axis.text.y = element_text(size = 18, margin = margin(t = 0,r = .08, b = 0,l = 0, unit = "in")),
-        axis.title = element_text(size = 20))
+        axis.text.x = element_text(size = 13, margin = margin(t = .15,r = 0, b = .05,l = 0,unit = "in")),
+        axis.text.y = element_text(size = 13, margin = margin(t = 0,r = .08, b = 0,l = 0, unit = "in")),
+        axis.title = element_text(size = 15))
 
 fig2
 
 ################################################################################
 
-# fig2 <- ggplot(figure2_agb, aes(y=Abg_C_Mg_Ha, x=Year, group = Group, col = Group)) +
 #   geom_line(data = figure2_agb  %>% filter(Group %in% c("Whole Plot")), lwd = 1.5, lty = 1) + 
 #   geom_line(data = figure2_agb %>% filter(Group %in% c("2", "3")), lwd = 1.2, lty = 5) + 
 #   geom_line(data = figure2_agb %>% filter(Group %in% c("1")), lwd = 1.2, lty = 3) +
@@ -114,22 +113,10 @@ fig2
 #   geom_point(data = figure2_agb %>% filter(Group %in% c("2")), pch = 25, cex = 3, fill = "#C7622B") +
 #   geom_point(data = figure2_agb %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
[…]

### 2024-03-21 code/figure5_recr_by_species.r
added deer exclosure shp and figure 5 changes

@@ -1,6 +1,8 @@
 #figure 5 Recruitment by Species
 library(tidyverse) 
 library(patchwork)
+library(glue)
+library(ggtext)
 
 ##### I - Plot recruitment by species #####
 recruit <- read.csv("data/processed_data/RecruitmentComposition_annual.csv")
@@ -9,24 +11,34 @@ grouped_quadrats <- read.csv("data/grouped_quadrats.csv")  %>%
   mutate(quadrat = sprintf("%04d",quadrat))
 
 
+spTable <- spTable %>%
+  mutate(scientific_name = paste(genus, species, sep = " ")) 
+
 missing <- data.frame(sp = c("acsp","casp","crsp","frsp","prsp","ulsp"), canopy_position = "canopy")
 
 load("data/census_data/all_censuses_agb.rdata")
 
 canopyposition <- spTable  %>% 
-  select(spcode,canopy_position, life_form)  %>% 
+  select(spcode,canopy_position, life_form, scientific_name)  %>% 
   mutate(canopy_position_new = case_when(spcode %in% "acne" ~ "understory",
                                          life_form %in% c("tree") ~ "canopy",
                                          .default = "understory"))  %>%
-  select(sp = spcode, canopy_position = canopy_position_new)  %>%
+  select(sp = spcode, canopy_position = canopy_position_new,scientific_name)  %>%
   bind_rows(missing)
 
 recruitment_by_sp <- recruit  %>% 
   left_join(canopyposition)  %>% 
   mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
-  mutate(sp = if_else(sp %in% c("qual", "quco", "qufa", "qumi", "qupr", "quru", "quve"), "Oak spp.", sp)) %>%
+  mutate(sp = if_else(sp %in% c("qual", "quco", "qufa", "qumi", "qupr", "quru", "quve"), "Oak spp.", sp),
+         sp = case_when(
+                                     sp == "Hickory spp." ~ "Carya spp.",
+                                     sp == "Oak spp." ~ "Quercus spp.",
+                                     sp == "Other canopy sp." ~ "Other canopy spp.",
+                                     sp == "Other understory sp." ~ "Other understory spp.",
+                                     .default = scientific_name)) %>%  
[…]

### 2024-03-21 code/figure6_canopy_understory_biomass_changes.r

@@ -6,7 +6,7 @@ library(allodb)
 grouped_quadrats <- read.csv("data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
 load("data/census_data/all_censuses_agb.rdata")
-deer_exclosure <- st_read("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/spatial_data/shapefiles/deer_exclosure_2011.shp")
+deer_exclosure <- st_read("data/20m_grid/deer_exclosure_2011.shp")
 quadrat_shp <- read_sf("data/20m_grid/20m_grid.shp")  
 spTable <- read.csv("https://raw.githubusercontent.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/master/species_lists/Tree%20ecology/SCBI_ForestGEO_sp_ecology.csv")
 

### 2024-03-21 doc/15yrsChange.Rmd
added deer exclosure shp and figure 5 changes
figure updates & adding #s to results

@@
 # DISPLAY ITEMS
-## I do not see any guidance as to the number of display items allowed in MEE.
 
 #Tables
-## example = 1
 
@@
 regen_failure = 6 #2-panel - map & barplot illustrating areas losing canopy species in the understory and biomass in the overstory
+
+```
+
+```{r eval = TRUE, echo = FALSE, warning = FALSE, results = 'hide'}
+library(tidyverse)
+### Numbers for text
+##load in text data
+load("results-text/Figure2_textdata.Rdata")
+load("results-text/Figure3_textdata.Rdata")
+load("results-text/Figure4_textdata.Rdata")
+load("results-text/Figure5_textdata.Rdata")
+
+##Figure 2
+
+#whole plot
+wp_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+wp_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha) %>% round(digits = 1)
+
+#high deer, high canopy vulnerability
+hdhn_ABG_2013 <- figure2_agb  %>% filter(Year == "2013" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+
+#low canopy vulnerability
+ldlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+ldlv_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "2")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1) 
+hdlv_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "2")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
 
@@
 For the first time since the establishment of  the SCBI ForestGEO site, whole-plot aboveground C stocks decreased (Figure 2).
-From 2018 to 2023, whole-plot C stocks have decreased from **"X to X"** Mg C/ha.
-Both low-nuisance regions show similar decreases in C stocks going from **"X to X"** and **"X to X"** Mg C/ha, respectively.
-In the high deer high nuisance regions C stocks have decreased for the past two censuses, with a higher magnitude decrease of **"X to X"** Mg C/ha from 2018 to 2023.
+From 2018 to 2023, whole-plot C stocks have decreased from `r wp_ABG_2018` to `r wp_ABG_2023` Mg C ha^-1^.
+In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses, from `r hdhn_ABG_2013` Mg C ha^-1^ in 2013, to `r hdhn_ABG_2018` Mg C ha^-1^ in 2018, and `r hdhn_ABG_2023` Mg C ha^-1^ in 2023.
+Areas with low density of vulnerable canopy species experienced minor decreases in AGB; C stocks inside and outside of the deer exclosure decreased from **"X to X"**  and **"X to X"** Mg C ha^-1^, respectively.
 C stock decrease **is related to/ is explained by** above ground net biomass change.
@@
 
-![**Figure `r regen_failure`.**](display/Figure6.jpeg)
+![**_Figure `r regen_failure`._**](display/Figure6.jpeg)
 

### 2024-03-21 code/figure4_annual_mort_by_species.r

@@ -1,6 +1,7 @@
 #install.packages("patchwork")
 library(tidyverse)
 library(patchwork)
+library(ggtext)
 allmort_awm <- read.csv("data/processed_data/MortalityComposition.csv")
 
 ##### I - Identify top species (by mortality flux) & reformat data for plotting #####
@@ -43,10 +44,13 @@ facet_years <- c(2013, 2023)
 #}
 
 ##### II - Plot mortality by species #####
+plotdf <- mort_plot_df  %>% 
+  mutate(plotspecies = factor(plotspecies, levels = facet_order))
+facet_labels <- if_else(!grepl("Other",facet_order), paste0("*",facet_order,"*"),facet_order)
+names(facet_labels) <- facet_order
 
-barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
-  facet_grid(~ factor(plotspecies, levels = facet_order)) +
-             #labeller = as_labeller(custom_labels)
+barp <- ggplot(plotdf, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +
+  facet_grid(~ plotspecies, labeller =  labeller(plotspecies = facet_labels )) +
   scale_x_continuous(breaks = facet_years) +
   geom_bar(stat = "identity" , position = "dodge", col = "grey20") + 
   theme_bw() +
@@ -59,8 +63,7 @@ barp <- ggplot(mort_plot_df, aes(x = survey_year, y = mort_woody, fill = as.orde
         axis.title.x = element_blank(),
         panel.spacing.x = unit(0, "lines"),
         #strip.background = element_blank(),
-        strip.text = element_text(size = 9.5, 
-                                  face = "italic"), 
+        strip.text = element_markdown(size = 9.5), 
         panel.grid.major.x = element_blank(),
         panel.grid.minor.x = element_blank(),
         legend.position = "none") 

### 2024-03-21 doc/15yrsChange.Rmd
adding data objects to the rmarkdown

@@
 
-#low canopy vulnerability
+#high deer, low canopy vulnerability
+hdlv_AGB_2013 <- figure2_agb %>% filter(Year == "2013" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+hdlv_AGB_2018 <- figure2_agb %>% filter(Year == "2018" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+hdlv_AGB_2023 <- figure2_agb %>% filter(Year == "2023" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+
+#low deer, low canopy vulnerability
 ldlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
@@
 
+##Figure 3
+
+#Net Biomass Change 
+#Group 1
+nbc_group1_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "1" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group1_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "1" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group1_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "1" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 2
+nbc_group2_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "2" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group2_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "2" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group2_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "2" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 3 
+nbc_group3_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "3" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group3_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "3" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_group3_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "3" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Whole Plot
+nbc_wholePlot_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "Whole Plot" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_wholePlot_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "Whole Plot" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+nbc_wholePlot_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "Whole Plot" & Flux == "NetFlux") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Aboveground Woody Growth 
+#Group 1
+awg_group1_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "1" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group1_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "1" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group1_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "1" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 2
+awg_group2_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "2" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group2_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "2" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group2_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "2" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 3
+awg_group3_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "3" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group3_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "3" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_group3_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "3" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Whole Plot
+awg_wholePlot_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "Whole Plot" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_wholePlot_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "Whole Plot" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awg_wholePlot_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "Whole Plot" & Flux == "AWP") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Aboveground Woody Mortality 
+#Group 1
+awm_group1_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "1" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group1_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "1" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group1_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "1" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 2
+awm_group2_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "2" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group2_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "2" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group2_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "2" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 3
+awm_group3_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "3" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group3_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "3" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_group3_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "3" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Whole Plot
+awm_wholePlot_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "Whole Plot" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_wholePlot_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "Whole Plot" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awm_wholePlot_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "Whole Plot" & Flux == "AWM") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Aboveground Woody Recruitment
+#Group 1
+awr_group1_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group1_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group1_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 2
+awr_group2_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group2_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group2_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Group 3
+awr_group3_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group3_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_group3_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+#Whole Plot
+awr_wholePlot_2013 <- fig3_woodyfluxes %>% filter(Census == "2" & Group == "Whole Plot" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_wholePlot_2018 <- fig3_woodyfluxes %>% filter(Census == "3" & Group == "Whole Plot" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+awr_wholePlot_2023 <- fig3_woodyfluxes %>% filter(Census == "4" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
+
+##Figure 4
+
+#Averages
+
+#5 Year Averages for top three species
+
+#Peak for each species
 ```

### 2024-03-22 doc/15yrsChange.Rmd
adding data objects to the rmarkdown

@@
 
-#5 Year Averages for top three species
+avg_fram <- allmort_awm %>% filter(sp == "fram") %>% pull(abgmort) %>% 
+  mean(na.rm = TRUE)
+avg_quve <- allmort_awm %>% filter(sp == "quve") %>% pull(abgmort) %>% 
+  mean(na.rm = TRUE)
+avg_quru <- allmort_awm %>% filter(sp == "quru") %>% pull(abgmort) %>%
+  mean(na.rm = TRUE)
+avg_qupr <- allmort_awm %>% filter(sp == "qupr") %>% pull(abgmort) %>% 
+  mean(na.rm = TRUE)
+avg_qual <- allmort_awm %>% filter(sp == "qual") %>% pull(abgmort) %>% 
+  mean(na.rm = TRUE)
+avg_litu <- allmort_awm %>% filter(sp == "litu") %>% pull(abgmort) %>% 
+  mean(na.rm = TRUE)
+avg_other <- allmort_awm %>%  mutate(plot_species = if_else(sp %in% top_sp$sp,sp, "Other")) %>% filter(plot_species == "Other") %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+#5 Year Averages for top three species -- Currently calculated with survey_years 2014-2023
+
+avg_fram_first5 <- allmort_awm %>% filter(sp == "fram" & (survey_year == "2014" | survey_year == "2015" | survey_year == "2016" | survey_year == "2017" | survey_year == "2018")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+avg_fram_second5 <- allmort_awm %>% filter(sp == "fram" & (survey_year == "2019" | survey_year == "2020" | survey_year == "2021" | survey_year == "2022" | survey_year == "2023")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+avg_quve_five5 <- allmort_awm %>% filter(sp == "quve" & (survey_year == "2014" | survey_year == "2015" | survey_year == "2016" | survey_year == "2017" | survey_year == "2018")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+avg_quve_second5 <- allmort_awm %>% filter(sp == "quve" & (survey_year == "2019" | survey_year == "2020" | survey_year == "2021" | survey_year == "2022" | survey_year == "2023")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+avg_quru_first5 <-  allmort_awm %>% filter(sp == "quru" & (survey_year == "2014" | survey_year == "2015" | survey_year == "2016" | survey_year == "2017" | survey_year == "2018")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
+
+avg_quru_second5 <- allmort_awm %>% filter(sp == "quru" & (survey_year == "2019" | survey_year == "2020" | survey_year == "2021" | survey_year == "2022" | survey_year == "2023")) %>% pull(abgmort) %>% mean(na.rm = TRUE)
 
 #Peak for each species
+
+fram_peak <-  allmort_awm %>% filter(sp == "fram") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+quve_peak <- allmort_awm %>% filter(sp == "quve") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+quru_peak <- allmort_awm %>% filter(sp == "quru") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+qupr_peak <- allmort_awm %>% filter(sp == "qupr") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+qual_peak <- allmort_awm %>% filter(sp == "quru") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+litu_peak <- allmort_awm %>% filter(sp == "litu") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+other_peak <- allmort_awm %>%  mutate(plot_species = if_else(sp %in% top_sp$sp,sp, "Other")) %>% filter(plot_species == "Other") %>% pull(abgmort) %>% max(na.rm = TRUE)
+
+##Figure 5
+
+#Total number of recruits in canopy and understory for each region
+
+#Canopy
+group1_canopy <- plotdf %>% filter(Group == "1" & canopy_position == "canopy")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+group2_canopy <- plotdf %>% filter(Group == "2" & canopy_position == "canopy")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+group3_canopy <- plotdf %>% filter(Group == "3" & canopy_position == "canopy")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+
+#Understory
+group1_understory <- plotdf %>% filter(Group == "1" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+group2_understory <- plotdf %>% filter(Group == "2" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+group3_understory <- plotdf %>% filter(Group == "3" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+
+#Number of recruits for oak spp, hickory spp, libe and astr in each of the groups
+
+#Oak recruits
+group1_oakRecruit <- plotdf %>% filter(Group == "1" & plot_sp == "Quercus spp.")%>% pull(n_stems)
+group2_oakRecruit <- plotdf %>% filter(Group == "2" & plot_sp == "Quercus spp.") %>% pull(n_stems)
+group3_oakRecruit <- plotdf %>% filter(Group == "3" & plot_sp == "Quercus spp.")%>% pull(n_stems)
+
+#Hickory Recruits
+group1_hickoryRecruit <- plotdf %>% filter(Group == "1" & plot_sp == "Carya spp.")%>% pull(n_stems)
+group2_hickoryRecruit <- plotdf %>% filter(Group == "2" & plot_sp == "Carya spp.")%>% pull(n_stems)
+group2_hickoryRecruit <- plotdf %>% filter(Group == "3" & plot_sp == "Carya spp.")%>% pull(n_stems)
+
+#libe Recruits
+group1_libeRecruit <- plotdf %>% filter(Group == "1" & plot_sp == "Lindera benzoin") %>% pull(n_stems)
+group2_libeRecruit <- plotdf %>% filter(Group == "2" & plot_sp == "Lindera benzoin") %>% pull(n_stems)
+group3_libeRecruit <- plotdf %>% filter(Group == "3" & plot_sp == "Lindera benzoin") %>% pull(n_stems)
+
+#astr Recruits
+group1_astrRecruit <- plotdf %>% filter(Group == "1" & plot_sp == "Asimina triloba") %>% pull(n_stems)
+group2_astrRecruit <- plotdf %>% filter(Group == "2" & plot_sp == "Asimina triloba") %>% pull(n_stems)
+group3_astrRecruit <- plotdf %>% filter(Group == "3" & plot_sp == "Asimina triloba") %>% pull(n_stems)
+
 ```

### 2024-03-22 code/figure4_annual_mort_by_species.r

@@ -13,7 +13,6 @@ top_sp <- allmort_awm  %>%
   summarize(abgmort = sum(abgmort))  %>% 
   arrange(desc(abgmort))  %>% 
   slice_head(n = 7)
-  
 
 allspyr <- allmort_awm  %>% 
   expand(sp,survey_year)

### 2024-03-26 doc/results-text-files.R
move results objects to their own script

@@ -1,0 +1,211 @@
+library(tidyverse)
+### Numbers for text
+##load in text data
+load("results-text/Figure2_textdata.Rdata")
+load("results-text/Figure3_textdata.Rdata")
+load("results-text/Figure4_textdata.Rdata")
+allmort_awm <- read.csv("../data/processed_data/MortalityComposition.csv")
+load("results-text/Figure5_textdata.Rdata")
+
+##Figure 2
+
+#whole plot
+wp_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+wp_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha) %>% round(digits = 1)
+
+#high deer, high canopy vulnerability
+hdhn_ABG_2013 <- figure2_agb  %>% filter(Year == "2013" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+
+#high deer, low canopy vulnerability
+hdlv_AGB_2013 <- figure2_agb %>% filter(Year == "2013" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+hdlv_AGB_2018 <- figure2_agb %>% filter(Year == "2018" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+hdlv_AGB_2023 <- figure2_agb %>% filter(Year == "2023" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
+round(digits = 1)
+
+#low deer, low canopy vulnerability
+ldlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
[…]

### 2024-03-27 doc/results-text-files.R
update fig 4 results objects

@@ -1,11 +1,10 @@
 library(tidyverse)
 ### Numbers for text
 ##load in text data
-load("results-text/Figure2_textdata.Rdata")
-load("results-text/Figure3_textdata.Rdata")
-load("results-text/Figure4_textdata.Rdata")
-allmort_awm <- read.csv("../data/processed_data/MortalityComposition.csv")
-load("results-text/Figure5_textdata.Rdata")
+load("doc/results-text/Figure2_textdata.Rdata")
+load("doc/results-text/Figure3_textdata.Rdata")
+load("doc/results-text/Figure4_textdata.Rdata")
+load("doc/results-text/Figure5_textdata.Rdata")
 
 ##Figure 2
 
@@ -127,85 +126,85 @@ awr_wholePlot_2023 <- fig3_textresults %>% filter(Census == "4" & Group == "3" &
 
 #Averages
 
-avg_fram <- allmort_awm %>% filter(sp == "fram") %>% pull(abgmort) %>% 
-  mean(na.rm = TRUE)
-avg_quve <- allmort_awm %>% filter(sp == "quve") %>% pull(abgmort) %>% 
-  mean(na.rm = TRUE)
-avg_quru <- allmort_awm %>% filter(sp == "quru") %>% pull(abgmort) %>%
-  mean(na.rm = TRUE)
-avg_qupr <- allmort_awm %>% filter(sp == "qupr") %>% pull(abgmort) %>% 
-  mean(na.rm = TRUE)
-avg_qual <- allmort_awm %>% filter(sp == "qual") %>% pull(abgmort) %>% 
-  mean(na.rm = TRUE)
-avg_litu <- allmort_awm %>% filter(sp == "litu") %>% pull(abgmort) %>% 
-  mean(na.rm = TRUE)
-avg_other <- allmort_awm %>%  mutate(plot_species = if_else(sp %in% top_sp$sp,sp, "Other")) %>% filter(plot_species == "Other") %>% pull(abgmort) %>% mean(na.rm = TRUE)
+avg_fram <- figure4_textresults %>% filter(plotspecies == "Fraxinus americana") %>% 
+  pull(mort_woody) %>% sum()/11
+avg_quve <- figure4_textresults %>% filter(plotspecies == "Quercus velutina") %>% 
+  pull(mort_woody) %>% sum()/11
+avg_quru <- figure4_textresults %>% filter(plotspecies == "Quercus rubra") %>% 
[…]

### 2024-03-27 doc/15yrsChange.Rmd
assimilating edits from co-authors

--- 2024-03-21 doc/15yrsChange.Rmd
@@
 ```{r eval = TRUE, echo = FALSE, warning = FALSE, results = 'hide'}
-library(tidyverse)
-### Numbers for text
-##load in text data
-load("results-text/Figure2_textdata.Rdata")
-load("results-text/Figure3_textdata.Rdata")
-load("results-text/Figure4_textdata.Rdata")
-load("results-text/Figure5_textdata.Rdata")
-
-##Figure 2
-
-#whole plot
-wp_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-wp_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha) %>% round(digits = 1)
-
-#high deer, high canopy vulnerability
-hdhn_ABG_2013 <- figure2_agb  %>% filter(Year == "2013" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-hdhn_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-hdhn_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-
-#low canopy vulnerability
-ldlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-ldlv_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "1")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-hdlv_ABG_2018 <-  figure2_agb  %>% filter(Year == "2018" & Group == "2")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1) 
-hdlv_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "2")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-
+source("results-text-files.R")
 ```
@@
 
-White-tailed deer (*Odocoileus virginianus*) are a nuisance species prevalent in eastern deciduous forests.*Start with a stronger sentence about the impact of deer on herbivory - Ex.
-White-tailed deer are the dominant herbivore in eastern deciduous forests (Reference)* Their populations have dramatically increased in the past 50 years (McShea et al.
-1997).
-*Find a reference with a specific number of increase in deer populations* due in large part to global change drivers.
-(Found in McGarvey et al.
-2013 – Brown et al.
+Another category of nuisance species is mammals whose population densities have been inflated due to human activities (Murphey et al.
+2013, Clout & Russel), including white-tailed deer (*Odocoileus virginianus*) in U.S.
+eastern deciduous forests.
+Their populations have dramatically increased in the past 50 years (McShea et al.
+1997), due in large part to global change (McGarvey et al.
+2013, Brown et al.
 2000, Côté et al.
 2004, Rooney 2001).
-White-tailed deer preferentially browse on woody plant species in their earliest life stages (McGarvey et al.
-2013), with overabundant herbivory reduce seedling and sapling survival, growth and density (Found in McGarvey et al.
-2013 – Dzieciolowski 1980, Gill and Beardall 2001, Healy 1997, Konig 1976, Putman et al.
-1989).
-Chronic overabundance has been shown to reduce understory diversity and decrease the abundance of traditionally dominant species (*NEW CITATION*).
-Due to climate change and human activity, many non-endemic plant species are being introduced or number in forest ecosystems (Reference).
-Deer find many of these species, including pawpaw (Asimina triloba), to be unpalatable (Found in McGarvey 2013 – Asnani 2006), and do not consume them at the same rates relative to their native counterparts (Reference), enabling them to form dense stands in forest understories (Found in Knauer et al.
-2023 – Horsley and Marquis 1983; Stromayer and Warren 1997; Royo and Carson 2006).
-This lack of browsing pressure, coupled with characteristics such as fast growth rate and greater adaptability to altered conditions precipitated by climate change, allow nuisance plant species to outcompete other species in the understory (Reference).
-The interaction of these nuisance species on the landscape contributes to a recruitment failure in forests, [*Where individuals from certain tree species do not enter the mature population - Do we need to define recruitment failure, or is that considered common knowledge?
-If so, reference*).
-The extent of the consequences derived from recruitment failure on species composition and forest structure is often not apparent for years (McGarvey et al.
-2013)
+White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density (McGarvey et al.
+2013).
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species (Bugalho et al.
+2013) while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS).
+The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory (McGarvey et al.
+2013; Russell et al., 2017; Miller & McGill, 2019; Gorchov et al., 2021; Miller et al., 2023).
 
-The capacity to regenerate following disturbances, including ongoing gap formation through mortality of canopy trees, critically influences long-term forest dynamics.
-Regeneration depends first upon seed production and then upon seedling recruitment, survival, and growth into trees.
-When any one of these steps fails, the stage is set for disturbance to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to it's pre-disturbance state in the foreseeable future [@refs].
-Global change pressures can set the stage for such critical transitions by gradually shifting baseline conditions, making post-disturbance recovery unlikely despite the persistence of mature trees [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @refs].
-In the mid-Atlantic region of eastern North America, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of current canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019;@miller_overabundant_2023].
-Low juvenile abundance in the region is driven by a combination of over-abundant deer, competition with non-endemic species, and possibly climate change [@russell_interactions_2017; @miller_compounding_2019;@miller_overabundant_2023;@gorchov_differential_2021].
-Deer are the biggest problem [@gorchov_differential_2021].
-The juveniles that are present tend to represent a different, more mesophytic set of species (*Acer spp.*, *Fagus grandifolia*) than currently dominate much of the region (*Quercus spp.*, *Carya spp.*, @miller_compounding_2019; @nowacki_climate_2015]--a dynamic driven by fire suppression and mesophication [@refs].
+In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.
+When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future  (Anderson-Teixeira et al., 2013; McDowell et al., 2020, CITATION).
+In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die (Miller & McGill, 2019; Miller et al., 2023).
+The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., Miller & McGill (2019); Nowacki & Abrams (2015)], but even these are often not present in sufficient abundance to replace the current canopy (??).
 
-Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest understory and canopy over extended time periods.
-Here, we use 15 years of detailed census data from a large forest dynamics plot including a 4-ha deer exclosure to test the following hypotheses: (1) Aboveground C stocks in living biomass have declined since the initial census (-∆AGB) (2) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species (3) Recruitment and biomass gains from tree growth have not kept pace with tree mortality, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
+Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest canopy and understory over an extended time.
+Here, we use 15 years of detailed census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.
+We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
 
@@
 
-The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
-The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
-Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
+The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
+The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
+As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
 The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al.
 2013).
-Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).
 The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
-It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the world.
+It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world (Anderson-Teixeira et al.
+2015, Davies et al.
+2021).
 
-In addition to our consideration of the plot as a whole, we give special consideration to three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams):
-(1) Low deer, low canopy vulnerability: A 4-ha portion of upland forest in the SE quarter of the plot, fenced in 1990 and maintained with only occasional deer presence for the past 24 years that has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm *et al.* 2013);
-(2) high deer, low canopy vulnerability: 16.3 ha of upland forest outside of the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens; 
-(3) high deer, high canopy vulnerability: 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens. 
-Upland forest was defined according to the topographic wetness index, originally calculated for the SCBI plot in @mcgregor_tree_2021.
-Vulnerability to non-indigenous insects and pathogens was defined based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment.
-The following canopy species were classified as affected: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Carya dentata*).
+In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig.
+1).
+The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
+1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
+This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm et al.
+2013).
+The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
+1b).
+The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by McGregor et al.
+(2021).
+We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
+The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) (Anderson-Teixeira et al.
+2021).
 
 ![**_Figure `r map_and_photos`.
-Map and photos of the Forest Global Earth Observatory (ForestGEO) plot at the Smithsonian Conservation Biology Institute (SCBI) in Virginia, USA._** Our analysis focuses on three habitat types: (a) low deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, (b) high deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, and (c) high deer density, high abundance of canopy trees vulnerable to non-indigenous insects & pathogens.
+Map and photos of the Forest Global Earth Observatory (ForestGEO) plot at the Smithsonian's Conservation Biology Institute (SCBI) in Virginia, USA._** Our analysis focuses on three portions of the plot: (a) low deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, (b) high deer density, low abundance of canopy trees vulnerable to non-indigenous insects & pathogens, and (c) high deer density, high abundance of canopy trees vulnerable to non-indigenous insects & pathogens.
 *All photos taken September 2023 by K.
@@
 This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh).
-The census records information regarding the dbh, species, living status and location of each stem included in the survey.
+The census records the dbh, species, living status, and spatial location of each censused stem.
 Each individual is marked with an identifying number for sequential data collection in subsequent censuses.
 In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag.
-The plot was established in 2008 and recensused three times since, comprising 15 years of forest dynamics data.
-In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated by GitHub actions (Kim et al., 2022).
-Each individual's location was recorded in the Field Maps application using a digitized map of the plot (*citation*).
-Additional checks and fixes were performed at the conclusion of the last census.
-In addition to the ForestGEO census, since 2014, a tree mortality survey has been conducted annually at the site.
-Initiated, in part, as a response to the presence of emerald ash borer (*Agrilus planipennis*), the mortality survey is conducted on all trees that had a dbh $\geq$ 10 centimeters in the previous census.
-Through assessing individual tree health, we document trends in forest mortality and identify factors that are associated with death.
-Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.
+The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data.
+In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions (Kim et al., 2022).
+Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation).
+In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census (Gonzalez-Akre et al.
+2016).
+Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.
 
@@
 
-We estimated total aboveground biomass (AGB) from the measured dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024).
-*allodb* creates species-specific allometric equations through weighting of all applicable allometries within its database filtered by tree size, geography, and taxa information.
-Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot regions and the plot as a whole.
-We evaluated trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region.
+We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024).
+The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information.
+Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot.
+We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region.
 We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.
@@
 Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species.
-For each .4 hectare quadrat, we calculated the net change in AGB from 2008 to 2023 and AGB stock of canopy species in the understory in 2023.
+We calculated the net change in AGB within each 0.4 hectare quadrat  between 2008 and 2023 and the AGB of canopy species in the understory in 2023.
 
@@
 
-For the first time since the establishment of  the SCBI ForestGEO site, whole-plot aboveground C stocks decreased (Figure 2).
-From 2018 to 2023, whole-plot C stocks have decreased from `r wp_ABG_2018` to `r wp_ABG_2023` Mg C ha^-1^.
-In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses, from `r hdhn_ABG_2013` Mg C ha^-1^ in 2013, to `r hdhn_ABG_2018` Mg C ha^-1^ in 2018, and `r hdhn_ABG_2023` Mg C ha^-1^ in 2023.
-Areas with low density of vulnerable canopy species experienced minor decreases in AGB; C stocks inside and outside of the deer exclosure decreased from **"X to X"**  and **"X to X"** Mg C ha^-1^, respectively.
+For the first time since the establishment of  the SCBI ForestGEO plot, whole-plot aboveground C stocks decreased over the 2018-2023 census interval (Figure 2) .
+From 2018 to 2023, whole-plot C stocks have decreased from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Prior to this census interval, biomass
+
+In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses, from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013, to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018, and `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
+Areas with low density of vulnerable canopy species experienced small losses of AGB between 2018 and 2023; inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
+
 C stock decrease **is related to/ is explained by** above ground net biomass change.

### 2024-03-27 doc/results-text-files.R
tweaks to results text file

@@ -9,13 +9,13 @@ load("doc/results-text/Figure5_textdata.Rdata")
 ##Figure 2
 
 #whole plot
-wp_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-wp_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha) %>% round(digits = 1)
+wp_AGB_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+wp_AGB_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "Whole Plot")  %>%  pull(Abg_C_Mg_Ha) %>% round(digits = 1)
 
 #high deer, high canopy vulnerability
-hdhn_ABG_2013 <- figure2_agb  %>% filter(Year == "2013" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-hdhn_ABG_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
-hdhn_ABG_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_AGB_2013 <- figure2_agb  %>% filter(Year == "2013" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_AGB_2018 <- figure2_agb  %>% filter(Year == "2018" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
+hdhn_AGB_2023 <- figure2_agb  %>% filter(Year == "2023" & Group == "3")  %>%  pull(Abg_C_Mg_Ha)  %>% round(digits = 1)
 
 #high deer, low canopy vulnerability
 hdlv_AGB_2013 <- figure2_agb %>% filter(Year == "2013" & Group == "2") %>% pull(Abg_C_Mg_Ha) %>%
[…]

### 2024-03-27 doc/15yrsChange.Rmd
small edits to results text script & co-authors

@@
 ---
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
 
@@
 Cameron Dow^1,3^
+
+Erika Gonzalez-Akre^1^
 
@@
 
-<!--JOURNAL
-
-Target journal:
-Instructions for authors:
-article type:  -->
-
 ```{r eval = TRUE, echo=FALSE, warning=FALSE, results='hide'}
@@
 
-```{r eval = TRUE, echo = FALSE, warning = FALSE, results = 'hide'}
+```{r eval = TRUE, echo = FALSE, warning = FALSE, results = 'hide', message = FALSE}
 source("results-text-files.R")
@@
 For the first time since the establishment of  the SCBI ForestGEO plot, whole-plot aboveground C stocks decreased over the 2018-2023 census interval (Figure 2) .
-From 2018 to 2023, whole-plot C stocks have decreased from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
-Prior to this census interval, biomass
-
-In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses, from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013, to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018, and `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
+From 2018 to 2023,ABG decreased from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Prior to this, ABG had steadily increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
+In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses: from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013, to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018, and `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
 Areas with low density of vulnerable canopy species experienced small losses of AGB between 2018 and 2023; inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
-
-C stock decrease **is related to/ is explained by** above ground net biomass change.
-In the first two census intervals, 2008 to 2013 and 2013 to 2018, there were positive C fluxes, but for the 2018 to 2023 interval there was a negative C flux of **VALUE** Mg C/ha/year (Figure 3a).
 

### 2024-03-27 doc/results-text-files.R

@@ -1,10 +1,10 @@
 library(tidyverse)
 ### Numbers for text
 ##load in text data
-load("doc/results-text/Figure2_textdata.Rdata")
-load("doc/results-text/Figure3_textdata.Rdata")
-load("doc/results-text/Figure4_textdata.Rdata")
-load("doc/results-text/Figure5_textdata.Rdata")
+load("results-text/Figure2_textdata.Rdata")
+load("results-text/Figure3_textdata.Rdata")
+load("results-text/Figure4_textdata.Rdata")
+load("results-text/Figure5_textdata.Rdata")
 
 ##Figure 2
 
@@ -126,7 +126,6 @@ awr_wholePlot_2023 <- fig3_textresults %>% filter(Census == "4" & Group == "3" &
 
 #Averages
 
-<<<<<<< Updated upstream
 avg_fram <- figure4_textresults %>% filter(plotspecies == "Fraxinus americana") %>% 
   pull(mort_woody) %>% sum()/11
 avg_quve <- figure4_textresults %>% filter(plotspecies == "Quercus velutina") %>% 
@@ -173,8 +172,6 @@ qual_peak <- figure4_textresults %>% filter(plotspecies == "Quercus alba") %>% p
 litu_peak <- figure4_textresults %>% filter(plotspecies == "Liriodendron tulipifera") %>% pull(mort_woody) %>% max()
 
 other_peak <- figure4_textresults %>% filter(plotspecies == "Other") %>% pull(mort_woody) %>% max()
-=======
->>>>>>> Stashed changes
 
 ##Figure 5
 

### 2024-03-28 doc/15yrsChange.Rmd
updating rmarkdown citations

@@
 Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
-One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@refs].
+One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; nunez_population_2010].
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
@@
 
-Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes (Lovett et al.
-2016, Fischer et al.
-2013, Herms and McCullough 2014, Fei et al.
-2014).
-Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement (Fischer et al.
-2013, Aukema et al.
-2010, Brockerhoff et al.
-2006).
-Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species (Aukema et.
-al 2010,Anderson-Teixeria 2021, Klooster et al.
-2013, Ellison et al.
-2018).
-Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species, further increasing potentials for biomass loss and altered carbon cycle dynamics (Fei et al.
-2013).
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @fischer_disturbances_2013; @herms_emerald_2014; @fei_biogeomorphic_2014].
+Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006].
+Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021a; @klooster_ash_2014; @ellison_present_2018].
+Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.
 
-Another category of nuisance species is mammals whose population densities have been inflated due to human activities (Murphey et al.
-2013, Clout & Russel), including white-tailed deer (*Odocoileus virginianus*) in U.S.
+Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S.
 eastern deciduous forests.
@@
 2004, Rooney 2001).
-White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density (McGarvey et al.
-2013).
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species (Bugalho et al.
-2013) while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS).
-The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory (McGarvey et al.
-2013; Russell et al., 2017; Miller & McGill, 2019; Gorchov et al., 2021; Miller et al., 2023).
+White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a].
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS).
+The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
 
 In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.
-When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future  (Anderson-Teixeira et al., 2013; McDowell et al., 2020, CITATION).
-In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die (Miller & McGill, 2019; Miller et al., 2023).
-The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., Miller & McGill (2019); Nowacki & Abrams (2015)], but even these are often not present in sufficient abundance to replace the current canopy (??).
+When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
+In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
+The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp.,[@miller_compounding_2019; @nowacki_climate_2015]), but even these are often not present in sufficient abundance to replace the current canopy (??).
 
@@
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al.
-2013).
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [SI archives, @bourg_initial_2013.
 The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
-It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world (Anderson-Teixeira et al.
-2015, Davies et al.
-2021).
+It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
 
@@
 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
-This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens (Holm et al.
-2013).
+This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
 The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
@@
 The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
-We defined upland forest according to topographic wetness index, which was calculated for the forest plot by McGregor et al.
-(2021).
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021.
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
-The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) (Anderson-Teixeira et al.
-2021).
+The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021a]
 

### 2024-03-29 doc/15yrsChange.Rmd
updating rmarkdown citations

@@
   #officedown::rdocx_document
-bibliography: references.bib
+bibliography: citations.bib
 csl: new-phytologist.csl
@@
 Although currently a net C sink, the future of the temperate forest biome remains uncertain.
-The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlström_robustness_2012].
-Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlström_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
-One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; nunez_population_2010].
+The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012].
+Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
+One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
@@
 
-Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @fischer_disturbances_2013; @herms_emerald_2014; @fei_biogeomorphic_2014].
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, and successional trajectories.
+[@lovett_nonnative_2016; @herms_emerald_2014].
 Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006].
-Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021a; @klooster_ash_2014; @ellison_present_2018].
+Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018].
 Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.
@@
 eastern deciduous forests.
-Their populations have dramatically increased in the past 50 years (McShea et al.
-1997), due in large part to global change (McGarvey et al.
-2013, Brown et al.
-2000, Côté et al.
-2004, Rooney 2001).
+Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [cote_ecological_2004; @rooney_deer_2001; @brown_future_2000].
 White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a].
@@
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
-The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021a]
+The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021]
 
@@
 
-As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in Condit (1998).
+As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998.
 This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh).
@@
 The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data.
-In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions (Kim et al., 2022).
+In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022].
 Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation).
-In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census (Gonzalez-Akre et al.
-2016).
+In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016].
 Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.
@@
 
-We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* (Gonzalez-Akre, 2024).
+We estimated total aboveground biomass (AGB) from the dbh of each stem in each census using extra-tropical allometric equations from the R package *allodb* [@gonzalez-akre_allodb_2022].
 The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information.

### 2024-04-03 doc/15yrsChange.Rmd
Results and discussion, misc. edits

@@
 
-Erika Gonzalez-Akre^1^
+Erika Gonzalez-Akre^1^  ( Orcid ID : 0000-0001-8305-6672)
 
@@
 
-Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, and successional trajectories.
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes.
 [@lovett_nonnative_2016; @herms_emerald_2014].
@@
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
-The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021]
+The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021].
 
@@
 We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region.
-We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data.
+We further assessed the role of nuisance species on mortality by calculating AWM flux for each year and each species using the annual mortality survey data (excluding stems < 10 cm dbh).
 
@@
 Finally, we assessed the spatial overlap of biomass loss and recruitment of canopy species.
-We calculated the net change in AGB within each 0.4 hectare quadrat  between 2008 and 2023 and the AGB of canopy species in the understory in 2023.
+We calculated the ∆AGB of canopy trees within each 0.4 hectare quadrat between 2008 and 2023 and the AGB of canopy species in the understory in 2023.
 
@@
 
-For the first time since the establishment of  the SCBI ForestGEO plot, whole-plot aboveground C stocks decreased over the 2018-2023 census interval (Figure 2) .
-From 2018 to 2023,ABG decreased from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
-Prior to this, ABG had steadily increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
-In regions with  high deer density and high canopy vulnerability AGB decreased significantly over the past two censuses: from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013, to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018, and `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
-Areas with low density of vulnerable canopy species experienced small losses of AGB between 2018 and 2023; inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
+For the first time since the establishment of  the SCBI ForestGEO plot, living aboveground C stocks decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
+In regions characterized by high deer density and high canopy vulnerability, AGB decreased significantly over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
+Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
 
 ![**_Figure `r biomass_trends`.
-Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)
+Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)
 
-The trend in net woody production can be explained by its constituent fluxes (Figure 3).
-Trends in woody mortality show that for the first two census intervals, the whole plot has lost **VALUE** and then **VALUE** Mg C/ha/year, increasing to a loss of **VALUE** Mg C/ha/year for the most recent census interval (Figure 3b).
-Woody productivity contributes positively to the flux, showing that the whole plot gained **VALUE** Mg C/ha/year in the first interval, **VALUE** Mg C/ha/year in the second, and **VALUE** Mg C/ha/year in the third (Figure 3c).
-Areas of low deer, low nuisance species were consistently the most productive ending at **VALUE** Mg C/ha/year in the interval 2018 to 2023, and areas of high deer, high nuisance species were the least productive at **VALUE** Mg C/ha/year in the interval 2018 to 2023.
-Recruitment has consistently added smaller magnitude amounts of carbon with the whole plot gaining **VALUE** Mg C/ha/year (Figure 3d).
-Recruitment was highest at **VALUE** Mg C/ha/year in areas with high deer and high vulnerable species, and lowest at **VALUE** Mg C/ha/year in areas with high deer and low vulnerable species.
+ The transition from net AGB accumulation to loss is primarily attributable to increased AWM across all plot regions (Figure 3b).
+Initially, AWM flux remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals.
+However, in the most recent period, AWM increased significantly to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
+Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval.
+Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.
+Simultaneously, the rate of biomass accrual from woody growth decreased across the plot (Figure 3c), declining from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over the 2008-2013 period to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
+Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the region with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
+Furthermore, AWG declined across all census intervals, even when AWM remained stable.
+Recruitment, constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d).
+AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.
 
 ![**_Figure `r loss_gain_net`.
-Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)
+Changes in aboveground woody biomass at the SCBI ForestGEO plot over each census interval since 2008 establishment, including (a) net change, (b) losses to mortality, and gains from (c) woody productivity and (d) recruitment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure3.jpeg)
 
@@
 
-Woody mortality was the primary driver of the negative trend in biomass, and certain species were responsible for higher levels of mortality.
-Fraxinus americana had the highest aboveground woody mortality rates of any species present on the plot, peaking at **VALUE** Mg C/ha/year lost in *2020* , and **VALUE** Mg C/ha/year lost in 2022 (Figure 4).
-Quercus velutina contributed the second-highest levels of mortality, losing **VALUE** Mg C/ha/year in 2022.
-Quercus rubra, Carya species, Quercus prinus, Quercus alba, and Liriodendron tulipifera were all significant contributors as well.
+Over the 15-year study period, AWM increased throughout the plot yet AWM rates remained stable in a majority of individual species.
+The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4).
+*Fraxinus americana* had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
+*Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^.
+The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^).
+Moreover, both species exhitibed significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.
 
 ![**_Figure `r mortality_by_genera`.
-Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the seven canopy taxa with highest average mortality rate in descending order and a category including all others.
-The census included all stems ≥ 10cm in the most recent full census.](display/Figure4.jpeg)
+Annual woody mortality by species at the SCBI ForestGEO plot from 2008-2023._** Shown are the six canopy taxa with highest average mortality rate in descending order and a category including all others.
+The census included all stems ≥ 10 cm in the most recent full census.](display/Figure4.jpeg)
 
-## Species composition of recruitment flux
-Certain species dominated aboveground woody recruitment, with understory species being recruited at higher rates than canopy species (Figure 5).
-Out of the canopy species, Carya species showed the highest levels of recruitment with a rate of **VALUE** Mg C/ha/year recruited from 2018 to 2023 (Figure 5a).
-Acer negundo had the second highest rate of recruitment for canopy species at **VALUE** Mg C/ha/year from 2018 to 2023.
-Fagus grandifolia, Fraxinus americana, and Ulmus rubra are the other species with the highest recruitment rates.
-Most canopy species had lower recruitment rates in the 2018 to 2023 census interval than the 2008 to 2013 interval.
-Lindera benzoin had by far the highest recruitment rate of the understory species at **VALUE** Mg C/ha/year from 2018 to 2023 (Figure 3b).
-Asimina triloba had the second highest rate of recruitment of the understory species at **VALUE** Mg C/ha/year recruited from 2018 to 2023.
-Other understory species with high recruitment were Carpinus caroliniana, Hamamelis virginiana, and Celtis occidentalis.
+## Species composition of recruitment
+
+We report strong mediating effects of deer density on the quantity and composition of woody stem recruitment (Figure 5).
+The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^.
+In particular, the recruitment rate of *Carya* spp.
+and *Quercus* spp., late-successional taxa characteristic of this forest's expected climax state, was reduced by > 50% in areas with high deer density.
+The prevalence of vulnerable canopy species in regions with low deer density showed no significant impact on the total quantity of canopy species recruitment into the understory.
+However, recruitment of *Fraxinus americana* and *Fagus grandifolia* was higher in areas with high canopy vulnerability, while recruitment of *Ulmus rubra* and *Carya* spp.
+was lower.
+Across all regions, the recruitment rate of understory taxa was nearly 10 times that of canopy taxa.
+Recruitment of *Lindera benzoin* was high across all regions (> 100 stems ha^-1^ yr^-1^), and significantly increased in areas with high deer densities while peaking in areas that also had low canopy vulnerability.
+*Asimina triloba* recruitment was also lowest within the deer exclosure (`r round(group1_astrRecruit,1)` stems ha^-1^ yr^-1^) but was highest in the region with high deer density and high canopy vulnerability (`r round(group3_astrRecruit,1)` stems ha^-1^ yr^-1^).
 
 ![**_Figure `r recruitment_by_genera`.
-Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across habitats differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment >20x that of canopy species recruitment.](display/Figure5.jpeg)
+Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20x that of canopy species recruitment.](display/Figure5.jpeg)
 
-## Regeneration & mortality mismatch
+## Overlapping biomass loss and recruitment
 
-![**_Figure `r regen_failure`._**](display/Figure6.jpeg)
+Our analyses reveal significant areas of decreasing AGB and suppressed regeneration of canopy species.
+Specifically, we found that 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory.
+Only 21.7% of the plot exhibits an understory with high AGB of species capable of reaching the canopy.
+Within the region characterized by high deer density and high canopy vulnerability 84% of the area has lost canopy biomass, only 5.4% has an understory containing high AGB of canopy taxa, and 0% of area exhibits both.
+In contrast, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB.
+Furthermore, 70.7% of this same region has high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration.
+
+![**_Figure `r regen_failure`.
+Characterizing the SCBI ForestGEO plot by changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023.
+(b) The % area of each region that falls within each classification.](display/Figure6.jpeg)
 
@@
 
-We found that the combination of decreasing biomass (driven by increasing woody mortality) and limited recruitment of canopy species in the understory is compromising the potential the carbon sink, now and into the future.
-Differences in AGB were the biggest in the regions of the plot with high vulnerability to nuisance species.
-These regions also had the lowest amount of canopy species in the understory.
-Areas with low deer population had the highest amount of canopy species in the understory.
-Some of the plot is increasing biomass, notably those with low proportions of vulnerable species. On net, these increases are not outpacing mortality. 
-Suggests that the combination of heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to pathogen-derived mortality where understory species that are resistant to deer and fast growers can take advantage, further competing with canopy species.
-While Emerald Ash-borer has had a large impact on biomass loss at our site, mortality among oak species, which is not linked to a non-indigenous pest, is substantial as well.
-Oak species in the region are generally found in older, even aged cohorts, which contributes to accelerated mortality rates (Oak et al., 2016).
+Our investigation supports our hypotheses and demonstrates that the combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future.
+Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens.
+These regions also exhibited the lowest recruitment of canopy species in the understory.
+The region with low deer density had the highest amount of canopy species in the understory.
+Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG.
+Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
+
+We find that while nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
+Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016).
 However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
@@
 
-Using current understory as a predictor of the future canopy is imperfect. 
-Some canopy species (examples) recruit in pulse events - if there is a shift in deer population and enough survival that coincides with a mast year than you could get regeneration into the canopy.
-However, current recruitment levels are below a rate at which we would expect in maturing secondary forests and may be insufficient to prevent large periods without a close canopy.
-Our data is from just the single site, which can make generalizability challenging. However, there is a lot of support that the drivers we identify here are representative of the region.
-Mixed hardwood forests are typical of mid-Atlantic forests and deer populations are high throughout the region.
-Vulnerable species (ash) are also common throughout these forests, and emerald ash borer continues to spread. Where it does spread it consistently decimates adult ash.
-Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forests. 
-Nuisance species (driven by human pressures/actions) interact with other anthropogenic changes to modify forest ecosystems. 
-Have the capacity to alter short-term dynamics: large carbon losses that would not be expected in models that don’t exclude it.
-Longer-term have capacity to alter successional dynamics: can delay recovery from disturbances (including other disturbances) or even suppress regeneration into a closed-canopy for sustained periods. 
-Decrease resilience to other pressures as pests and pathogens decrease diversity through shrinking the number of possible canopy species.
-Positive takeaways: this work highlights the potential of active management. Preserving the forests that we have is more valuable than planting new forests. 
-Pests and pathogens are driving mortality - may be opportunities for treatment/prevention of spread. 
-Research into natural immunity (ash species in original range of EAB have immunity).
-Management of deer populations to decrease browsing pressure - deer exclosure shows the importance to maintain an understory.
+Relying on the current understory as a proxy for future canopy presents challenges.
+White-tailed deer populations have varied greatly over the last century (citation).
+Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
+A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse, could allow for substantial, rapid regeneration of canopy species.
+However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
+When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.
+
+Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
+Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
+The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
+Furthermore,the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
+EAB consistently induces near-total mortality in *Fraxinus* spp.
+once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
+
+Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
+Nuisance species, driven by human introduction and perturbations to ecological equilibriums, interact with other anthropogenic changes to modify forest ecosystems.
+By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.
+Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics.
+They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods.
+They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.
+
+Nuisance species pose a serious threat to the health of temperate forests.
+But, this work highlights the potential of active management to protect them.
+Increased understory recruitment within the deer exclosure demonstrates that lowering deer populations can reduce browsing pressure.
+The direct linkage between non-indigenous pests and pathogens and decreased carbon storage demonstrates the need for natural climate solutions that focus on preserving the forests that we have instead of planting new ones.
+Opportunities exist to expand efforts that prevent further spread, increase the application of extant treatments, and continue research into natural immunity.
 

### 2024-04-03 code/figure4_annual_mort_by_species.r

@@ -20,10 +20,10 @@ mort_plot_df <- allmort_awm  %>%
   right_join(allspyr)  %>% 
   mutate(abgmort = if_else(is.na(abgmort),0, abgmort))  %>% 
   mutate(sp = if_else(sp %in% c("caco","cagl","caovl","cato"), "Hickory spp.",sp))  %>% 
-  mutate(plotspecies = if_else(sp %in% top_sp$sp,sp, "Other"))  %>% 
+  mutate(plotspecies = if_else(sp %in% top_sp$sp,sp, "All other taxa"))  %>% 
   group_by(survey_year, plotspecies)  %>% 
   summarize(mort_woody = sum(abgmort))  %>% 
-  mutate(plotspecies = factor(plotspecies, levels = c(top_sp$sp, "Other")))
+  mutate(plotspecies = factor(plotspecies, levels = c(top_sp$sp, "All other taxa")))
 
 mort_plot_df$plotspecies <- gsub('fram', 'Fraxinus americana', mort_plot_df$plotspecies)
 mort_plot_df$plotspecies <- gsub('Hickory spp.', 'Carya spp.', mort_plot_df$plotspecies)
@@ -34,7 +34,7 @@ mort_plot_df$plotspecies <- gsub('quru', 'Quercus rubra', mort_plot_df$plotspeci
 mort_plot_df$plotspecies <- gsub('quve', 'Quercus velutina', mort_plot_df$plotspecies)
 
 facet_order <- c("Fraxinus americana", "Quercus velutina", "Quercus rubra", 
-                 "Quercus prinus", "Quercus alba", "Liriodendron tulipifera", "Other")
+                 "Quercus prinus", "Quercus alba", "Liriodendron tulipifera", "All other taxa")
 
 facet_years <- c(2013, 2023)
 
@@ -62,7 +62,7 @@ barp <- ggplot(plotdf, aes(x = survey_year, y = mort_woody, fill = as.ordered(su
         axis.title.x = element_blank(),
         panel.spacing.x = unit(0, "lines"),
         #strip.background = element_blank(),
-        strip.text = element_markdown(size = 9.5), 
+        strip.text = element_markdown(size = 10.5), 
         panel.grid.major.x = element_blank(),
         panel.grid.minor.x = element_blank(),
         legend.position = "none") 
@@ -70,7 +70,7 @@ barp <- ggplot(plotdf, aes(x = survey_year, y = mort_woody, fill = as.ordered(su
 
 barp
[…]

### 2024-04-03 doc/results-text-files.R

@@ -106,16 +106,18 @@ awr_group1_2013 <- fig3_textresults %>% filter(Census == "2" & Group == "1" & Fl
 awr_group1_2018 <- fig3_textresults %>% filter(Census == "3" & Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 awr_group1_2023 <- fig3_textresults %>% filter(Census == "4" & Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 
+awr_group1_avg <- fig3_textresults %>% filter(Group == "1" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% mean()  %>% round(digits = 3)
+
 #Group 2
 awr_group2_2013 <- fig3_textresults %>% filter(Census == "2" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 awr_group2_2018 <- fig3_textresults %>% filter(Census == "3" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 awr_group2_2023 <- fig3_textresults %>% filter(Census == "4" & Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
-
+awr_group2_avg <- fig3_textresults %>% filter(Group == "2" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% mean()  %>% round(digits = 3)
 #Group 3
 awr_group3_2013 <- fig3_textresults %>% filter(Census == "2" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 awr_group3_2018 <- fig3_textresults %>% filter(Census == "3" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
 awr_group3_2023 <- fig3_textresults %>% filter(Census == "4" & Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% round(digits = 1)
-
+awr_group3_avg <- fig3_textresults %>% filter(Group == "3" & Flux == "AWR") %>% pull(MgC_Yr_Ha) %>% mean()  %>% round(digits = 3)
 #Whole Plot
[…]

### 2024-04-04 doc/15yrsChange.Rmd
figure 1 changes and citation updates

--- 2024-03-29 doc/15yrsChange.Rmd
@@
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [SI archives, @bourg_initial_2013.
-The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [SI archives, @bourg_initial_2013, @mcgarvey_effects_2013a] The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
 It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].

### 2024-04-04 code/figure3_woodyfluxes.R

@@ -40,7 +40,7 @@ fig3_woodyfluxes <- group_trends  %>%
                               Census == 4 ~ "2018-\n2023")
   )
 
-
+ 
 colz <- c("#E7BC40","#C7622B", "#750000","#7e937f")
 flux_names <- as_labeller(c("AWM" = "Aboveground Woody Mortality",
                             "AWP" = "Aboveground Woody Growth",
@@ -67,6 +67,7 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
         axis.text.x = element_text(size = 18, margin = margin(t = .15,r = 0, b = .05,l = 0,unit = "in"), angle = 45, hjust = 1),
         axis.text.y = element_text(size = 18, margin = margin(t = 0,r = .08, b = 0,l = 0, unit = "in")),
         axis.title = element_text(size = 20)) 
+
 
 fig3
 

### 2024-04-04 code/identify-plot-regions.r

@@ -4,9 +4,9 @@ library(terra)
 library(tidyterra)
 
 
-setwd("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/")
-load("tree_main_census/data/scbi.stem1.rdata")
-load("tree_main_census/data/scbi.spptable.rdata")
+#setwd("C:/Work/Smithsonian/Repos/SCBI-ForestGEO-Data/")
+scbi.stem1 <- load("data/census_data/scbi.stem1.corrected.rdata")
+scbi.spptable <- load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
 
 quadrats <- st_read("spatial_data/shapefiles/20m_grid.shp")
 dem <- rast("spatial_data/elevation/rasters/plot_elevation.tif")

### 2024-04-04 doc/15yrsChange.Rmd
Citation edits & other

@@
 Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
-One mechanism that is not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in undesirable ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
+For instance, one mechanism not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
@@
 
-Another category of nuisance species is mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S.
+Another category of nuisance species includes mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S.
 eastern deciduous forests.
@@
 White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a].
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., XX ) species (REFS).
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018).
 The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
@@
 When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
-In the mid-Atlantic region of the eastern United States, forests face a severe “regeneration debt”, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
-The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp.,[@miller_compounding_2019; @nowacki_climate_2015]), but even these are often not present in sufficient abundance to replace the current canopy (??).
+In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort of those same species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
+The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., [@nowacki_climate_2015], Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
 
 Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest canopy and understory over an extended time.
-Here, we use 15 years of detailed census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.
-We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
+Here, we use 15 years of detailed tree census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.
+We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future tree mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.
 
@@
 
-In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure non-indigenous insects and pathogens and white-tailed deer (Fig.
+In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig.
 1).
@@
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
-The following canopy species were identified as vulnerable: ash (Fraxinus spp.), elm (Ulmus spp.), butternut (Juglans cinerea), and American chestnut (Carya dentata) [@anderson-teixeira_longterm_2021].
+The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].
 
@@
 The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data.
-In the last census, data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022].
+In the last census, systematic data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022].
 Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation).
-In addition to the ForestGEO census, a tree health and mortality survey was conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016].
+In addition to the ForestGEO census, a tree health and mortality survey has been conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016].
 Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.
@@
 
-In addition to the estimation of AWR biomass flux, we analyzed the effects of nuisance species on recruitment through counts of newly recruited stems.
+In addition to the estimation of AWR biomass flux, we analyzed the possible effects of nuisance species on recruitment based on counts of newly recruited stems.
 We categorized each species as either canopy or understory based on expert knowledge of life history traits.
@@
 Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG.
+Total ABG stocks in the plot were lower in 2023 than in 2013.
 Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
@@
 
-# Conclusions /Future Directions
+# Future Directions
 
-Expand analyses across other temperate forests.
-Different forest compositions and herbivory dynamics, but many pests and pathogens.
-Leverage other forest dynamics sites to quantify the impact of nuisance species across large regions.
-Investigate productivity responses of surviving canopy. We would expect survivors adjacent to gaps increase productivity as competition decreases, but may be hampered by climate change.
-Interactions of global change drivers are critical to understand (though difficult to study).
+Understanding interactions between global change drivers, an ongoing area of research, is critical.
+We would expect that trees unaffected by nuisance species (*e.g.* *Liriodendron tulipifera*, *Quercus rubra*) might see productivity increases as canopy gaps form and competition from nearby trees decreases.
+However, continuous decreases in AWG throughout the plot suggests that other mechanisms, including increasing maximum temperatures during the growing season (Dow et al 2022), may be hindering tree growth.
+Tree seedlings of many species, already decimated by deer browsing, are further threatened by rising maximum temperatures (McEwan et al 2011).
+Increasing mortality within *Quercus* species is another focal area of future research, as species within the *Quercus* genera constitute > 33% of woody productivity at this site (McGregor et al 2020).
+
+Future research should build on our results by expanding these analyses to other temperate forest regions.
+While other areas will have different forest compositions and herbivory dynamics, increasing pressure from pests and pathogens is pervasive throughout temperate forests (Santini et al.
+2012, Ayres and Lombardero 2018).
+Moreover, while not a large factor at this site, the spread of non-indigenous plant species can further threaten the  regeneration of woody canopy species (Pile Knapp et al, 2023).
+Large forest dynamics sites (*i.e.* other sites in the ForestGEO network) are well-suited to this research and can be leveraged to quantify how nuisance species compromise present and future forest carbon sequestration.
+Expanding the scale of these analyses may also facilitate the incorporation of nuisance species into global vegetation and carbon models.
+Future evaluations of temperate forest health and carbon storage must account for the substantial effects of nuisance species along with other global changes.
 

### 2024-04-04 code/Woodyflux_Calculations_debugging.r

@@ -1,0 +1,181 @@
+
+
+library(tidyverse)
+library(allodb)
+
+##### 0 - helper functions #####
+create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
+  census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
+  return(census_df)
+}
+# Can skip to III and read-in all_censuses #
+
+##### I - Read in data from ALL censuses and the species table #####
+grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+  mutate(quadrat = sprintf("%04d",quadrat))
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
+load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem4.corrected.rdata")
+
+load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
+
+latlong <- c(-78.1454, 38.8935)
+
+scbi.stem1.corrected <- scbi.stem1.corrected  %>% 
+  mutate(Census = 1)
+scbi.stem2.corrected <- scbi.stem2.corrected  %>% 
+  mutate(Census = 2)
+scbi.stem3.corrected <- scbi.stem3.corrected  %>% 
+  mutate(Census = 3)
+scbi.stem4.corrected <- scbi.stem4.corrected  %>% 
+  mutate(Census = 4, hom = as.character(hom), dbh = as.character(dbh))
+
+##### II - Combine the census data - run this before either section III or IV #####
+all_censuses <- scbi.stem1.corrected  %>% 
+  bind_rows(scbi.stem2.corrected, scbi.stem3.corrected, scbi.stem4.corrected)  %>% 
+  left_join(scbi.spptable, by = c("sp" = "sp"))  %>% 
[…]

### 2024-04-04 doc/results-text-files.R

@@ -189,7 +189,7 @@ group3_canopy <- figure5_textresults %>% filter(Group == "3" & canopy_position =
 group2_3_canopy <- figure5_textresults %>% filter(Group %in% c("2","3") & canopy_position == "canopy") %>% group_by(plot_sp)  %>% summarize(n_stems = mean(n_stems,na.rm = T))  %>%  pull(n_stems) %>% sum(na.rm = TRUE)
 
 #Understory
-group1_understory <- figure5_textresults %>% filter(Group == "1" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
+group1_understory <- figure5_textresults %>% filter(Group == "1" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)  %>% round()
 group2_understory <- figure5_textresults %>% filter(Group == "2" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
 group3_understory <- figure5_textresults %>% filter(Group == "3" & canopy_position == "understory")%>% pull(n_stems) %>% sum(na.rm = TRUE)
 

### 2024-04-04 doc/15yrsChange.Rmd
minor text changes

@@
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [SI archives, @bourg_initial_2013, @mcgarvey_effects_2013a] The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (Smithsonian Institution archives, @bourg_initial_2013, @mcgarvey_effects_2013a).
+The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
 It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that facilitates comparative forest ecology studies around the world [@anderson-teixeira_ctfsforestgeo_2015; @davies_forestgeo_2021].
@@
 1).
-The first portion we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
+The first regoin we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
-This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
+This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
 The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
@@
 The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
-We defined upland forest according to topographic wetness index, which was calculated for the forest plot by mcgregor_tree_2021.
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021.
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
@@
 
-For the first time since the establishment of  the SCBI ForestGEO plot, living aboveground C stocks decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+For the first time since the establishment of  the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
 Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
-In regions characterized by high deer density and high canopy vulnerability, AGB decreased significantly over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
+In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
 Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
@@
 
- The transition from net AGB accumulation to loss is primarily attributable to increased AWM across all plot regions (Figure 3b).
+ The transition from net AGB accumulation to loss is primarily attributable to increased mortality (AWM) across all plot regions (Figure 3b).
 Initially, AWM flux remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals.
-However, in the most recent period, AWM increased significantly to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
+However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
 Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval.
 Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.
-Simultaneously, the rate of biomass accrual from woody growth decreased across the plot (Figure 3c), declining from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over the 2008-2013 period to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
+Simultaneously, the rate of biomass accrual from woody growth (AWG) decreased across the plot (Figure 3c), declining from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over the 2008-2013 period to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
 Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the region with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
@@
 Increasing mortality within *Quercus* species is another focal area of future research, as species within the *Quercus* genera constitute > 33% of woody productivity at this site (McGregor et al 2020).
+Multiple global changes already impact the SCBI ForestGEO plot; carbon storage in the site is, and will be, mediated by multifaceted, interacting pressures resulting from human actions.
 
@@
 <div id="refs"></div>
-
-# Supporting Information

### 2024-04-05 doc/15yrsChange.Rmd
summary, results reordering

@@
 
-**Authors :**
+**Authors:**
 
@@
 
-Rachel Hoffman^1^
+Rachel M. Hoffman^1^
 
-Krystal Bagnaschi^1^
+Krystal J. Bagnaschi^1^
 
-Iris Kennedy^1^
+Iris A. Kennedy^1^
 
@@
 
-*Summary format for New Phytologist must be under 200 words and should be organized using four bullet points to indicate: (1) the research conducted, including the rationale, (2) methods, (3) key results, and (4) the main conclusion, including the key points of discussion*
-
-Temperate deciduous forests play a critical role in the global carbon (C) cycle, accounting for a substantial portion of the global forest C sink.
-The dominant view imbued in Earth System Models is that, at least within the eastern United States, the biome is likely to remain a C sink throughout the 21st century.
-However, these models do not incorporate nuisance species -- *i.e.*, species whose local abundance has increased as a result of human activities and that are causing ecological harm.
-Nuisance species are known to increase tree mortality (*e.g.*, non-endemic pests and pathogens) and decrease recruitment (*e.g.*, deer and non-endemic plants) throughout the temperate deciduous biome, yet the net effect on current and future C cycling is unknown.
-Here, we use 15 years of detailed census data from a large (25.6 ha) forest dynamics plot in a Mid-Atlantic temperate forest, including a 4-ha deer exclosure, to understand how nuisance species are affecting C cycling.
-We show increased biomass mortality, a net reduction in aboveground C storage, and decreased abundance of canopy tree species in the understory.
-Under current trends this forest will continue to lose C sequestration capacity, indicating that the C sequestration of temperate deciduous forests is likely overestimated.
+- The future of the important temperate forest carbon (C) sink is dependent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood.
+Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration.
+- We analyzed biomass change and regenerative capacity in sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
+- Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens.
+There was also severe depletion of canopy tree species in the understory in areas with high deer browsing, endangering future canopy regeneration.
+- Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem.
+This indicates that the C sequestration of U.S.
+Eastern deciduous forests is likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.
 
@@
 *New Phytologist allows 5-8 keywords. Short phrases are acceptable*
+
+Pests and pathogens; temperate forests; carbon sequestration; white-tailed deer; ecosystem processes;
 
@@
 Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
-This is problematic in that nuisance species – including non-indigenous insect pests and pathogens, non-indigenous plants, and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.
+This is problematic in that nuisance species – including non-indigenous insect pests and pathogens and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.
 
-Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes.
-[@lovett_nonnative_2016; @herms_emerald_2014].
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @herms_emerald_2014; @fischer_disturbances_2013] (Fei et al 2014).
 Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006].
@@
 eastern deciduous forests.
-Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [cote_ecological_2004; @rooney_deer_2001; @brown_future_2000].
+Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000].
 White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a].
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (e.g., pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (e.g., garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018).
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018).
 The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
@@
 When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
-In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt, meaning that there are insufficient juveniles of canopy tree species to replace the mature cohort of those same species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
-The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., Acer spp., Fagus grandifolia) than historically dominated much of the region (e.g., Quercus spp., Carya spp., [@nowacki_climate_2015], Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
+In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt; there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
+The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., *Acer* spp., *Fagus grandifolia*) than historically dominated much of the region (e.g., *Quercus* spp., *Carya* spp., [@nowacki_climate_2015]; Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
 
@@
 1).
-The first regoin we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
+The first focal area we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
 1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
 This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
-The second focal area is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
+The second portion is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
 1b).
-The third focal portion of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
+The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
 We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021.
@@
 
-For the first time since the establishment of  the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
-Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
-In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
-Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
-
-![**_Figure `r biomass_trends`.
-Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)
-
- The transition from net AGB accumulation to loss is primarily attributable to increased mortality (AWM) across all plot regions (Figure 3b).
-Initially, AWM flux remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals.
+All sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3).
+The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens.
+Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals.
 However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
@@
 Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.
-Simultaneously, the rate of biomass accrual from woody growth (AWG) decreased across the plot (Figure 3c), declining from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over the 2008-2013 period to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
-Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the region with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
-Furthermore, AWG declined across all census intervals, even when AWM remained stable.
-Recruitment, constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d).
-AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.
 
@@
 
-## Species composition of mortality flux
-
-Over the 15-year study period, AWM increased throughout the plot yet AWM rates remained stable in a majority of individual species.
+Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species.
 The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4).
-*Fraxinus americana* had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
+*Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
 *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^.
@@
 
-## Species composition of recruitment
+For the first time since the establishment of  the SCBI ForestGEO plot, biomass losses to mortality outpaced biomass gains from tree growth (Figure 3a).
+Living aboveground C stocks (AGB) decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
+In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
+Conversely, areas with low density of vulnerable canopy species experienced marginal losses of AGB between 2018 and 2023; both inside and outside of the deer exclosure, biomass decreased from `r hdlv_AGB_2018` to `r hdlv_AGB_2023` Mg C ha^-1^  and `r ldlv_AGB_2018` to `r ldlv_AGB_2023` Mg C ha^-1^, respectively.
 
-We report strong mediating effects of deer density on the quantity and composition of woody stem recruitment (Figure 5).
+![**_Figure `r biomass_trends`.
+Aboveground biomass at the SCBI ForestGEO plot since 2008 establishment._** Values are shown for the plot as a whole and for regions differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens.](display/Figure2.jpeg)
+
+ Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c).
+Plot-wide AWG decreased by `r (awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013 * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
+Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
+Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d).
+AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.
+
+Deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5).
 The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^.
@@
 
-## Overlapping biomass loss and recruitment
-
-Our analyses reveal significant areas of decreasing AGB and suppressed regeneration of canopy species.
-Specifically, we found that 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory.
+There were significant areas of decreasing AGB and suppressed regeneration of canopy species.
+Specifically, 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory.
 Only 21.7% of the plot exhibits an understory with high AGB of species capable of reaching the canopy.
@@
 
-Our investigation supports our hypotheses and demonstrates that the combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future.
+The combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future.
 Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens.
 These regions also exhibited the lowest recruitment of canopy species in the understory.
-The region with low deer density had the highest amount of canopy species in the understory.
+Deer severely depleted regeneration of canopy species and the region with low deer density maintained the highest amount of canopy species in the understory.
 Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG.
@@
 
-We find that while nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
+While nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
 Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016).
@@
 # Author Contributions {-#Authors}
-*[Name of author 1] and [Name of author 2] conceived the ideas and designed methodology; [Name of author 1] and [Name of author 3] collected the data; [Name of author 2] and [Name of author 4] analysed the data; [Name of author 1] and [Name of author 4] led the writing of the manuscript.
-All authors contributed critically to the drafts and gave final approval for publication.*
+
+LLM, RMH, KJB, IAK, EM, CT, JS, KAT, WJM and NAB conceived the ideas and designed methodology; RMH, KJB, IAK, EM, CT, CP, EGA, JJ, JS and LM collected the data; LLM, RH, KJB, IAK, EM and VH analyzed the data; LLM and RMH led the writing of the manuscript.
+All authors contributed critically to the drafts and gave final approval for publication.
 

### 2024-04-05 doc/15yrsChange.Rmd
acknowledgements and author contributions

@@
 
-- Funding: ForestGEO
+We gratefully acknowledge the many researchers who have collected data at the SCBI ForestGEO plot over the years. 
+This research was funded by Smithsonian's Forest Global Earth Observatory, Smithsonian's National Zoo and Conservation Biology Institute, NSF Macrosystems grant #2106015, and grants from Virginia Native Plants Society<!--2017 and 2019 mortality censuses--> and Shenandoah National Park Trust<!--2018 mortality census-->.
 
@@
 
-LLM, RMH, KJB, IAK, EM, CT, JS, KAT, WJM and NAB conceived the ideas and designed methodology; RMH, KJB, IAK, EM, CT, CP, EGA, JJ, JS and LM collected the data; LLM, RH, KJB, IAK, EM and VH analyzed the data; LLM and RMH led the writing of the manuscript.
+LLM, RMH, KJB, IAK, EM, CT, DM, JS, NAB, WJM, and KAT conceived the ideas and designed methodology; all researchers contributed to data collection; LLM, RH, KJB, IAK, EM and VH analyzed the data; LLM and RMH led the writing of the manuscript.
 All authors contributed critically to the drafts and gave final approval for publication.

### 2024-04-05 doc/15yrsChange.Rmd
data availability statement
@llmorreale , ideally we'll want to make this repo public today. Once it's public, we can archive in Zenodo, which will assign a DOI (which can refer to all versions). We'll want to clean it up before archiving in Zenodo.

@@
 
-# Data availability {-#Data}
+# Data and code availability {-#Data}
+
+All data and code supporting this analysis are available via the public GitHub repository "15yrsChange" within the SCBI-ForestGEO organization (https://github.com/SCBI-ForestGEO/15yrsChange) and archived in Zenodo (DOI: TBD)<!--this can be obtained once the repo is made public-->.
+The most up-to-date versions of the data products analyzed here are publicly available via the ForestGEO data portal (https://forestgeo.si.edu/explore-data) and the SCBI-ForestGEO GitHub organization (https://github.com/SCBI-ForestGEO).
 

### 2024-04-05 doc/15yrsChange.Rmd
integrated edits to results from google doc

@@
 
-Pests and pathogens; temperate forests; carbon sequestration; white-tailed deer; ecosystem processes;
+Pests and pathogens; temperate forests; carbon sequestration; white-tailed deer; ecosystem processes; invasive species
 
@@
 
-All sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3).
+Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3b).
 The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens.
-Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first census intervals.
+Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals.
 However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
@@
 The average annual AWM of these two species (`r avg_fram_quve` Mg C ha^-1^ yr^-1^), was higher than that of all other species combined (`r avg_all_other` Mg C ha^-1^ yr^-1^).
-Moreover, both species exhitibed significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.
+Moreover, both species exhibited significant increases in AWM since the start of the census, a trend also observed in three other species in the Oak genera—*Quercus rubra*, *Quercus prinus* and *Quercus alba*.
 
@@
 
-For the first time since the establishment of  the SCBI ForestGEO plot, biomass losses to mortality outpaced biomass gains from tree growth (Figure 3a).
-Living aboveground C stocks (AGB) decreased over the 2018-2023 census interval (Figure 2), with AGB declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure 3a).
+For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure 2), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
 Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
@@
  Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c).
-Plot-wide AWG decreased by `r (awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013 * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
+Plot-wide AWG decreased by `r round((awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013,1) * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
 Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
@@
 
-Deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5).
+Consistent with hypotehsis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5).
 The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^.
@@
 
-There were significant areas of decreasing AGB and suppressed regeneration of canopy species.
-Specifically, 86% of the plot has either experienced -∆AGB from 2008-2023 or has a low AGB of canopy taxa in the understory.
-Only 21.7% of the plot exhibits an understory with high AGB of species capable of reaching the canopy.
-Within the region characterized by high deer density and high canopy vulnerability 84% of the area has lost canopy biomass, only 5.4% has an understory containing high AGB of canopy taxa, and 0% of area exhibits both.
-In contrast, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB.
-Furthermore, 70.7% of this same region has high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration.
+There were significant areas of suppressed regeneration of canopy species, often coincident with decreasing AGB.
+Only 21.7% of the plot (mostly inside the deer exlosure) exhibited an understory with high AGB of species capable of reaching the canopy.
+Within the plot areas characterized by high deer density and high canopy vulnerability, 84% of the area had lost canopy biomass, only 5.4% had an understory containing high AGB of canopy taxa, and 0% of area exhibited both a healthy understory and increasing AGB.
+In contrast, 70.7% of the plot area with low deer density and low canopy vulnerability had high AGB of canopy species in the understory, indicating a higher likelihood of future C sequestration.
+Furthermore, 72.7% of the region with low deer density and low canopy vulnerability had a net increase of AGB.
 
 ![**_Figure `r regen_failure`.
-Characterizing the SCBI ForestGEO plot by changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023.
+Map of changes in canopy biomass (2008-2023) and the biomass of canopy taxa in the understory in 2023 within the SCBI ForestGEO plot._** (a) Each .04 ha quadrat is classified by the ∆AGB of canopy trees over the 15-year study period and the AGB of canopy taxa in the understory in 2023.
 (b) The % area of each region that falls within each classification.](display/Figure6.jpeg)

### 2024-04-05 code/0_Clean_2023_Census.r
cleaning up code directory

@@ -35,7 +35,7 @@ census2023_hom_corrections <- census2023  %>%
          taper_sp = if_else(true_sp %in% scbi_taper_coefficients$spcd, true_sp, "other"),
          hom = if_else((is.na(hom)) & status_current %in% c("LI"), 1.3, hom))  %>%  #correct living stems that have a HoM of 0 or NA -> 1.3m 
   left_join(scbi_taper_coefficients,by = c("taper_sp" = "spcd"))  %>% 
-  mutate(dbhc_mm = if_else(!(status_current %in% c("LI")), NA,  usfs_taper(dh = dbh_current / 1000, h = hom, stump_taper = Stump.taper.rate, stem_taper = Stem.taper.rate)) * 10,
+  mutate(dbhc_mm = if_else(!(status_current %in% c("LI")), NA,  usfs_taper(dh = dbh_current / 1000, h = hom, stump_taper = Stump.taper.rate, stem_taper = Stem.taper.rate)) * 10, #apply tapering corrections
          dbh_mm = dbh_current)  
 
 

### 2024-04-05 code/5_plot_annual_mort_by_species.r

@@ -45,7 +45,7 @@ facet_years <- c(2013, 2023)
 ##### II - Plot mortality by species #####
 plotdf <- mort_plot_df  %>% 
   mutate(plotspecies = factor(plotspecies, levels = facet_order))
-facet_labels <- if_else(!grepl("Other",facet_order), paste0("*",facet_order,"*"),facet_order)
+facet_labels <- if_else(!grepl("All other taxa",facet_order), paste0("*",facet_order,"*"),facet_order)
 names(facet_labels) <- facet_order
 
 barp <- ggplot(plotdf, aes(x = survey_year, y = mort_woody, fill = as.ordered(survey_year))) +

### 2024-04-05 code/exploratory/figure3_woodyGrowthDataPrep- temporary.R
clean up repository pt. 2

@@ -11,12 +11,12 @@ create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
 }
 
 #Read in data from ALL censuses and the species table
-grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
+grouped_quadrats <- read.csv("data/grouped_quadrats.csv")  %>% 
   mutate(quadrat = sprintf("%04d",quadrat))
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem1.corrected.rdata")
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem2.corrected.rdata")
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem3.corrected.rdata")
-load("C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/scbi.stem4.corrected.rdata")
+load("data/census_data/scbi.stem1.corrected.rdata")
+load("data/census_data/scbi.stem2.corrected.rdata")
+load("data/census_data/scbi.stem3.corrected.rdata")
+load("data/census_data/scbi.stem4.corrected.rdata")
 
 load(url("https://github.com/SCBI-ForestGEO/SCBI-ForestGEO-Data/raw/master/tree_main_census/data/scbi.spptable.rdata"))
 

### 2024-04-05 doc/15yrsChange.Rmd
discussion
- new line per sentence (easier for editing)
- a few changes in first couple sentences
- commented out future directions

@@
 
-The combination of -∆AGB and limited recruitment of canopy species in the understory is compromising the potential carbon sink of this forest, now and into the future.
-Decreases in AGB were largest in the regions of the plot with high canopy vulnerability to pests and pathogens.
+Our results reveal that aThe combination of declining biomass (-∆AGB) and limited recruitment of canopy species in the understory is compromising the potential carbon sink of the SCBI ForestGEO plot, now and into the future.
+Decreases in AGB were largest in the plot areas with high canopy vulnerability to pests and pathogens.
 These regions also exhibited the lowest recruitment of canopy species in the understory.
@@
 Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
-Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
-The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
+Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
 Furthermore,the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
@@
 
-# Future Directions
+<!--# Future Directions
 
@@
 Future evaluations of temperate forest health and carbon storage must account for the substantial effects of nuisance species along with other global changes.
+-->
 

### 2024-04-05 doc/15yrsChange.Rmd
discussion - minor changes from google doc

@@
 However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
-These mortality events are often termed “oak decline” events due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
+These mortality events are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
 Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).
@@
 Relying on the current understory as a proxy for future canopy presents challenges.
-White-tailed deer populations have varied greatly over the last century (citation).
+White-tailed deer populations have varied greatly over the last century [@citation].
 Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
-A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse, could allow for substantial, rapid regeneration of canopy species.
+A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
 However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
@@
 Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
-Furthermore,the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
+Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
 EAB consistently induces near-total mortality in *Fraxinus* spp.
@@
 Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
-Nuisance species, driven by human introduction and perturbations to ecological equilibriums, interact with other anthropogenic changes to modify forest ecosystems.
+Nuisance species, driven by human introduction and perturbations to ecological equilibria, interact with other anthropogenic changes to modify forest ecosystems.
 By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.

### 2024-04-05 code/4_calculate_woody_flux.r
figure 4 fixes

@@ -8,7 +8,7 @@ create_stem_UID <- function(census_df, TID_col = "tag", SID_col = "StemTag") {
   census_df[["UID"]] <- paste(census_df[[TID_col]], census_df[[SID_col]], sep = "_")
   return(census_df)
 }
-# Can skip to III and read-in all_censuses #
+# Can skip to III and read-in all_censuses if it is already created#
 
 ##### I - Read in data from ALL censuses and the species table #####
 grouped_quadrats <- read.csv("C:/Work/Smithsonian/Repos/15yrsChange/data/grouped_quadrats.csv")  %>% 
@@ -44,8 +44,10 @@ all_censuses <- scbi.stem1.corrected  %>%
 
 #save(all_censuses, file = "C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
 
-load(file = "C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+
 ##### III - Calculate woody fluxes by quadrat #####
+load(file = "C:/Work/Smithsonian/Repos/15yrsChange/data/census_data/all_censuses_agb.rdata")
+
 AWP <- all_censuses  %>% 
   filter(DFstatus %in% c("alive"))  %>% 
   mutate(WoodyGrowth = (ABG - lag(ABG)) / as.numeric(Meas_Int))  %>% 
@@ -86,16 +88,16 @@ write.csv(woody_fluxes, "C:/Work/Smithsonian/Repos/15yrsChange/data/processed_da
 
 ##### IV - Decomposing census mortality & recruitment by species #####
 
-### Mortality by species ###
-AWM_sp <- all_censuses  %>% 
-  group_by(UID) %>%  
-  arrange(UID,Census) %>% 
-  mutate(FutMeasInt = lead(Meas_Int))  %>% 
-  filter(DFstatus %in% c("alive") & lead(DFstatus) %in% c("dead","broken_below","stem dead"))  %>% 
-  mutate(WoodyMort = ABG / as.numeric(FutMeasInt), Census = Census + 1)  %>% 
-  ungroup()  %>% 
-  group_by(Census,sp)  %>%
-  summarize(AWM = sum(WoodyMort,na.rm = T) / 1000 /.47)
+# ### Mortality by species ###
+# AWM_sp <- all_censuses  %>% 
+#   group_by(UID) %>%  
[…]

### 2024-04-05 doc/15yrsChange.Rmd
discussion: revamp par. 1

@@
 
-for submission to Anthromes joint special collection in New Phytologist/ Plants, People, Planet: https://www.newphytologist.org/news/view/332
+for submission to Anthromes joint special collection in New Phytologist: https://www.newphytologist.org/news/view/332
 
@@
 
-Our results reveal that aThe combination of declining biomass (-∆AGB) and limited recruitment of canopy species in the understory is compromising the potential carbon sink of the SCBI ForestGEO plot, now and into the future.
-Decreases in AGB were largest in the plot areas with high canopy vulnerability to pests and pathogens.
-These regions also exhibited the lowest recruitment of canopy species in the understory.
-Deer severely depleted regeneration of canopy species and the region with low deer density maintained the highest amount of canopy species in the understory.
-Despite localized increases in AGB, notably in areas with low canopy vulnerability, on net, these increases are not outpacing AGB lost through a combination of increased AWM and decreasing AWG.
-Total ABG stocks in the plot were lower in 2023 than in 2013.
-Moreover, the low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
+Our results reveal that nuisance species have compromised biomass stocks and regenerative capacity at the SCBI ForestGEO plot, reducing the recent and future carbon sink of this forest.
+Canopy tree mortality and associated AWM increased, particularly in plot areas affected most heavily by non-indigenous insects and pathogens (hypothesis 1, Figs. 2-3).
+These increases in AWM were not offset by biomass increases through AWP, which decreased, or recruitment, resulting in net biomass loss (hypothesis 2, Figs. 2,4).
+This net AGB loss occurred despite localized increases in AGB across most of the plot, particularly in areas with low canopy vulnerability (Fig. 6).
+Recruitment of canopy tree species was low throughout the plot (Fig. 5), and there was low biomass of canopy species in the understory, particularly in areas with high deer density (Fig. 6). 
+Moreover, low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.
+Thus, nuisance species have severely compromised both recent and future carbon sequestration in this forest, contradicting models projecting increases in AGB [@wu_uncertainty_2023].
 

### 2024-04-05 doc/15yrsChange.Rmd
renumber figures (code)

@@
 
-biomass_trends = 2 #line plot showing biomass through time for entire plot and each of the 3 focal sections
+loss_gain_net = 2 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard
 
-loss_gain_net = 3 # 4-panel figure showing ∆AGB, loss, and gain for (a) whole plot, (b) deer exclosure, (c) upland, (d) graveyard
+mortality_by_genera = 3 # 2-panel figure showing annual biomass mortality by genera for (a) canopy species and (b) understory species
 
-mortality_by_genera = 4 # 2-panel figure showing annual biomass mortality by genera for (a) canopy species and (b) understory species
+biomass_trends = 4 #line plot showing biomass through time for entire plot and each of the 3 focal sections
 

### 2024-04-05 doc/15yrsChange.Rmd
update & automate figure numbering
@llmorreale , FYI

@@
 
-Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure 3b).
+Consistent with hypothesis 1, all sections of the plot experienced increased mortality (AWM) since plot establishment in 2008 (Figure `r loss_gain_net`b).
 The largest increases in mortality occurred in areas with the highest vulnerability to non-indigenous pests and pathogens.
@@
 Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species.
-The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure 4).
+The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure `r mortality_by_genera`).
 *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
@@
 
-Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure 3a).
-For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure 2), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
-Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure 3a).
+Consistent with hypothesis 2, biomass losses to mortality outpaced biomass gains from tree growth over the 2018-23 census interval (Figure `r loss_gain_net`a).
+For the first time since establishment of the SCBI ForestGEO plot, living aboveground C stocks (AGB) decreased between 2018 and 2023 (Figure `r biomass_trends`), declining from `r wp_AGB_2018` to `r wp_AGB_2023` Mg C ha^-1^.
+Prior to this period, AGB had consistently increased at a rate of `r mean(awg_wholePlot_2013,awg_wholePlot_2018)` Mg C ha^-1^ yr ^-1^ between 2008 and 2018 (Figure `r loss_gain_net`a).
 In regions characterized by high deer density and high canopy vulnerability, > 33%  of AGB was lost over the past two censuses: dropping from `r hdhn_AGB_2013` Mg C ha^-1^ in 2013 to `r hdhn_AGB_2018` Mg C ha^-1^ in 2018 and then to `r hdhn_AGB_2023` Mg C ha^-1^ in 2023.
@@
 
- Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure 3c).
+ Transition from net AGB accumulation to loss was not solely attributable to increased mortality; biomass accrual from woody growth (AWG) declined across all census intervals, even when AWM did not increase (Figure `r loss_gain_net`c).
 Plot-wide AWG decreased by `r round((awg_wholePlot_2023 - awg_wholePlot_2013) /  awg_wholePlot_2013,1) * 100` %, from `r awg_wholePlot_2013` Mg C ha^-1^ yr^-1^ over 2008-2013 to `r awg_wholePlot_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 interval.
 Areas with low canopy vulnerability consistently displayed the highest AWG (`r awg_group1_2023` Mg C ha^-1^ yr^-1^ in the 2018-2023 period), while the AWG in the portion with high canopy vulnerability was significantly lower (`r awg_group3_2023`Mg C ha^-1^ yr^-1^).
-Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure 3d).
+Woody recruitment flux (AWR), constituting a minor fraction of net biomass change, showed no significant temporal trend (Figure `r loss_gain_net`d).
 AWR consistently remained lowest in areas characterized by high deer density and low canopy vulnerability, whereas it reached its highest levels in regions with high canopy vulnerability, recording rates of `r awr_group2_avg` Mg C ha^-1^ yr^-1^ and `r awr_group3_avg` Mg C ha^-1^ yr^-1^, respectively.
@@
 Our results reveal that nuisance species have compromised biomass stocks and regenerative capacity at the SCBI ForestGEO plot, reducing the recent and future carbon sink of this forest.
-Canopy tree mortality and associated AWM increased, particularly in plot areas affected most heavily by non-indigenous insects and pathogens (hypothesis 1, Figs. 2-3).
-These increases in AWM were not offset by biomass increases through AWP, which decreased, or recruitment, resulting in net biomass loss (hypothesis 2, Figs. 2,4).
-This net AGB loss occurred despite localized increases in AGB across most of the plot, particularly in areas with low canopy vulnerability (Fig. 6).
-Recruitment of canopy tree species was low throughout the plot (Fig. 5), and there was low biomass of canopy species in the understory, particularly in areas with high deer density (Fig. 6). 
+Canopy tree mortality and associated AWM increased, particularly in plot areas affected most heavily by non-indigenous insects and pathogens (hypothesis 1, Figs.
+`r loss_gain_net`- `r mortality_by_genera`).
+These increases in AWM were not offset by biomass increases through AWP, which decreased, or recruitment, resulting in net biomass loss (hypothesis 2, Figs. `r loss_gain_net`, `r biomass_trends`).
+This net AGB loss occurred despite localized increases in AGB across most of the plot, particularly in areas with low canopy vulnerability (Fig. `r regen_failure`).
+Recruitment of canopy tree species was low throughout the plot (Fig.
+`r recruitment_by_genera`), and there was low biomass of canopy species in the understory, particularly in areas with high deer density (Fig.
+`r regen_failure`).
 Moreover, low recruitment of canopy species in areas with high canopy vulnerability suggests that heavy deer browsing on canopy species may be exacerbated in areas with canopy openings due to nuisance-species driven mortality where fast-growing understory species that are resistant to deer can take advantage, further competing with canopy recruitment.

### 2024-04-05 code/6_plot_woody_fluxes.R
add panel letters to woody flux fig

@@ -57,6 +57,11 @@ fig3 <- ggplot(fig3_woodyfluxes, aes(x = as.ordered(cens_int), y = MgC_Yr_Ha, gr
   geom_point(data = fig3_woodyfluxes %>% filter(Group %in% c("3")), pch = 22, cex = 3, fill = "#750000") +
   geom_point (data = fig3_woodyfluxes %>% filter(Group %in% c("Whole Plot")), pch = 16, cex = 3, fill = "#7e937f") +
   geom_hline(data = fig3_woodyfluxes  %>% filter(Flux %in% c("NetFlux")),aes(yintercept = 0), lty = "dashed") +
+  geom_text(data = fig3_woodyfluxes  %>% 
+              group_by(Flux)  %>%  
+              slice(1)  %>% 
+              mutate(label = paste0("(",letters[as.numeric(Flux)],")")),
+              aes(label = label, x = .6, y = c(1.2,7.4,2.85,0.0295)), color = "black", fontface = "bold", cex = 8) +
   scale_color_manual(name = element_blank(),labels = c("Low deer, low canopy vulnerability","High deer, low canopy vulnerability", "High deer, high canopy vulnerability", "Whole Plot"), values = colz, guide = guide_legend(nrow = 2)) +
   labs(y = expression("Carbon Flux"~(Mg~C~ha^-1~yr^-1)), x = "Census") +
   theme_bw() +

### 2024-04-05 doc/15yrsChange.Rmd
updates to discussion

@@
 ---
-title: ''
+title: '15yrsChange'
 output: 
@@
 The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
-We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021.
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a.
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
@@
 
-Consistent with hypotehsis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5).
+Consistent with hypothesis 3, deer density had strong mediating effects on the composition of woody stem recruitment (Figure 5).
 The region characterized by low deer density exhibited significantly higher annual recruitment of canopy taxa compared to those areas with higher deer density— specifically, `r round(group1_canopy,1)` stems ha^-1^ yr^-1^ compared to `r round(group2_3_canopy)` ha^-1^ yr^-1^.
@@
 Relying on the current understory as a proxy for future canopy presents challenges.
-White-tailed deer populations have varied greatly over the last century [@citation].
-Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events (Koenig, 2021).
+White-tailed deer populations have varied greatly over the last century [@hanberry_regaining_2020].
+Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events [@koenig_brief_2021].
 A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
@@
 
-Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting broader applicability.
+Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
 Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
@@
 once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
+The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
+Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020].
 

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 1

@@
 
-The temperate forest biome plays a critical role in the global carbon cycle, accounting for almost half of the net global forest carbon (C) sink [@harris_global_2021], with temperate deciduous forests alone sequestering >300 Tg C yr^-1^ [@pugh_role_2019].
-Although currently a net C sink, the future of the temperate forest biome remains uncertain.
+The temperate forest biome plays a critical role in the global carbon cycle (C), accounting for almost half of the net global forest C sink [@harris_global_2021], and temperate deciduous forests alone sequester >300 Tg C yr^-1^ [@pugh_role_2019].
+Although it is currently a net C sink, the future of the temperate forest biome remains uncertain.
 The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012].
-Yet, global C models predict a wide range of future trajectories of CO~2~ sequestration [@ahlstrom_robustness_2012;@arora_carbon_2020]<!--check refs-->, and current global C models do not represent some influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
-For instance, one mechanism not represented in global C models is the impact of nuisance species – i.e., indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, resulting in detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
-Similarly, the impact of nuisance species on forest carbon budgets is not considered in machine learning or niche models that seek to project future forest distribution and carbon stocks, nor in carbon offset projects [@wu_uncertainty_2023].
-This is problematic in that nuisance species – including non-indigenous insect pests and pathogens and over-abundant herbivores – are dramatically impacting carbon cycling in temperate deciduous forests around the world.
+Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020]<!--check refs-->, and current global C models do not represent certain influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
+For instance, the impact of nuisance species, is not represented in global C - i.e., the impact of indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, leading to detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
+The impact of nuisance species on forest C budgets is also not considered in machine learning or niche models that seek to project future forest distribution and C stocks, nor is it in C offset projects [@wu_uncertainty_2023].
+Nuisance species, including non-indigenous insect pests and pathogens and over-abundant herbivores, are dramatically impacting carbon cycling in temperate deciduous forests around the world, and failure to account for them significantly limits understanding of future forest health and C sequestration potential.
 
-Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling, successional trajectories, and geomorphic processes [@lovett_nonnative_2016; @herms_emerald_2014; @fischer_disturbances_2013] (Fei et al 2014).
+Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling and successional trajectories [@lovett_nonnative_2016; @herms_emerald_2014].
 Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006].

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 2

@@
 Non-indigenous pests and pathogens can have a protracted impact on forests, often modifying forest productivity, nutrient cycling and successional trajectories [@lovett_nonnative_2016; @herms_emerald_2014].
-Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013; @aukema_historical_2010; @brockerhoff_ecology_2006].
+Historically introduced to new ecosystems through intercontinental trade, non-indigenous forest insects and diseases will continue to proliferate with increased globalization and human movement [@fischer_disturbances_2013a; @aukema_historical_2010; @brockerhoff_ecology_2006].
 Many of the prominent pests and pathogens in temperate deciduous forests have dramatic effects on canopy composition and structure, decimating once-abundant canopy species [@aukema_historical_2010; @anderson-teixeira_longterm_2021; @klooster_ash_2014; @ellison_present_2018].
-Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion from nuisance species [@fei_biomass_2019], further increasing potentials for biomass loss and altered carbon cycle dynamics.
+Since many pests and pathogens do not currently occupy the full extent of their host species range, a significant portion of the remaining forest canopy is susceptible to future invasion by other nuisance species [@fei_biomass_2019].
 
@@
 The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
-Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020].
+Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
 

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 3

@@
 
-Another category of nuisance species includes mammals whose population densities have been inflated due to human activities [@murphy_invasive_2014; @clout_invasion_2008], including white-tailed deer (*Odocoileus virginianus*) in U.S.
-eastern deciduous forests.
+Mammals whose population densities have been inflated due to human activities are also considered nuisance species [@murphy_invasive_2014; @clout_invasion_2008], and in United States (US) eastern deciduous forests, this includes white-tailed deer (*Odocoileus virginianus*).
 Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000].
-White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces survival, growth and density [@mcgarvey_effects_2013a].
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant species [@bugalho_effects_2013a] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (Averill *et al* 2018).
-The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change  results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
+White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a].
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species [*e.g.*, oak, *Quercus* spp.; @bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017).
+The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
 

### 2024-04-05 doc/15yrsChange.Rmd
revamp discussion par 2
@llmorreale , FYI

--- 2024-04-05 doc/15yrsChange.Rmd
@@
 
-While nuisance species have had a large impact on AGB loss at our site (primarily through increased AWM of *Fraxinus americana*; Anderson-Teixeira et al, 2021), mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well.
-Oak species in the region are generally found in older, even aged cohorts, which can contribute to accelerated mortality rates (Oak et al., 2016).
-However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century.
-These mortality events are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
-Oak decline is believed to be the result of an imbalance in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).
+While nuisance species have had a large impact on AGB loss at our site, simultaneous changes including stand aging and climate change have also contributed to observed trends.
+The majority of AGB loss has occurred through increased EAB-driven mortality of *Fraxinus spp.* [@anderson-teixeira_longterm_2021], yet mortality among oak species, which is not linked to a non-indigenous nuisance species, is substantial as well (Fig.
+`r mortality_by_genera`).
+As is the case at our site, oak species in the region are generally found in older, relatively even aged cohorts, and this can contribute to accelerated mortality rates (Oak et al., 2016).
+However, large oak mortality events, which often occur over protracted periods of time, have been common globally over the past century and are often termed “oak decline” due to the complex abiotic and biotic factors which interact to incite mortality (Manion and Lachance, 1992).
+Oak decline is believed to be the result of climate-driven imbalances in carbohydrate dynamics, which weakens oak trees’ resistance to pest and pathogen invasion, eventually leading to mortality (Manion and Lachance, 1992).<!--KAT added "climate-driven" .
+Not entirely sure that's true -- need to verify.-->
+Similarly, stand aging and climate change may both have contributed to the observed decline in AWP, which in turn contributed to reduced AGB.
+Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures **(Anderson-Teixeira et al.
+2022, Dow et al.
+2022)**.<!--refs need to be inserted.-->
+These complexities underscore the importance of multiple anthropogenic stressors, including but not limited to nuisance species, in shaping 21^st^ century forest carbon dynamics.
+
+The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as simple directional trends in the future.
+The top dirver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
+...
 

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 4

@@
 
-In combination, increased canopy tree mortality and reduced canopy species recruitment in the understory is likely to have profound impacts on deciduous forests of eastern North America.
-When global change pressures (e.g., climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, whereby there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
-In the mid-Atlantic region of the eastern United States, forests face a severe regeneration debt; there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
-The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., *Acer* spp., *Fagus grandifolia*) than historically dominated much of the region (e.g., *Quercus* spp., *Carya* spp., [@nowacki_climate_2015]; Thompson *et al.* 2013), but even these are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
+Increased canopy tree mortality in combination with reduced canopy species recruitment in the understory is likely to have profound impacts on the deciduous forests of eastern North America.
+When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
+In the mid-Atlantic region of the eastern US, forests face a severe regeneration debt: there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
+The juvenile canopy trees that are most common in contemporary mid-Atlantic forests represent a more mesophytic set of species (e.g., maple, *Acer* spp.; American beech, *Fagus grandifolia*) than historically dominated much of the region (e.g., oak, *Quercus* spp.; hickory, *Carya* spp.; [@nowacki_climate_2015; @thompson_four_2013], but even these species associated with mesic conditions are often not present in sufficient abundance to replace the current canopy [@miller_compounding_2019].
 

### 2024-04-05 doc/15yrsChange.Rmd
update introduction pt. 5

@@
 Understanding the combined effects of multiple nuisance species on forest ecosystem processes requires data on both the forest canopy and understory over an extended time.
-Here, we use 15 years of detailed tree census from a large forest dynamics plot the mid-Atlantic region of the eastern United States to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest carbon balance and regeneration potential.
+Here, we use 15 years of detailed tree census data from a large forest dynamics plot in the mid-Atlantic region of the eastern US to understand the effects of non-indigenous pests and pathogens and white-tailed deer on the forest C balance and regeneration potential.
 We test the following hypotheses: (1) canopy tree mortality and associated woody biomass loss are increasing, in large part due to non-indigenous nuisance species, (2) biomass gains from tree growth and recruitment have not kept pace with tree mortality, resulting in a decline in aboveground living biomass (-∆AGB), and (3) deer browsing has decimated the abundance of canopy tree saplings, suggesting that future tree mortality will result in substantial net biomass loss and sustained decrease in the C storage potential of this ecosystem.

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 1

@@
 
-The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
+The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (NZCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
 The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (Smithsonian Institution archives, @bourg_initial_2013, @mcgarvey_effects_2013a).
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]..
 The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 2

@@
 
-In addition to our consideration of the entire forest plot, we focus further on three portions of the plot, all upland forest (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig.
+In addition to our consideration of the entire forest plot, we focus further on three upland forest portions of the plot (i.e., excluding low-lying areas around streams), with varying levels of exposure to non-indigenous insects and pathogens and white-tailed deer (Fig.
 1).
-The first focal area we define as“Low deer, low canopy vulnerability”: A 4-ha portion of upland forest in the SE quarter of the plot (Fig.
-1a), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
-This area has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
-The second portion is a “high deer, low canopy vulnerability” 16.3 ha upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
-1b).
-The third section of the plot has high deer, high canopy vulnerability: corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
-We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a.
-We defined vulnerability to non-indigenous insects and pathogens based on the abundance of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment (2008).
+The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig.
+**1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
+This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
+The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
+**1b**).
+The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by [@mcgregor_tree_2021a].
+We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008.
 The following canopy species were identified as vulnerable: ash (*Fraxinus* spp.), elm (*Ulmus* spp.), butternut (*Juglans cinerea*), and American chestnut (*Castanea dentata*) [@anderson-teixeira_longterm_2021].

### 2024-04-05 doc/15yrsChange.Rmd
discussion: add paragraph + tweaks

@@
 Not entirely sure that's true -- need to verify.-->
-Similarly, stand aging and climate change may both have contributed to the observed decline in AWP, which in turn contributed to reduced AGB.
-Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures **(Anderson-Teixeira et al.
-2022, Dow et al.
-2022)**.<!--refs need to be inserted.-->
+Similarly, stand aging and climate change may both have contributed to the observed decline in AWP (Fig.
+`r loss_gain_net`b), which in turn contributed to reduced AGB (Figs.
+`r loss_gain_net`, `r biomass_trends`).
+Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures [@dow_warm_2022; @anderson-teixeira_joint_2022].
 These complexities underscore the importance of multiple anthropogenic stressors, including but not limited to nuisance species, in shaping 21^st^ century forest carbon dynamics.
-
-The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as simple directional trends in the future.
-The top dirver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
-...
 
@@
 When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.
+
+The observed trends in AGB loss, including increasing AWM and decreasing AWP, will not necessarily be maintained as continuous directional trends in the future.
+The top driver of increasing AWM has been EAB-driven mortality of *Fraxinus spp.*, but *Fraxinus* AWM has started declining as most individuals ≥10 cm have already died (Fig. `r mortality_by_genera`).
+It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021].
+While it is also possible that AWM will return to its previously low level as EAB recedes, we consider it very likely that oak decline, additional non-indigenous insects or pathogen species (e.g., beech leaf disease), and climate stressors will continue to elevate tree mortality at this site in the coming decades.
+Similarly, the observed long-term declines at this site in AWP (Fig.
+`r loss_gain_net`b) and individual tree growth [@anderson-teixeira_joint_2022] remain poorly understood, hindering prediction of their future trends.
+Overall, there remains large uncertainty as to the future trend in AGB at this site, but our analysis makes it clear that these trends have been and will continue to be substantially impacted by nuisance species.
 

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 3

@@
 The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
-We defined upland forest according to topographic wetness index, which was calculated for the forest plot by [@mcgregor_tree_2021a].
+We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a.
 We defined vulnerability to non-indigenous insects and pathogens based on the abundance (>20% of tree basal area) of canopy species affected by non-indigenous pests and pathogens at the time of plot establishment in 2008.
@@
 
-As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years according to the protocol detailed in @condit_tropical_1998.
+As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years following the protocol detailed in @condit_tropical_1998.
 This inventory, hereafter referred to as the census, includes all stems greater than one centimeter in diameter at breast height (1.3 meters; dbh).
@@
 In multi-stemmed individuals, each additional stem that surpasses the dbh threshold receives a stem number and associated tag.
-The plot was established in 2008 and recensused three times (2013, 2018, 2023) since, comprising 15 years of forest dynamics data.
+The plot was established in 2008 and recensused three times since (2013, 2018, 2023), providing 15 years of forest dynamics data.
 In the last census, systematic data cleaning was continuously implemented 1) in an ESRI ArcGIS Field Maps application via built-in checks and 2) on a daily basis, via error reports generated through GitHub actions [@kim_implementing_2022].
-Each individual’s location was recorded in the Field Maps application using a digitized map of the plot (citation).
+Each stem’s location was recorded in the Field Maps application using a digitized map of the plot.
 In addition to the ForestGEO census, a tree health and mortality survey has been conducted annually starting in 2014 on all trees with a dbh ≥ 10 cm in the previous census [@gonzalez-akre_patterns_2016].
-Data were collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of tree health, such as physical damage, potential pathogens and insect infestation.
+Data were collected on the current status, canopy position, percentage of living and intact crown, and visible indicators of tree health, (*e.g.* physical damage, potential pathogens, insect infestation).
 

### 2024-04-05 doc/15yrsChange.Rmd
tweaks to recruitment paragraph

--- 2024-04-05 doc/15yrsChange.Rmd
@@
 
-Relying on the current understory as a proxy for future canopy presents challenges.
-White-tailed deer populations have varied greatly over the last century [@hanberry_regaining_2020].
-Moreover, some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events [@koenig_brief_2021].
-A decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
+Similarly, a complex chain of processes -- from seed production to germination and seedling survival -- affects sapling recruitment, and relying on the current understory as a proxy for future canopy presents challenges.
+Some canopy species, including some *Quercus* spp., are hypothesized to recruit in synchronized pulse events [@koenig_brief_2021].
+Moreover, white-tailed deer populations have varied greatly over the last century [@hanberry_regaining_2020], and a decline in deer population that reduces browsing pressure and coincides with a recruitment pulse could allow for substantial, rapid regeneration of canopy species.
 However, current recruitment levels are below a rate which we would expect in maturing secondary forests.
-When overlaid with increasing mortality, regeneration may be insufficient to prevent extended periods without a closed canopy.
+When overlaid with increasing mortality, regeneration is likely to be insufficient to prevent extended periods without a closed canopy.
 

### 2024-04-05 doc/15yrsChange.Rmd
discussion: edited another paragraph, plus a bit of rearrangement

@@
 It is possible that increasing AWM will be maintained by trends of increasing oak mortality or by other non-indigenous insects or pathogen, which have arrived in the region in multiple waves over the past century [@anderson-teixeira_longterm_2021].
-While it is also possible that AWM will return to its previously low level as EAB recedes, we consider it very likely that oak decline, additional non-indigenous insects or pathogen species (e.g., beech leaf disease), and climate stressors will continue to elevate tree mortality at this site in the coming decades.
+The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
+Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
+Thus, we consider it likely that oak decline, additional non-indigenous insects or pathogen species, and climate stressors will continue to elevate tree mortality at this site in the coming decades.
 Similarly, the observed long-term declines at this site in AWP (Fig.
@@
 
-Despite inherent limitations in extrapolating findings from a single site, we find robust evidence suggesting a broader applicability of our results. 
-Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region. The persistently elevated deer populations documented at this site are reported throughout eastern US forests.
+Despite inherent limitations in extrapolating findings into the future and from a single site to a broader region, we find robust evidence suggesting a broader applicability of our results. 
+Mixed-hardwood forests, akin to our study site, are typical of the mid-Atlantic region.
 Furthermore, the widespread occurrence of vulnerable species (*e.g.* *Fraxinus* spp.) and continued spread of non-indigenous nuisance species (*e.g.* emerald ash borer, EAB) reinforces the relevance of our findings to the broader regional context.
@@
 once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
+Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
+<!--MOVED UP:
 The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
 Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
+-->
+Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].
 

### 2024-04-05 doc/15yrsChange.Rmd
update methods pt. 4
update authors and methods

@@
 
-Luca L. Morreale^1,2^*
+Luca L. Morreale^1,2^* ( Orcid ID: 0000-0003-2023-9150))
 
@@
 
-William J. McShea^1^
+William J. McShea^1^ ( Orcid ID : 0000-0002-8102-0200)
 
@@
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a]..
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a].
 The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
@@
 1).
-The first portion we define as “Low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig.
+The first focal area we define as “low deer, low canopy vulnerability”: a 4-ha portion of upland forest in the SE quarter of the plot (Fig.
 **1a**), fenced in 1990 and maintained with only occasional deer presence over the past 24 years.
 This portion has a low abundance of canopy species vulnerable to non-indigenous pests and pathogens [@holm_interactive_2013].
-The second focal area is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
+The second portion is “high deer, low canopy vulnerability”: 16.3 ha of upland forest area outside the deer exclosure with low abundance of canopy species affected by non-indigenous pests and pathogens (Fig.
 **1b**).
-The third portion of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
+The third focal area of the plot is "high deer, high canopy vulnerability": corresponding to 2.2 ha of upland forest outside the deer exclosure with high abundance of canopy species affected by non-indigenous pests and pathogens.
 We defined upland forest according to topographic wetness index, which was calculated for the forest plot by @mcgregor_tree_2021a.
@@
 The *allodb* package creates species-specific allometric equations by weighting all applicable allometries within its database filtered by tree size, geography, and taxonomic information.
-Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot study regions and the entire 25.6-ha plot.
+Using the calculated stem AGB, we estimated AGB stocks at each census year for each of the plot regions, as well as the entire 25.6-ha plot.
 We quantified trends in AGB over time by calculating annualized AGB fluxes including: 1) the net change in AGB, 2) aboveground woody growth (AWG; the addition of woody biomass from tree growth), 3) aboveground woody mortality (AWM; the change of living biomass to necromass due to tree mortality), and 4) aboveground woody recruitment (AWR; the addition of living biomass through recruitment) for each census interval and plot region.

### 2024-04-05 doc/15yrsChange.Rmd
figure 6 fixes & edits to discussion

@@
 ---
-title: '15yrsChange'
 output: 
@@
 ![**_Figure `r recruitment_by_genera`.
-Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20x that of canopy species recruitment.](display/Figure5.jpeg)
+Average annual recruitment (2008-2023) of canopy and understory tree species at the SCBI ForestGEO plot across areas differentiated by intensity of deer browsing and 2008 abundance of canopy tree species vulnerable to non-indigenous pests and pathogens._** Note that y-axes differ, with understory species recruitment > 20 times that of canopy species recruitment.](display/Figure5.jpeg)
 
@@
 Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
-<!--MOVED UP:
-The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
-Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
--->
 Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].
 
-Our results underscore the importance of accounting for nuisance species in higher-level models and understanding of temperate forest ecosystem processes.
-Nuisance species, driven by human introduction and perturbations to ecological equilibria, interact with other anthropogenic changes to modify forest ecosystems.
-By increasing mortality, nuisance species have the capacity to alter short-term carbon dynamics, resulting in large carbon losses that would not be expected in models that exclude them.
-Through recruitment suppression, nuisance species can have long-term effects that may alter successional dynamics.
-They can delay recovery from disturbances, anthropogenic or natural, and even suppress regeneration into a closed-canopy for sustained periods.
-They also can decrease resilience to other pressures as pests and pathogens decrease canopy diversity by shrinking the number of possible canopy species.
+Accounting for nuisance species in carbon models and future forest predictions is imperative.
+Our results underscore how ignoring nuisance species at our site, and likely much of the region, creates an expectation that temperate decidious forests will keep growing steadily and accumulating C [@wu_uncertainty_2023].
+Instead, by increasing tree mortality, nusicance species alter short-term C dynamics and result in large decreases in living woody biomass that are not predicted by models that fail to account for their impact.
+A result of human introduction and perturbations to ecological equilibria, nuisance species co-occur with other anthropogenic changes and may have interactive effects on forest ecosystems.
+Through recruitment suppression, nuisance species can delay recovery from disturbances, anthropogenic or natural [@bradshaw_impacts_2016; @harvey_emergent_2023], and may even suppress regeneration into a closed-canopy for sustained periods.
+Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threat_2020; @reed_distribution_2022; @knight_factors_2013].
+Elimination of canopy species reduces biodiversity, which can decrease resilience to other disturbances (*e.g.* drought; [@anderegg_hydraulic_2018]) further endagerning forest health and C sequestration potential.
 
-Nuisance species pose a serious threat to the health of temperate forests.
-But, this work highlights the potential of active management to protect them.
-Increased understory recruitment within the deer exclosure demonstrates that lowering deer populations can reduce browsing pressure.
-The direct linkage between non-indigenous pests and pathogens and decreased carbon storage demonstrates the need for natural climate solutions that focus on preserving the forests that we have instead of planting new ones.
-Opportunities exist to expand efforts that prevent further spread, increase the application of extant treatments, and continue research into natural immunity.
-
-<!--# Future Directions
-
-Understanding interactions between global change drivers, an ongoing area of research, is critical.
-We would expect that trees unaffected by nuisance species (*e.g.* *Liriodendron tulipifera*, *Quercus rubra*) might see productivity increases as canopy gaps form and competition from nearby trees decreases.
-However, continuous decreases in AWG throughout the plot suggests that other mechanisms, including increasing maximum temperatures during the growing season (Dow et al 2022), may be hindering tree growth.
-Tree seedlings of many species, already decimated by deer browsing, are further threatened by rising maximum temperatures (McEwan et al 2011).
-Increasing mortality within *Quercus* species is another focal area of future research, as species within the *Quercus* genera constitute > 33% of woody productivity at this site (McGregor et al 2020).
-Multiple global changes already impact the SCBI ForestGEO plot; carbon storage in the site is, and will be, mediated by multifaceted, interacting pressures resulting from human actions.
-
-Future research should build on our results by expanding these analyses to other temperate forest regions.
-While other areas will have different forest compositions and herbivory dynamics, increasing pressure from pests and pathogens is pervasive throughout temperate forests (Santini et al.
-2012, Ayres and Lombardero 2018).
-Moreover, while not a large factor at this site, the spread of non-indigenous plant species can further threaten the  regeneration of woody canopy species (Pile Knapp et al, 2023).
-Large forest dynamics sites (*i.e.* other sites in the ForestGEO network) are well-suited to this research and can be leveraged to quantify how nuisance species compromise present and future forest carbon sequestration.
-Expanding the scale of these analyses may also facilitate the incorporation of nuisance species into global vegetation and carbon models.
-Future evaluations of temperate forest health and carbon storage must account for the substantial effects of nuisance species along with other global changes.
--->
+Nuisance species pose a serious threat to the health of temperate forests. However, active forest management offers pathways to protect these ecosystems.
+At our site, increased understory recruitment within the deer exclosure supports  that lowering deer populations can reduce browsing pressure. 
+Interventions that help reduce, or even eliminate, pests and pathogens  (*e.g.* monitoring programs, eradication technologies) exist and are improving  [@simberloff_impacts_2013].
+Management of nuisance species at the scales necessary unquestionably will require enormous investment and effort.
+However, protecting existing forests represents a far more effective natural climate than emerging efforts to plant new ones [@cook-patton_protect_2021].
 

### 2024-04-05 code/8_plot_canopy_understory_biomass_changes.r

@@ -105,17 +105,22 @@ plotdf <- quadrat_shp  %>%
   left_join(bivariate_color_scale) 
 
 map <- ggplot(data = plotdf) +
+    labs(tag = "(a)") +
     scale_alpha(name = "",range = c(.6,0),guide = F) +
-    geom_sf(aes(fill = fill),color = "white",size = .1) +
+    geom_sf(aes(fill = fill),color = "black",size = .1) +
     geom_sf(data = deer_exclosure, colour = "black", fill = NA,lwd = 1.5) +
     scale_fill_identity() +
     # geom_sf(data = deer_exclosure, colour = "black", fill = NA,lwd = 1.5) +
-    theme_classic() +
-    theme(plot.margin = unit(c(0,0,0,10), "cm"),
+#    theme_classic() +
+    theme_bw() +
+    theme(plot.margin = unit(c(0,0,0,0), "cm"),
           legend.position = "none",
-          axis.line = element_blank(),
+          #axis.line = element_blank(),
           axis.text = element_blank(),
-          axis.ticks = element_blank())
+          #axis.ticks = element_blank(),
+       #   plot.background = element_rect(color = "black", linewidth = 1),
+          plot.tag = element_text(face = "bold",size = 18),
+          plot.tag.position = c(.05,.98))
 
 
 legend <- ggplot() +
@@ -134,8 +139,9 @@ legend <- ggplot() +
   theme_classic() +
   # make font small enough
   theme(
-    axis.text.x = element_text(angle = 45, hjust = 1),
-    axis.title = element_text(size = 12)
+    axis.text.x = element_text(angle = 45, hjust = 1,size = 12),
+    axis.text.y = element_text(size = 14),
+    axis.title = element_text(size = 16)
   ) +
   # quadratic tiles
   coord_fixed()
@@ -172,23 +178,32 @@ barp_df2 <- barp_df  %>%
 barp <- ggplot(barp_df) +
   geom_bar(stat = "identity",aes(x = as.character(Group), y = pct_area,group = plot_group ,fill = fill)) +
 #  geom_bar(data = barp_df2, stat = "identity", aes( x = as.character(Group), y = pct_area,group = factor(understory)), col = "white", fill = "transparent", lwd = 1.2) +
-  geom_vline(xintercept = 3.5) +
+  geom_vline(xintercept = 3.5,lwd = 1.3) +
[…]

### 2024-04-05 doc/15yrsChange.Rmd
minor edits

@@
 
-for submission to Anthromes joint special collection in New Phytologist: https://www.newphytologist.org/news/view/332
+<!-- for submission to Anthromes joint special collection in New Phytologist: https://www.newphytologist.org/news/view/332  -->
 
@@
 
-*(Word limit: 6,500 words)*
-
 # Summary  {-#Abstract}
 
-- The future of the important temperate forest carbon (C) sink is dependent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood.
+- The future of the critical temperate forest carbon (C) sink is contigent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood.
 Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration.
-- We analyzed biomass change and regenerative capacity in sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens. 
+- We analyzed biomass change and regenerative capacity in varying sections of the plot classified by deer browsing pressure and the abundance of species vulnerable to non-indigenous pests and pathogens.
 - Biomass mortality increased, driving a net reduction in aboveground C storage, particularly in areas where canopy trees were vulnerable to non-indigenous pests and pathogens.
@@
 - Under current trends this forest will continue to lose living biomass and be in danger of a critical transition to a lower biomass ecosystem.
-This indicates that the C sequestration of U.S.
-Eastern deciduous forests is likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.
+The C sequestration of Eastern US deciduous forests is therefore likely overestimated in current global carbon models, and these forests could even transition to a net source of C in the future.
 
 **Keywords**:
-
-*New Phytologist allows 5-8 keywords. Short phrases are acceptable*
 
@@
 The dominant pattern projected by Earth System Models is a continued, albeit weakening, temperate forest C sink for the remainder of the 21st century [@wu_uncertainty_2023; @ahlstrom_robustness_2012].
-Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020]<!--check refs-->, and current global C models do not represent certain influential mechanisms [e.g., @fatichi_moving_2014; @clark_continentwide_2021].
+Yet, global C models predict a wide range of future trajectories of carbon dioxide (CO~2~) sequestration [@ahlstrom_robustness_2012; @arora_carbon_2020], and current global C models do not represent certain influential mechanisms [*e.g.*, @fatichi_moving_2014; @clark_continentwide_2021].
 For instance, the impact of nuisance species, is not represented in global C - i.e., the impact of indigenous or non-indigenous species that, as a result of human influence, have much greater abundance in an ecosystem than they did historically, leading to detrimental ecological consequences [@skandrani_nuisance_2014; @nunez_population_2010].
@@
 White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a].
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species [*e.g.*, oak, *Quercus* spp.; @bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017).
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017).
 The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
@@
 
-The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (NZCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
+The study was conducted in the 25.6-hectare Forest Dynamics plot at the Smithsonian’s National Zoo and Conservation Biology Institute (SCBI) in Front Royal, Virginia (38$^\circ$ 53' 36.6''N, 78$^\circ$  08’ 43.4''W).
 The forest dynamics plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest.
@@
 Initially, plot-level AWM remained constant, maintaining rates of `r awm_wholePlot_2013` and `r awm_wholePlot_2018` Mg C ha^-1^ yr^-1^ in the first two census intervals.
-However, in the most recent period, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
-Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first interval.
+However, in the most recent period from 2018 to 2023, AWM increased to `r awm_wholePlot_2023` Mg C ha^-1^ yr^-1^.
+Despite initial uniformity in AWM across all plot regions, high canopy vulnerability areas experienced substantial increases in AWM across all censuses, resulting in a three-fold increase in AWM from the first to most recent interval.
 Low canopy vulnerability regions displayed smaller rises in AWM and only in the final survey period.
@@
 Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species.
-The increasing trend of AWM was driven primarily by two species: *Fraxinus americana* and *Quercus velutina* (Figure `r mortality_by_genera`).
-*Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
+The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak(*Quercus velutina*) (Figure `r mortality_by_genera`).
+*Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and loss peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.
 *Quercus velutina* experienced the second-highest rate of AWM at a rate of `r format(round(avg_quve,1),nsmall = 1)` Mg C ha^-1^ yr^-1^.

### 2024-04-05 doc/15yrsChange.Rmd
more cleaning

@@
 
-- The future of the critical temperate forest carbon (C) sink is contigent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood.
+- The future of the critical temperate forest carbon (C) sink is contingent on the role of nuisance species, including non-indigenous pests and pathogens and deer, yet their net effect on current and future C cycling is little understood.
 Here, we use 15 years of census data from a 25.6-ha plot in a Mid-Atlantic temperate forest to understand how nuisance species are affecting C sequestration.
@@
 `r loss_gain_net`, `r biomass_trends`).
-Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures [@dow_warm_2022; @anderson-teixeira_joint_2022].
+Specifically, reductions in individual tree growth rates at this site are associated with increasing stem diameters, an aging stand, and rising summer temperatures [@dow_warm_2022a; @anderson-teixeira_joint_2022].
 These complexities underscore the importance of multiple anthropogenic stressors, including but not limited to nuisance species, in shaping 21^st^ century forest carbon dynamics.
@@
 The ongoing emergence of new threats endangers other important canopy species: beech leaf disease, a non-indigenous pathogen lethal to *Fagus grandifolia* and first detected in 2012, is rapidly spreading throughout eastern US deciduous forests [@reed_distribution_2022].
-Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threat_2020].
+Another non-indigenous pathogen, Sudden Oak Death, is currently confined to the Western US but is causing widespread mortality in multiple Oak species and poses huge risks if it were to successfully spread to this region [@conrad_threats_2020].
 Thus, we consider it likely that oak decline, additional non-indigenous insects or pathogen species, and climate stressors will continue to elevate tree mortality at this site in the coming decades.
@@
 once infected, to such an extent that five species in North America (including *Fraxinus americana*) are now critically endangered (Knight et al, 2013, IUCN Red list).
-Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threat_2020] on forests across the region.
+Similarly, temperate forests of eastern North America continue to be affected by a variety of non-indigenous pests and pathogens [@anderson-teixeira_longterm_2021; @aukema_historical_2010; @fei_biomass_2019], and emerging threats (e.g., beach leaf disease, sudden oak death) could also have profound impacts [@reed_distribution_2022;@conrad_threats_2020] on forests across the region.
 Also paralleling our site, persistently elevated deer populations are reported throughout eastern US forests, and pressures from deer brousing have contributed to regneration debts of canopy species throughout the region [@miller_compounding_2019; @miller_overabundant_2023].
@@
 Through recruitment suppression, nuisance species can delay recovery from disturbances, anthropogenic or natural [@bradshaw_impacts_2016; @harvey_emergent_2023], and may even suppress regeneration into a closed-canopy for sustained periods.
-Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threat_2020; @reed_distribution_2022; @knight_factors_2013].
+Unlike many other disturbances, many non-indigenous pests and pathogens not only induce mortality, but then prevent recovery of their host species [@conrad_threats_2020; @reed_distribution_2022; @knight_factors_2013].
 Elimination of canopy species reduces biodiversity, which can decrease resilience to other disturbances (*e.g.* drought; [@anderegg_hydraulic_2018]) further endagerning forest health and C sequestration potential.
@@
 
-All data and code supporting this analysis are available via the public GitHub repository "15yrsChange" within the SCBI-ForestGEO organization (https://github.com/SCBI-ForestGEO/15yrsChange) and archived in Zenodo (DOI: TBD)<!--this can be obtained once the repo is made public-->.
+All data and code supporting this analysis are available via the public GitHub repository "scbi_nuisance_species_biomass" within the SCBI-ForestGEO organization (https://github.com/SCBI-ForestGEO/scbi_nuisance_species_biomass).
+<!-- and archived in Zenodo (DOI: TBD)this can be obtained once the repo is made public-->
 The most up-to-date versions of the data products analyzed here are publicly available via the ForestGEO data portal (https://forestgeo.si.edu/explore-data) and the SCBI-ForestGEO GitHub organization (https://github.com/SCBI-ForestGEO).

### 2024-04-09 doc/15yrsChange.Rmd
final submission

@@
 Their populations have dramatically increased in the past 50 years [@mcshea_science_2003], due in large part to global change [@cote_ecological_2004; @rooney_deer_2001; @brown_future_2000].
-White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013a].
-Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013a; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017).
-The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013a; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
+White-tailed deer preferentially browse on woody plant species in the seedling and sapling life stages, which reduces woody plant survival, growth and density [@mcgarvey_effects_2013b].
+Chronic white-tailed deer overabundance can reduce understory plant diversity and decrease the abundance of traditionally dominant woody species (*e.g.*, oak, *Quercus* spp.; [@bugalho_effects_2013; @habeck_communitylevel_2015] while giving an advantage to species they find less palatable, including both indigenous (*e.g.*, pawpaw, *Asimina tribola*; spicebush, *Lindera benzoin*) and non-indigenous (*e.g.*, garlic mustard, *Alliaria petiolata*) species (@averill_regional_2017).
+The strong impact of deer browsing, exacerbated by competition with dense deer-resistant understory vegetation and possibly climate change, results in dramatic reduction of canopy tree species in the understory [@mcgarvey_effects_2013b; @russell_interactions_2017; @miller_compounding_2019; @gorchov_differential_2021; @miller_overabundant_2023].
 
 Increased canopy tree mortality in combination with reduced canopy species recruitment in the understory is likely to have profound impacts on the deciduous forests of eastern North America.
-When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023a].
+When global change pressures (*e.g.*, climate change, nuisance species) inhibit recruitment of canopy species, the stage is set for disturbances to push forest ecosystems over a tipping point, after which there is little chance that a forest will recover to its pre-disturbance state in the foreseeable future [@anderson-teixeira_altered_2013; @mcdowell_pervasive_2020; @turner_novel_2023].
 In the mid-Atlantic region of the eastern US, forests face a severe regeneration debt: there are insufficient juveniles of canopy tree species to replace the mature cohort of that species when they eventually die [@miller_compounding_2019; @miller_overabundant_2023].
@@
 As part of the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus* spp.), and hickory (*Carya* spp.) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*), and witch hazel (*Hamamelis virginiana*).
-The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@anonymous_smithsonian_2011, @bourg_initial_2013, @mcgarvey_effects_2013a].
+The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 [@bourg_initial_2013b; @mcgarvey_effects_2013b].
 The plot, which includes a four-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters.
@@
 Though AWM increased throughout the plot AWM rates remained stable in a majority of individual species.
-The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak(*Quercus velutina*) (Figure `r mortality_by_genera`).
+The increasing trend of AWM was driven primarily by two species: white ash (*Fraxinus americana*) and black oak (*Quercus velutina*) (Figure `r mortality_by_genera`).
 *Fraxinus americana*, the host species for non-indigenous pest emerald ash borer (*Agrilus planipennis*), had the highest annual AWM of any species present on the plot, with an average flux of `r round(avg_fram,1)` Mg C ha^-1^ yr^-1^ and loss peaking at `r fram_peak` Mg C ha^-1^ yr^-1^ in 2020.

### 2024-05-31 doc/response_to_reviews.Rmd
start response to reviews
@llmorreale , I started an .Rmd for response to reviews

---
output: 
    #pdf_document: 
    bookdown::word_document2: 
      md_extensions: +raw_html-markdown_in_html_blocks
      pandoc_args: ['--lua-filter', 'read_html.lua']
      reference_docx: word-styles-reference.docx #make sure this line is indented! see this guide for formatting: https://rmarkdown.rstudio.com/articles_docx.html
  #officedown::rdocx_document
bibliography: citations.bib
csl: new-phytologist.csl

---
# RESPONSE TO REVIEWS {-}

NPH-MS-2024-46851 Nuisance species compromise carbon sequestration potential in an Eastern US temperate deciduous forest by Morreale, Luca; Hoffman, Rachel; Bagnaschi, Krystal; Kennedy, Iris; MacMonigle, Erin; Troy, Caroline; Dow, Cameron; Gonzalez-Akre, Erika; Herrmann, Valentine; Jordan, Jennifer; Magee, Lukas; Mitre, David; Pate, Christopher; Shue, Jessica; Bourg, Norman; McShea, William; Anderson-Teixeira, Kristina

# Editor {-}

The paper you submitted to New Phytologist, " Nuisance species compromise carbon sequestration potential in an Eastern US temperate deciduous forest" (NPH-MS-2024-46851) has been read by three reviewers, whose comments are below. The reviewers agreed that this is a well written paper on an interesting and important topic, and they recognize the value of long-term observations at a well characterized site. Nevertheless, they identified some substantial problems with the current analysis that preclude publication in New Phytologist. However, if you are able to fully address the current shortcomings, you are invited to submit a revised manuscript as a new paper for consideration.

Although your focus is on temporal trends in carbon storage sequestration, you do not present quantitative data on ecosystem C. Reviewer 1 points out that you have said little about the fate of C in dead trees and have not considered soil C at all. Reviewer 2 discusses the lack of any consideration of variability or any statistical support for your conclusions. The reviewer suggests analyzing variation among quadrants – yes, this would be pseudoreplication, but that is probably better than no replication at all.
Reviewer 2 and 3 both question your unsubstantiated statements about long-term trends and the provocative conclusion about a tipping point or possible transition to a net C source. I think that in this regard, some analysis of both near-term and long-term trends of forest growth and dynamics at the site prior to the current study could be useful (site history, tree-ring analysis?). For example, had you done this study a century ago when a nuisance species was decimating the American chestnut (perhaps not at this particular site, but certainly at similar Appalachian forests), would you have made similar conclusions about transition to lower biomass forest and perhaps a C source?

You emphasize that the effect of nuisance species should be included in models, but you provide no guidance on how this might be accomplished.

If you choose to revise your manuscript for reconsideration by New Phytologist, please address the other suggestions and concerns of the reviewers in addition to the major ones highlighted above. You should submit your revision via Manuscript Central as a new paper (see instructions below), and it will be sent out for review again. You will be asked to include a detailed description of how you responded to each of the reviewers' comments. Please note that it is the policy of New Phytologist to permit only one such resubmission, and I think that in this case the bar is high for a revision that meets all of these concerns. If on the other hand you decide instead to send your paper to another journal, I do hope you will find these reviewers' comments helpful.

Thank you for submitting your work to New Phytologist.

Sincerely,

Richard Norby
Editor, New Phytologist

# Referee: 1 {-}

In this paper titled “Nuisance species compromise carbon sequestration potential in an Eastern US temperate deciduous forest” Morreale et al. describe different aspects of  carbon accumulation at a  25.6-hectare Forest plot that are very well studied. Results are very interesting and eye opening, since they found that problematic animals and microbes can drastically reduce carbon sequestration and storage.  I have some questions about the paper and some suggestions.

My main concern is how some of these numbers were calculated. Based on my reading (but I may be confused) they seem to imply that dead trees do not sequester carbon. Although it is clear that dead trees stop increasing their biomass, they keep their carbon for decades (carbon storage) and in the meantime younger trees occupy their space since more light reach the understory. The authors show a 50% decrease in late succession species in areas with high density of deer. But you only need one tree to replace another, so even a 50% decrease may not imply that there will not be replacement.  I am not an expert on the models used here (and I guess that most readers are not), so some clarification on this can be important.

The first thing I thought when I read the title was about invasive plants, not animals and pathogens. You are focused here only is a subset of “nuisance species”. I wonder if results would change if you considered weeds in the mix (including nonnative trees). We have done some research on related topics (tree invasion in treeless areas, see for example papers titled “Unintended consequences of planting native and non-native trees in treeless ecosystems to mitigate climate change” or “Should tree invasions be used in treeless ecosystems to mitigate climate change?”) and it likely that even in your sites invasive plants actually reduce carbon, but it would be interesting if you mention something about nonnative plants in the discussion. Alternatively, you can change the title to make it more clear to what you are referring to. Lots of readers of this journal are plant focused.

The article highlighted the limitations that this was done only in one area, but there is another key limitation in my opinion. It is about the lack of data of soil carbon and belowground biomass. If we are curious about carbon sequestration in forests, soils are truly a fundamental part. Some discussion on this would be important.

All in all, this paper has some limitations but the data is indeed very interesting. I hope the authors can address these issues.

# Referee: 2 {-}

## Synopsis: {-}

Morreale et al. present a study of forest carbon dynamics in portions of the SCBI ForestGEO plot representing different scenarios of canopy tree pest/pathogens and white-tailed deer browsing. They evaluate trends in biomass change, growth, mortality, and recruitment of both the canopy and understory in these different scenarios of pathogens and deer browsing. Overall, they provide evidence that biomass is declining at SCBI, primarily driven by high mortality of Ash trees due to emerald ash borer, that woody growth has not kept pace to offset the high mortality, and that the species dominating woody recruitment are being heavily influenced by deer browsing. The authors posit that this evidence suggest we are overestimating the future temperate forest C sink and that we should be directly incorporating nuisance species into our Earth System Models.

## General Comments: {-}

Overall, I found this study very interesting and the paper to be extremely well written. The topic of nuisance species is an important one in temperate forests, and the interaction between pathogens and deer browsing is critical to the topic. That said, I had substantial concerns when reading the manuscript – mostly relating to the quantitative evidence for the claims made in the text. My largest concerns are the lack of statistical evidence for the claims made and the incomplete study design of the different “treatments” assessed. I have provided sets of general and specific comments that I hope assist the authors in improving the work:

While I appreciate the challenges of running statistics within a single plot and single exclosure area, most of the results section feels weakened by the lack of statistical support for the claims you make. Using 20x20m quadrats as replicates, accounting for spatial autocorrelation, and rarefying to get equal sample sizes or something like that could allow you to run statistical models to evaluate the claims you make with the 4-group figures you currently present. An analysis and paper of this sort needs statistics behind it.

The lack of a “high canopy vulnerability, low deer” designation in your study creates some fundamental challenges when interpreting the results – particularly on the influence of deer. The fact that within the exclosure, it looks like 75% of your quadrats have healthy canopies (Fig 6), means that the dynamics within the exclosure could be some combination of few deer, and relatively little canopy mortality. Are there areas within the exclosure that you could use as “high canopy vulnerability” such as the ~17 light blue quadrats in Fig inside the exclosure, or is the reduced canopy biomass in these driven by Oaks?

The substantial differences in plot area represented in your three “treatment” groups creates some challenges in the interpretation. While you don’t use statistics, in the broad sense you’re pulling information from groups with very different sample sizes. This makes it hard to know how much the deer story is attributable to the exclosure or to the limited area of the plot where the exlosure sits. This is even more prominent for the high-deer, high-vulnerability group, which is a few small portions of the plot with what look to be fairly unique topographic qualities (see additional general comment below).

I suggest that the assertion of a potential “tipping point” for these forests is too strong given the data presented here (and certainly too strong to be a bullet point in the Abstract). While I suspect many would agree with this assertion, at least qualitatively, the data presented here are a bit too site-specific and short term to evaluate this claim (the recruitment following Ash mortality just hasn’t had enough time to play out at this point in my opinion).

I think the study would benefit a lot from analyses focusing on the trees that aren’t experiencing large mortality events. Figure 2 presents things in terms of plot-level C, which is certainly important. That said, assessing the growth and mortality rates of the non Ash/Oak canopy trees might give the reader a better impression of how the rest of the forest is doing (i.e. are these baseline mortality rates in other trees higher or lower than normal?)

I think there’s important opportunity here to fold in abiotic characteristics into the analyses. Without an elevation map it’s hard to be sure, but the arrangement of the grey “streams” in figure 1 strongly suggests that the “high-deer, high vulnerability” quadrats are in low-lying areas at the top of these streams or something like that. This seems really important as these quadrats are the ones really driving the trends, and at the moment this seems like an important confounding factor that could/should be directly accounted for in your analyses. This could potentially change the conclusion to “there’s a lot of mortality in riparian areas which is strong enough to drive plot-level declines in biomass,” but also add in the capacity to discuss the effects of pathogens on these riparian areas specifically (similar dynamics to the well-documented hemlock wolly adelgid driven effects on streams).

## Specific Comments: {-}

L85-87: Or also the spread of the focal pest/pathogen to the full extent of the host range, right?

L227: You need a comma after “plot” to delineate the initial clause

L262: Particularly if you use a term like “significantly,” it’s important to provide statistical evidence for the statement

L268-271: It seems plausible that at least part of the effect described here (high annual recruitment of canopy taxa in low deer areas) is due to the fact that the low deer area also had low canopy vulnerability (ie the canopy trees were, on average, healthier and more fecund potentially).

L273-275: In regions of low deer density, these vulnerable canopy species are already in low abundance so I don’t think we would expect for them to have a substantial impact on total canopy recruitment.

L289-297: This seems like an ideal section to refer to Fig 6 for the first time.

L343-346: This is a critical statement to much of the language you use around tipping points, etc. As currently stated, this feels like the authors’ opinion, which is a well-informed one, but would be much stronger with some harder evidence to support the statement.

L366-380: This paragraph would be much stronger with some citation support for the actual spatial extent of this forest type, Fraxinus spp, EAB infection, increases in deer populations, etc. You provide citations that support that these nuisance species are important (as stated previously in the manuscript), but the argument of broad applicability would be much stronger if you provide numbers from the literature on how widespread the focal forest and nuisance species ranges are.

L381-395: I am not a modeler but work with them enough to know that they struggle with this sort of passage – effectively claiming that we need to incorporate some highly complex process into ESMs without any concrete ideas on how to do so. Modelers are flooded with things that could be incorporated into their models, but that incorporation is, of course, challenging. I would suggest either making a set of clear suggestions for how we might incorporate nuisance species into ESMs or cutting this paragraph.

L393-394: I understand and appreciate the sentiment with this statement, but it’s not uncommon for the elimination of canopy species to increase biodiversity – particularly in the understory and in the soil food web. This statement certainly isn’t a given and needs better support (or rewording/removal).

L403: I think you might mean “natural climate solution” here

L396-403: The final paragraph is very well worded and does a great job of wrapping up the discussion and the topic writ large.

**Thank you.**

# Referee: 3 {-}

## Comments to the Author {-}
In this manuscript, the author presents an interesting work about how nuisance species (i.e., non-indigenous pests and pathogens and deer density) could impact the carbon sequestration potential in an Eastern US temperate deciduous forest by analyzing 15 years of census data from a 25.6-ha plot deciduous forest. They found that pest and pathogen-driven tree mortality and reduced tree recruitment mainly led to the net carbon loss in this area, finally, the authors underscore the importance of including nuisance species in the carbon cycle model to better understand the future carbon changes and propose some forest management actions which could help to protect this ecosystem from the perspective of carbon sink. In general, I think this is a well-designed study and most of the manuscript is well-structured and easy to follow. This is a valuable contribution to the field of carbon sequestration potential and its impact from the biotic agents. However, I have two main comments that should be addressed.

## Major comments: {-}

1. The main conclusion in this study that nuisance species mainly led to the net carbon loss in the historical period is dependent on the census data in the past 15 years, I agree with this. However, the authors think this trend will also be in the future (Line 305), which I cannot fully agree with because there is not strong evidence in this analysis. How would the density of pests and pathogens and deer population change in the future? There are a lot of uncertainties here. I think necessary assumptions should be clearly made if the authors prefer to link their analysis to future predictions.

2. Expect nuisance species, many other factors impact forest carbon sequestration, for example, drought, windstorms, or even fires (although fires are much earlier to be identified). That is, the question here is how authors exclude other factors that can impact carbon change when you conclude that nuisance species are the most important factors. In addition, sorry I am not an expert on pests and pathogens, is there any difference between “non-indigenous” and “indigenous” pests and pathogens? If they are different, have the authors considered indigenous pests?

## Other comments: {-}

Line 52: change it to “carbon cycle models that have not considered [your conclusion]”

Line 271 and Figure 5: what are the units here? What does “n ha-1 year-1” mean here? 38.7 and 19 are far larger than the figure maximum numbers in the left panel.

Figure 6 is very different to understand, i.e., I even do not understand the authors’s main takeaways from this figure. I suggest modifying this figure and reorganizing the results part regarding this figure.

Line 308: What is AWP? Aboveground woody productivity? The authors used AWG in the first part of the manuscript.

Line 321: define “EAB” here

Lines 381. I totally agree that the inclusion of nuisance species in the carbon cycle model is a good point, but how? If the authors could elaborate on this a bit, that would be better.

# References {-}

<div id="refs"></div>

### 2024-06-03 doc/response_to_reviews.Rmd
begin outline of r2r

@@
 
+- Clarify language about carbon uptake vs. storage.
+  - Pests & pathogens are prematurely decreasing uptake
+  - Climate change is increasing decomposition, but unable to say anything about downed woody debris at this site specifically?
+- Look for numbers in literature about needed density of understory to replace overstory
+
 The first thing I thought when I read the title was about invasive plants, not animals and pathogens.
@@
 
+- Add discussion on role of non-indigenous plants
+  - Do we have data on non-indigenous plants?
+
 The article highlighted the limitations that this was done only in one area, but there is another key limitation in my opinion.
@@
 Some discussion on this would be important.
+
+- Add discussion about soils and belowground biomass
+  - We expect belowground biomass to track with aboveground
+  - Including how increasing canopy gaps would likely increase soil respiration (CO2 efflux) through increased temperatures
 
@@
 
+- Use quadrats as replicates to calculate statistical differences
+  - Calculate bootstrapped confidence intervals using quadrats
+  - Look into Rutishauser et al for use of quadrats
+
 The lack of a “high canopy vulnerability, low deer” designation in your study creates some fundamental challenges when interpreting the results – particularly on the influence of deer.
@@
 Are there areas within the exclosure that you could use as “high canopy vulnerability” such as the ~17 light blue quadrats in Fig inside the exclosure, or is the reduced canopy biomass in these driven by Oaks?
+
+- Doublecheck that the canopy mortality in exclosure is driven by Oaks
 
@@
 
+- Soften language around "tipping point"
+
 I think the study would benefit a lot from analyses focusing on the trees that aren’t experiencing large mortality events.
@@
 
+- Compare baseline mortality rates of other species with those from literature
+- Analysis of growth rates of trees not experiencing mortality events?
+
 I think there’s important opportunity here to fold in abiotic characteristics into the analyses.
@@
 This could potentially change the conclusion to “there’s a lot of mortality in riparian areas which is strong enough to drive plot-level declines in biomass,” but also add in the capacity to discuss the effects of pathogens on these riparian areas specifically (similar dynamics to the well-documented hemlock wolly adelgid driven effects on streams).
+
+- Analyze differences in TWI/topographical position in areas with high mortality
 
@@
 
+- Provide area from USFS or FIA about extent of this forest type. 
+  - Also extent of Fraxinus spp.
+
 L381-395: I am not a modeler but work with them enough to know that they struggle with this sort of passage – effectively claiming that we need to incorporate some highly complex process into ESMs without any concrete ideas on how to do so.
@@
 
+- Do we want to make specific suggestions or cut this?
+
 L393-394: I understand and appreciate the sentiment with this statement, but it’s not uncommon for the elimination of canopy species to increase biodiversity – particularly in the understory and in the soil food web.
 This statement certainly isn’t a given and needs better support (or rewording/removal).
+
+- Remove reference to biodiversity
 
@@
 
+- Is softening language around tipping points/future trends enough to address this?
+
 2.
@@
 If they are different, have the authors considered indigenous pests?
+
+- Add text clarifying that the mortality survey confirms that this site has not been affected by any large disturbances of windstorms or fires
+
+- Add some text discussing role of indigenous pest
 
